### 2017-09-23 inst/doc/ggalluvial.rmd
include usage in + reorganize documentation

---
title: "alluvial diagrams in ggplot2"
author: "Jason Cory Brunson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alluvial diagrams in ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The **ggalluvial** package strives to adapt the style and flexibility of the [**alluvial**](https://github.com/mbojan/alluvial) package to the principles and frameworks of the [**tidyverse**](https://github.com/tidyverse).
This vignette

- defines the essential components of alluvial diagrams as used in the naming schemes and documentation (*axis*, *alluvium*, *stratum*, *lode*, *flow*),
- describes the alluvial data structures recognized by **ggalluvial**,
- illustrates the new stats and geoms, and
- showcases some popular variants on the theme and how to produce them.

Many other resources exist for visualizing categorical data in R, including several more basic plot types that are likely to more accurately convey proportions to viewers when the data are not so structured as to warrant an alluvial diagram. In particular, check out Michael Friendly's [**vcd** and **vcdExtra** packages (PDF)](https://cran.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf) for a variety of statistically-motivated categorical data visualization techniques, Hadley Wickham's [**productplots** package](https://github.com/hadley/productplots) and Haley Jeppson and Heike Hofmann's descendant [**ggmosaic** package](https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html) for product or mosaic plots, and Nicholas Hamilton's [**ggtern** package](http://www.ggtern.com/) for ternary coordinates. Other related packages are mentioned below.

```{r setup, echo=FALSE, message=FALSE, results='hide'}
library(ggalluvial)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
```

## Alluvial diagrams

Here's a quintessential alluvial diagram:

```{r example alluvial diagram using Titanic dataset, echo=FALSE}
ggplot(data = to_lodes(as.data.frame(Titanic),
                       key = "Demographic",
                       axes = 1:3),
       aes(x = Demographic, stratum = value, alluvium = id,
           weight = Freq, label = value)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

The next section details how the elements of this image encode information about the underlying dataset.
For now, we use the image as a point of reference to define the following elements of a typical alluvial diagram:

- An *axis* is a dimension (variable) along which the data are vertically grouped at a fixed horizontal position. The diagram above uses three categorical axes: `Class`, `Sex`, and `Age`.
- The groups at each axis are depicted as opaque blocks called *strata*. For example, the `Class` axis contains four strata: `1st`, `2nd`, `3rd`, and `Crew`.
- Horizontal (x-) splines called *alluvia* span the width of the diagram. In this diagram, each alluvium corresponds to a fixed value of each axis variable, indicated by its vertical position at the axis, as well as of the `Survived` variable, indicated by its fill color.
- The segments of the alluvia between pairs of adjacent axes are *flows*.
- The alluvia intersect the strata at *lodes*, The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.

As the examples in the next section will demonstrate, which of these elements are incorporated into an alluvial diagram depends on both how the underlying data is structured and what the creator wants the diagram to communicate.

## Alluvial data

**ggalluvial** recognizes two formats of "alluvial data", detailed in the following subsections. A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.

[^tableform]: See Friendly's tutorial, linked above, for a discussion.

### One row per alluvium

The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": Each row corresponds to a subset or amount of observations that take a specific value at each variable, and each variable has its own column. An additional column contains the weight of each row. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

```{r Berkeley admissions dataset}
head(as.data.frame(UCBAdmissions), n = 12)
is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE)
```

This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

```{r alluvial diagram of UC Berkeley admissions dataset}
ggplot(as.data.frame(UCBAdmissions),
       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
  geom_alluvium(aes(fill = Admit), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_continuous(breaks = 1:2, labels = c("Gender", "Dept")) +
  ggtitle("UC Berkeley admissions and rejections, by sex and department")
```

An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations. The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).
**ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]

[^ggparallel]: A greater variety of parallel sets plots are implemented in [the **ggparallel** package](https://github.com/heike/ggparallel).

```{r parallel sets plot of Titanic dataset}
ggplot(as.data.frame(Titanic),
       aes(weight = Freq,
           axis1 = Survived, axis2 = Sex, axis3 = Class)) +
  geom_alluvium(aes(fill = Class),
                width = 1/8, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", label.strata = TRUE, reverse = FALSE) +
  scale_x_continuous(breaks = 1:3, labels = c("Survived", "Sex", "Class")) +
  coord_flip() +
  ggtitle("Titanic survival by class and sex")
```

This format and functionality are useful and will be retained in future versions. They also involve some conspicuous deviations from **ggplot2** norms:

- The `axis[0-9]*` position aesthetics are non-standard.
- `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
- The horizontal axis must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.

Furthermore, format aesthetics like `fill` are necessarily fixed for each alluvium; they cannot, for example, change from axis to axis according to the value taken at each. This means that, although they can reproduce the branching-tree structure of parallel sets, this format and functionality cannot produce alluvial diagrams with the color schemes featured [here](https://epijim.uk/code-snippets/eq5d/) ("Alluvial diagram") and [here](https://developers.google.com/chart/interactive/docs/gallery/sankey) ("Controlling colors"), which are "reset" at each axis.

### One row per lode

The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat. This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

```{r lode form of Berkeley admissions dataset}
UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
head(UCB_lodes, n = 12)
is_alluvial(UCB_lodes, logical = FALSE)
```

The same stat and geom can receive this data format using a different set of positional aesthetics, also specific to **ggalluvial**:

- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged along the horizontal axis;
- `stratum`, the "value" taken by the axis variable indicated by `x`; and
- `alluvium`, the indexing scheme that links the rows of a single alluvium.

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:

[^geom-area]: If bumping is unnecessary, consider using [`geom_area()`](https://chrisalbon.com/r-stats/stacked-area-graph.html) instead.
[^arguments]: `stat_stratum()` will similarly accept arguments for `x` and `stratum` without `alluvium`. If both strata and either alluvia or flows are to be plotted, though, all three parameters need arguments.

```{r time series alluvia diagram of refugees dataset}
data(Refugees, package = "alluvial")
country_regions <- c(
  Afghanistan = "Middle East",
  Burundi = "Central Africa",
  `Congo DRC` = "Central Africa",
  Iraq = "Middle East",
  Myanmar = "Southeast Asia",
  Palestine = "Middle East",
  Somalia = "Horn of Africa",
  Sudan = "Central Africa",
  Syria = "Middle East",
  Vietnam = "Southeast Asia"
)
Refugees$region <- country_regions[Refugees$country]
ggplot(data = Refugees,
       aes(x = year, weight = refugees, alluvium = country)) +
  geom_alluvium(aes(fill = country, colour = country),
                alpha = .75, decreasing = FALSE) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_color_brewer(type = "qual", palette = "Set3") +
  facet_wrap(~ region, scales = "fixed") +
  ggtitle("refugee volume by country of origin and geographic region")
```

The format allows us to assign aesthetics that change from axis to axis along the same alluvium, which is useful for repeated measures datasets. This requires generating a separate graphical object for each flow, as implemented in `geom_flow()`.
The plot below uses a set of (changes to) students' academic curricula over the course of several semesters.
Since `geom_flow()` calls `stat_flow()` by default (see the next example), we override it with `stat_alluvium()` in order to track each student across all semesters:

```{r alluvial diagram of majors dataset}
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum, label = curriculum)) +
  geom_flow(stat = "alluvium", lode.guidance = "rightleft") +
  geom_stratum() +
  theme(legend.position = "bottom") +
  ggtitle("student curricula across several semesters")
```

No `weight` is specified, so each row is assigned a weight of `1`.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

[^na.rm]: Be sure to set `na.rm` consistently in each layer, in this case both the flows and the strata.

Finally, lode format gives us the option to aggregate the flows between adjacent axes, which may be appropriate when the transitions between adjacent axes are of primary importance.
We can demonstrate this option on data from the influenza vaccination surveys conducted by the [RAND American Life Panel](https://alpdata.rand.org/):

```{r alluvial diagram of vaccinations dataset}
data(vaccinations)
levels(vaccinations$response) <- rev(levels(vaccinations$response))
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           weight = freq,
           fill = response, label = response)) +
  geom_flow() +
  geom_stratum(alpha = .5) +
  geom_text(stat = "stratum", size = 3) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses at three points in time")
```

This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes. This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.

## Appendix

[Michał Bojanowski](https://github.com/mbojan) makes a habit of including R session info in each vignette. This makes eminent sense to me, so i'm doing it here.

```{r session info}
sessionInfo()
```

### 2017-09-23 R/aaa-.r

@@ -1,0 +1,6 @@
+#' Base ggproto classes for ggalluvial
+#'
+#' @name ggalluvial-ggproto
+#' @seealso \code{\link[ggplot2]{ggplot2-ggproto}}
+#' @keywords internal
+NULL

### 2017-09-23 R/geom-alluvium.r

@@ -23,19 +23,17 @@
 #' }
 #' Currently, \code{group} is ignored.
 #' 
-#' @name geom-alluvium
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{stat_stratum}} and \code{\link{geom_stratum}} for 
 #'   intra-axis boxes, and
 #'   \code{\link{ggalluvial}} for a shortcut method.
-#' @inheritParams geom-lode
+#' @inheritParams geom_lode
 #' @param knot.pos The horizontal distance between a stratum (\code{width/2}
 #'   from its axis) and the knot of the x-spline, as a proportion of the
 #'   separation between strata. Defaults to 1/6.
 #' @param ribbon_bend Deprecated; alias for \code{knot.pos}.
 #' @example inst/examples/ex-geom-alluvium.r
-#' @usage NULL
 #' @export
 geom_alluvium <- function(mapping = NULL,
                           data = NULL,
@@ -64,7 +62,7 @@ geom_alluvium <- function(mapping = NULL,
   )
 }
 
-#' @rdname geom-alluvium
+#' @rdname ggalluvial-ggproto
 #' @usage NULL
 #' @export
 GeomAlluvium <- ggproto(

### 2017-09-23 R/geom-flow.r

@@ -25,17 +25,15 @@
 #' }
 #' Currently, \code{group} is ignored.
 #' 
-#' @name geom-flow
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{stat_stratum}} and \code{\link{geom_stratum}} for 
 #'   intra-axis boxes, and
 #'   \code{\link{ggalluvial}} for a shortcut method.
-#' @inheritParams geom-alluvium
+#' @inheritParams geom_alluvium
 #' @param aes.flow Character; how inter-lode flows assume aesthetics from lodes.
 #'   Options are "forward" and "backward".
 #' @example inst/examples/ex-geom-flow.r
-#' @usage NULL
 #' @export
 geom_flow <- function(mapping = NULL,
                       data = NULL,
@@ -69,7 +67,7 @@ geom_flow <- function(mapping = NULL,
   )
 }
 
-#' @rdname geom-flow
+#' @rdname ggalluvial-ggproto
 #' @usage NULL
 #' @export
 GeomFlow <- ggproto(

### 2017-09-23 R/geom-lode.r

@@ -22,20 +22,20 @@
 #' }
 #' Currently, \code{group} is ignored.
 #' 
-#' @name geom-lode
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{geom_alluvium}} for alluvial flows,
 #'   \code{\link{stat_stratum}} and \code{\link{geom_stratum}} for 
 #'   intra-axis stratum rectangles, and
 #'   \code{\link{ggalluvial}} for a shortcut method.
+#' @inheritParams ggplot2::layer
+#' @template common-params
 #' @param stat The statistical transformation to use on the data;
 #'    override the default.
 #' @param width Numeric; the width of each stratum, as a proportion of the
 #'   distance between axes. Defaults to 1/3.
 #' @param axis_width Deprecated; alias for \code{width}.
 #' @example inst/examples/ex-geom-lode.r
-#' @usage NULL
 #' @export
 geom_lode <- function(mapping = NULL,
                       data = NULL,
@@ -62,7 +62,7 @@ geom_lode <- function(mapping = NULL,
   )
 }
 
-#' @rdname geom-lode
+#' @rdname ggalluvial-ggproto
 #' @usage NULL
 #' @export
 GeomLode <- ggproto(

### 2017-09-23 man-roxygen/common-params.r

@@ -1,0 +1,5 @@
+#' @param na.rm Logical:
+#'   if \code{FALSE}, the default, \code{NA} lodes are not included;
+#'   if \code{TRUE}, \code{NA} lodes constitute a separate category,
+#'   plotted in grey (regardless of the color scheme).
+#' @param ... Additional arguments passed to \code{\link[ggplot2]{layer}}.

### 2017-10-22 inst/doc/ggalluvial.rmd
edit vignette

@@
 - The segments of the alluvia between pairs of adjacent axes are *flows*.
-- The alluvia intersect the strata at *lodes*, The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.
+- The alluvia intersect the strata at *lodes*, rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.
+The lodes are generated as parts of the alluvia in the above diagram, but they are obscured from view by the strata.
 
@@
 
-**ggalluvial** recognizes two formats of "alluvial data", detailed in the following subsections.
+**ggalluvial** recognizes two formats of "alluvial data", in which rows record either alluvia or lodes.
+The formats are related by a pair of operations, `to_lodes()` and `to_alluvia()`.
+These functions are adapted from `tidyr::spread()` and `tidyr::gather()` in order to also maintain the entanglement between alluvia and strata.
 A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.
@@
 This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**.
-It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:
+It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` must recognize and process in a consistent way:
 
@@
 
-An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations.
+The vertical axis scale is an important feature of these diagrams: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight (in this case, the count) of the observations.
 The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).
+
 **ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]
@@
 - `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
-- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
-- The horizontal axis must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.
+- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()` in the above code chunk) to take the values of the axis variables as labels.
+- The horizontal scale must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable indicating which axis variable each horizontal position corresponds to.
 
@@
 
-The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat.
+The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of `tidyr::gather()`ing the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat.
 This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:
@@
 
-- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged along the horizontal axis;
+- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged horizontally;
 - `stratum`, the "value" taken by the axis variable indicated by `x`; and
-- `alluvium`, the indexing scheme that links the rows of a single alluvium.
+- `alluvium`, the indexing scheme that links the lodes of a single alluvium.
 
 Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted.
-For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:
+For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:
 
@@
 
-The format allows us to assign aesthetics that change from axis to axis along the same alluvium, which is useful for repeated measures datasets.
+The format allows us to assign aesthetics that change from axis to axis along an alluvium, which is useful for repeated measures datasets.
 This requires generating a separate graphical object for each flow, as implemented in `geom_flow()`.
-The plot below uses a set of (changes to) students' academic curricula over the course of several semesters.
+The plot below visualizes a set of (changes to) students' academic curricula over the course of several semesters.
 Since `geom_flow()` calls `stat_flow()` by default (see the next example), we override it with `stat_alluvium()` in order to track each student across all semesters:
@@
 
-No `weight` is specified, so each row is assigned a weight of `1`.
+No `weight` is specified, so each row is assigned unit weight.
 This example demonstrates one way **ggalluvial** handles missing data.
@@
 This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes.
-This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.
+This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each individual or group longitudinally across the diagram.
 

### 2017-10-28 inst/doc/ggalluvial.rmd
arrange alluvia by aesthetics for consistent z-ordering

--- 2017-09-23 inst/doc/ggalluvial.rmd
@@
   geom_alluvium(aes(fill = Class),
-                width = 1/8, knot.pos = 0, reverse = FALSE) +
+                width = 0, knot.pos = 0, reverse = FALSE) +
   guides(fill = FALSE) +

### 2017-10-28 inst/doc/ggalluvial.R

@@ -31,7 +31,7 @@ ggplot(as.data.frame(Titanic),
        aes(weight = Freq,
            axis1 = Survived, axis2 = Sex, axis3 = Class)) +
   geom_alluvium(aes(fill = Class),
-                width = 1/8, knot.pos = 0, reverse = FALSE) +
+                width = 0, knot.pos = 0, reverse = FALSE) +
   guides(fill = FALSE) +
   geom_stratum(width = 1/8, reverse = FALSE) +
   geom_text(stat = "stratum", label.strata = TRUE, reverse = FALSE) +

### 2017-11-04 inst/doc/ggalluvial.R
add 'silent' argument to ''is_alluvial()'

@@ -15,7 +15,7 @@ ggplot(data = to_lodes(as.data.frame(Titanic),
 
 ## ----Berkeley admissions dataset-----------------------------------------
 head(as.data.frame(UCBAdmissions), n = 12)
-is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE)
+is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE, silent = TRUE)
 
 ## ----alluvial diagram of UC Berkeley admissions dataset------------------
 ggplot(as.data.frame(UCBAdmissions),
@@ -42,7 +42,7 @@ ggplot(as.data.frame(Titanic),
 ## ----lode form of Berkeley admissions dataset----------------------------
 UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
 head(UCB_lodes, n = 12)
-is_alluvial(UCB_lodes, logical = FALSE)
+is_alluvial(UCB_lodes, logical = FALSE, silent = TRUE)
 
 ## ----time series alluvia diagram of refugees dataset---------------------
 data(Refugees, package = "alluvial")

### 2017-11-04 R/stat-alluvium.r
change 'to_lodes()' defaults

@@ -133,11 +133,9 @@ StatAlluvium <- ggproto(
     # ensure that data is in lode form
     if (type == "alluvia") {
       axis_ind <- get_axes(names(data))
-      data <- to_lodes(data = data,
-                       key = "x", value = "stratum", id = "alluvium",
-                       axes = axis_ind)
+      data <- to_lodes(data = data, axes = axis_ind)
       # positioning requires numeric 'x'
-      #data$x <- as.numeric(as.factor(data$x))
+      data <- dplyr::arrange(data, x, stratum, alluvium)
       data$x <- cumsum(!duplicated(data$x))
     }
     

### 2017-11-04 R/stat-stratum.r

@@ -14,13 +14,22 @@
 #' and \code{axis[0-9]*} for data in alluvia format
 #' (see \code{\link{is_alluvial}}).
 #' Arguments to parameters inconsistent with the format will be ignored.
+#' 
 #' Additionally, \code{stat_stratum} accepts the following optional aesthetics:
 #' \itemize{
 #'   \item \code{weight}
+#'   \item \code{label}
 #'   \item \code{group}
 #' }
 #' \code{weight} controls the vertical dimensions of the alluvia
 #' and are aggregated across equivalent observations.
+#' \code{label} is used to label the strata and must take a unique value across
+#' the observations within each stratum.
+#' These and any other aesthetics are aggregated as follows:
+#' Numeric aesthetics, including \code{weight}, are summed.
+#' Character and factor aesthetics, including \code{label},
+#' are assigned to strata provided they take unique values across the
+#' observations within each stratum (otherwise \code{NA} is assigned).
 #' \code{group} is used internally; arguments are ignored.
 #' 
 #' @import ggplot2
@@ -43,9 +52,9 @@
 #'   so that they match the order of the values in the legend.
 #'   Ignored if \code{decreasing} is not \code{NA}.
 #'   Defaults to \code{TRUE}.
-#' @param label.strata Logical; whether to assign the values of the axis
-#'   variables to the strata. Defaults to FALSE, and requires that no label
-#'   aesthetic is assigned.
+#' @param label.strata Logical; whether to assign the values of the axis 
+#'   variables to the strata. Defaults to FALSE, and requires that no
+#'   \code{label} aesthetic is assigned.
 #' @example inst/examples/ex-stat-stratum.r
 #' @export
 stat_stratum <- function(mapping = NULL,
@@ -118,15 +127,13 @@ StatStratum <- ggproto(
     # ensure that data is in lode form
     if (type == "alluvia") {
       axis_ind <- get_axes(names(data))
-      data <- to_lodes(data = data,
-                       key = "x", value = "stratum", id = "alluvium",
-                       axes = axis_ind)
+      data <- to_lodes(data = data, axes = axis_ind)
       # positioning requires numeric 'x'
-      #data$x <- as.numeric(as.factor(data$x))
+      data <- dplyr::arrange(data, x, stratum, alluvium)
       data$x <- cumsum(!duplicated(data$x))
     }
     
-    # nullify 'group' and 'alluvium' fields
+    # nullify 'group' and 'alluvium' fields (to avoid confusion with geoms)
[…]

### 2017-11-05 inst/doc/ggalluvial.rmd
debug 'aggregate.wts' + default FALSE

--- 2017-10-28 inst/doc/ggalluvial.rmd
@@
                        axes = 1:3),
-       aes(x = Demographic, stratum = value, alluvium = id,
-           weight = Freq, label = value)) +
+       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
+           weight = Freq, label = stratum)) +
   geom_alluvium(aes(fill = Survived)) +
@@
 head(as.data.frame(UCBAdmissions), n = 12)
-is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE)
+is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE, silent = TRUE)
 ```
@@
 head(UCB_lodes, n = 12)
-is_alluvial(UCB_lodes, logical = FALSE)
+is_alluvial(UCB_lodes, logical = FALSE, silent = TRUE)
 ```
@@
            fill = curriculum, label = curriculum)) +
-  geom_flow(stat = "alluvium", lode.guidance = "rightleft") +
+  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
+            color = "black") +
   geom_stratum() +

### 2017-11-05 R/geom-alluvium.r

@@ -6,6 +6,7 @@
 #' It plots both the lodes themselves, using \code{\link{geom_lode}}, and the
 #' flows between them, using \code{\link{geom_flow}}.
 #' 
+
 #' @section Aesthetics:
 #' \code{geom_alluvium} understands the following aesthetics
 #' (required aesthetics are in bold):
@@ -23,6 +24,7 @@
 #' }
 #' \code{group} is used internally; arguments are ignored.
 #' 
+
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{stat_stratum}} and \code{\link{geom_stratum}} for 

### 2017-11-05 R/geom-lode.r

@@ -5,6 +5,7 @@
 #' of an alluvial diagram, the intersections of the alluvia with the strata.
 #' It plots rectangles for these lodes of a provided \code{width}.
 #' 
+
 #' @section Aesthetics:
 #' \code{geom_lode} understands the following aesthetics
 #' (required aesthetics are in bold):
@@ -22,6 +23,7 @@
 #' }
 #' \code{group} is used internally; arguments are ignored.
 #' 
+
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{geom_alluvium}} for alluvial flows,

### 2017-11-05 R/geom-stratum.r

@@ -5,6 +5,7 @@
 #' alluvial diagram.
 #' It plots rectangles for these strata of a provided \code{width}.
 #' 
+
 #' @section Aesthetics:
 #' \code{geom_stratum} understands the following aesthetics
 #' (required aesthetics are in bold):
@@ -22,6 +23,7 @@
 #' }
 #' \code{group} is used internally; arguments are ignored.
 #' 
+
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments, and
 #'   \code{\link{stat_stratum}} for the corresponding geom.

### 2017-11-05 R/stat-alluvium.r

@@ -6,6 +6,7 @@
 #' \code{ymin} and \code{ymax}). It leverages the \code{group} aesthetic for 
 #' plotting purposes (for now).
 #' 
+
 #' @section Aesthetics:
 #' \code{stat_alluvium} requires one of two sets of aesthetics:
 #' \itemize{
@@ -25,6 +26,7 @@
 #' and are aggregated across equivalent observations.
 #' \code{group} is used internally; arguments are ignored.
 #' 
+
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{geom_alluvium}} for the corresponding geom,
@@ -32,11 +34,13 @@
 #'   intra-axis boxes, and
 #'   \code{\link{ggalluvial}} for a shortcut method.
 #' @inheritParams stat_flow
-#' @param aggregate.wts Logical; whether to aggregate weights across otherwise
-#'   equivalent rows before computing lode and flow positions. Defaults to TRUE.
+#' @param aggregate.wts Whether to aggregate weights across otherwise equivalent
+#'   rows before computing lode and flow positions. Set to \code{TRUE} to group
+#'   observations into cohorts.
+#'   \strong{Warning}: This is currently an expensive operation.
 #' @param lode.guidance The function to prioritize the axis variables for 
-#'   ordering the lodes within each stratum. Defaults to "zigzag", other options
-#'   include "rightleft", "leftright", "rightward", and "leftward" (see 
+#'   ordering the lodes within each stratum. Options are "zigzag", "rightleft",
+#'   "leftright", "rightward", and "leftward" (see 
 #'   \code{\link{lode-guidance-functions}}).
 #' @param lode.ordering A list (of length the number of axes) of integer vectors
 #'   (each of length the number of rows of \code{data}) or NULL entries 
@@ -52,7 +56,7 @@ stat_alluvium <- function(mapping = NULL,
                           position = "identity",
                           decreasing = NA,
                           reverse = TRUE,
-                          aggregate.wts = TRUE,
+                          aggregate.wts = FALSE,
                           lode.guidance = "zigzag",
                           lode.ordering = NULL,
                           na.rm = FALSE,
@@ -70,7 +74,7 @@ stat_alluvium <- function(mapping = NULL,
     params = list(
       decreasing = decreasing,
       reverse = reverse,
-      aggregate.wts = TRUE,
+      aggregate.wts = FALSE,
[…]

### 2017-11-05 R/stat-stratum.r

@@ -3,6 +3,7 @@
 #' Given a dataset with alluvial structure, \code{stat_stratum} calculates the
 #' centroids of the strata at each axis, together with their weights (heights).
 #' 
+
 #' @section Aesthetics:
 #' \code{stat_stratum} requires one of two sets of aesthetics:
 #' \itemize{
@@ -32,6 +33,7 @@
 #' observations within each stratum (otherwise \code{NA} is assigned).
 #' \code{group} is used internally; arguments are ignored.
 #' 
+
 #' @import ggplot2
 #' @seealso \code{\link[ggplot2]{layer}} for additional arguments,
 #'   \code{\link{geom_stratum}} for the corresponding geom,

### 2017-11-05 R/stat-utils.r

@@ -52,3 +52,15 @@ na_keep <- function(data, type) {
   }
   data
 }
+
+# arrange data by aesthetics for consistent (reverse) z-ordering
+z_order_colors <- function(data) {
+  stopifnot("group" %in% names(data))
+  # arrange by fill, then color
+  z_aes <- intersect(c("fill", "colour", "group"), names(data))
+  if (length(z_aes) > 0) {
+    data <- data[do.call(order, data[, z_aes, drop = FALSE]), , drop = FALSE]
+    data$group <- cumsum(!duplicated(data$group))
+  }
+  data
+}

### 2017-11-05 inst/doc/ggalluvial.R

@@ -74,7 +74,8 @@ majors$curriculum <- as.factor(majors$curriculum)
 ggplot(majors,
        aes(x = semester, stratum = curriculum, alluvium = student,
            fill = curriculum, label = curriculum)) +
-  geom_flow(stat = "alluvium", lode.guidance = "rightleft") +
+  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
+            color = "black") +
   geom_stratum() +
   theme(legend.position = "bottom") +
   ggtitle("student curricula across several semesters")

### 2017-11-06 inst/doc/ggalluvial.rmd

@@
 
-### One row per alluvium
+### Alluvia format
 
-The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": Each row corresponds to a subset or amount of observations that take a specific value at each variable, and each variable has its own column.
-An additional column contains the weight of each row.
+The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g.
+the number of observational units in the cohort.
 This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:
 
-```{r Berkeley admissions dataset}
+```{r alluvia format of Berkeley admissions dataset}
 head(as.data.frame(UCBAdmissions), n = 12)
@@
 
-This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**.
+This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**.
 It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:
@@
 
-### One row per lode
+### Lodes format
 
-The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat.
-This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:
+The second format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the `key` and the stratum as the `value`.
+This format requires an additional indexing `id` column that links the rows corresponding to a common cohort, i.e.
+the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:
 
-```{r lode form of Berkeley admissions dataset}
+```{r lodes format of Berkeley admissions dataset}
 UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
@@
 Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted.
-For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:
+For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:
 
@@
                 alpha = .75, decreasing = FALSE) +
+  scale_x_continuous(breaks = seq(2003, 2013, 2)) +
+  theme(axis.text.x = element_text(angle = -30, hjust = 0)) +
   scale_fill_brewer(type = "qual", palette = "Set3") +
@@
   facet_wrap(~ region, scales = "fixed") +
-  ggtitle("refugee volume by country of origin and geographic region")
+  ggtitle("refugee volume by country and region of origin")
 ```
@@
   geom_flow(stat = "alluvium", lode.guidance = "rightleft",
-            color = "black") +
+            color = "darkgray") +
   geom_stratum() +
@@
 
-This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes.
+This diagram ignores any continuity between the flows between axes.
 This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.

### 2017-11-12 inst/doc/ggalluvial.rmd

@@
                        axes = 1:3),
-       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
-           weight = Freq, label = stratum)) +
+       aes(x = Demographic, stratum = value, alluvium = id,
+           weight = Freq, label = value)) +
   geom_alluvium(aes(fill = Survived)) +
@@
            fill = curriculum, label = curriculum)) +
-  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
-            color = "darkgray") +
+  geom_flow(stat = "alluvium", lode.guidance = "rightleft", color = "darkgrey") +
   geom_stratum() +

### 2017-11-18 inst/doc/ggalluvial.rmd

--- 2017-11-06 inst/doc/ggalluvial.rmd
@@
 
-- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged along the horizontal axis;
+- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged horizontally;
 - `stratum`, the "value" taken by the axis variable indicated by `x`; and
-- `alluvium`, the indexing scheme that links the rows of a single alluvium.
+- `alluvium`, the indexing scheme that links the lodes of a single alluvium.
 

### 2017-11-18 inst/doc/ggalluvial.rmd
mention ggforce in vignette

--- 2017-11-06 inst/doc/ggalluvial.rmd
@@
 An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations.
-The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).
+The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).[^ggforce]
 **ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]
 
+[^ggforce]: [The **ggforce** package](https://github.com/thomasp85/ggforce) includes parallel set geom and stat layers to produce similar diagrams that can be allowed to free-float.
 [^ggparallel]: A greater variety of parallel sets plots are implemented in [the **ggparallel** package](https://github.com/heike/ggparallel).

### 2018-02-03 README.rmd
rmd README

---
output: github_document
---

<!-- README.md is generated from README.rmd. Please edit that file -->

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ggalluvial

## Design

This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial diagrams. The alluvial plots implemented here can be used to visualize frequency distributions over time or frequency tables involving several categorical variables. The design is derived mostly from the [**alluvial**](https://github.com/mbojan/alluvial) package, but the **ggplot2** framework induced several conspicuous differences:

- **alluvial** understands a variety of inputs (vectors, lists, data frames), while **ggalluvial** requires a single data frame;
- **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](http://tidyr.tidyverse.org/);
- **alluvial** produces both the *alluvia*, which link cohorts across multiple dimensions, and (what are here called) the *strata*, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called *lodes* and *flows*.

## Installation

The latest stable release can be installed from CRAN:
```{r eval=FALSE}
install.packages("ggalluvial")
```

The [cran branch](https://github.com/corybrunson/ggalluvial/tree/cran) will contain the version most recently submitted to [CRAN](https://cran.r-project.org/package=ggalluvial).

Development versions can be installed from GitHub:
```{r eval=FALSE}
devtools::install_github("corybrunson/ggalluvial", build_vignettes = TRUE)
```
The
[optimization branch](https://github.com/corybrunson/ggalluvial/tree/optimization) contains a development version with experimental functions to reduce the number or area of alluvial overlaps (see issue [#6](/../../issues/6)). Install it as follows:
```{r eval=FALSE}
devtools::install_github("corybrunson/ggalluvial", ref = "optimization")
```

## Usage

```{r eval=TRUE, echo=FALSE, message=FALSE}
devtools::load_all()
```

### Example

Here is how to generate an alluvial diagram representation of the multi-dimensional categorical dataset of passengers on the Titanic:

```{r eval=TRUE}
titanic_wide <- data.frame(Titanic)
head(titanic_wide)
ggplot(data = titanic_wide,
       aes(axis1 = Class, axis2 = Sex, axis3 = Age,
           weight = Freq)) +
  scale_x_continuous(breaks = 1:3, labels = c("Class", "Sex", "Age")) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

The data is in "wide" format, but **ggalluvial** also recognizes data in "long" format and can convert between the two:

```{r eval=TRUE}
titanic_long <- to_lodes(data.frame(Titanic),
                         key = "Demographic",
                         axes = 1:3)
head(titanic_long)
ggplot(data = titanic_long,
       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
           weight = Freq, label = stratum)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

### Resources

For detailed discussion of the data formats recognized by **ggalluvial** and several examples that illustrate its flexibility and limitations, read the vignette:

```{r eval=FALSE}
vignette(topic = "ggalluvial", package = "ggalluvial")
```

The documentation contains several examples; use `help()` to call forth examples of any layer (`stat_*` or `geom_*`).

## Feedback

### Cite

If you use **ggalluvial**-generated figures in publication, i'd be grateful to hear about it! You can also cite the package according to `citation("ggalluvial")`.

### Contribute

Issues and pull requests are more than welcome! Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.

### 2018-02-05 inst/examples/ex-alluvial-data.r
organize pkgdown reference

@@ -50,17 +50,17 @@ head(majors_alluvia2)
 gg <- ggplot(majors_alluvia,
              aes(axis1 = CURR1, axis2 = CURR7, axis3 = CURR13))
 gg +
-  geom_alluvium(aes(fill = as.factor(student)), discern = TRUE) +
-  geom_stratum(discern = TRUE) +
+  geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = TRUE) +
+  geom_stratum(width = 2/5, discern = TRUE) +
   geom_text(stat = "stratum", discern = TRUE, label.strata = TRUE)
 gg +
-  geom_alluvium(aes(fill = as.factor(student)), discern = FALSE) +
-  geom_stratum(discern = FALSE) +
+  geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = FALSE) +
+  geom_stratum(width = 2/5, discern = FALSE) +
   geom_text(stat = "stratum", discern = FALSE, label.strata = TRUE)
 # warning when inappropriate
 ggplot(majors[majors$semester %in% paste0("CURR", c(1, 7, 13)), ],
        aes(x = semester, stratum = curriculum, alluvium = student,
            label = curriculum)) +
-  geom_alluvium(aes(fill = as.factor(student)), discern = TRUE) +
-  geom_stratum(discern = TRUE) +
+  geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = TRUE) +
+  geom_stratum(width = 2/5, discern = TRUE) +
   geom_text(stat = "stratum", discern = TRUE)

### 2018-02-05 README.rmd
organize pkgdown reference
use scale_x_discrete for categorical axis set

@@
 
+<!--
+[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
+-->
+
+This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial diagrams.
+
 ## Design
 
-This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial diagrams.
 The alluvial plots implemented here can be used to visualize frequency distributions over time or frequency tables involving several categorical variables.
@@
            weight = Freq)) +
-  scale_x_continuous(breaks = 1:3, labels = c("Class", "Sex", "Age")) +
+  scale_x_discrete(limits = c("Class", "Sex", "Age")) +
   geom_alluvium(aes(fill = Survived)) +

### 2018-02-05 inst/examples/ex-stat-alluvium.r

@@ -7,14 +7,13 @@ ggplot(as.data.frame(Titanic),
   geom_line(stat = "alluvium") +
   stat_alluvium(geom = "pointrange") +
   geom_text(stat = "stratum", label.strata = TRUE) +
-  scale_x_continuous(breaks = 1:3,
-                     labels = c("Class", "Sex", "Age"))
+  scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 gg <- ggplot(as.data.frame(Titanic),
              aes(weight = Freq,
                  axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
-  scale_x_continuous(breaks = 1:3, labels = c("Class", "Sex", "Age"))
+  scale_x_discrete(limits = c("Class", "Sex", "Age"))
 # use of lode controls
 gg + geom_flow(aes(fill = Survived, alpha = Sex), stat = "alluvium",
                aes.bind = TRUE, lode.guidance = "rightward")

### 2018-02-11 README.rmd
use 'expand' in 'scale_discrete' to reduce margins

@@
            weight = Freq)) +
-  scale_x_discrete(limits = c("Class", "Sex", "Age")) +
+  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
   geom_alluvium(aes(fill = Survived)) +

### 2018-02-11 inst/doc/ggalluvial.rmd

@@
 ---
-title: "alluvial diagrams in ggplot2"
+title: "Alluvial Diagrams in ggplot2"
 author: "Jason Cory Brunson"
@@
 Many other resources exist for visualizing categorical data in R, including several more basic plot types that are likely to more accurately convey proportions to viewers when the data are not so structured as to warrant an alluvial diagram.
-In particular, check out Michael Friendly's [**vcd** and **vcdExtra** packages (PDF)](https://cran.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf) for a variety of statistically-motivated categorical data visualization techniques, Hadley Wickham's [**productplots** package](https://github.com/hadley/productplots) and Haley Jeppson and Heike Hofmann's descendant [**ggmosaic** package](https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html) for product or mosaic plots, and Nicholas Hamilton's [**ggtern** package](http://www.ggtern.com/) for ternary coordinates.
+In particular, check out Michael Friendly's [**vcd** and **vcdExtra** packages (PDF)](https://CRAN.R-project.org/package=vcdExtra/vignettes/vcd-tutorial.pdf) for a variety of statistically-motivated categorical data visualization techniques, Hadley Wickham's [**productplots** package](https://github.com/hadley/productplots) and Haley Jeppson and Heike Hofmann's descendant [**ggmosaic** package](https://CRAN.R-project.org/package=ggmosaic/vignettes/ggmosaic.html) for product or mosaic plots, and Nicholas Hamilton's [**ggtern** package](http://www.ggtern.com/) for ternary coordinates.
 Other related packages are mentioned below.
@@
 - The segments of the alluvia between pairs of adjacent axes are *flows*.
-- The alluvia intersect the strata at *lodes*, The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.
+- The alluvia intersect the strata at *lodes*.
+The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.
 
@@
 
-**ggalluvial** recognizes two formats of "alluvial data", detailed in the following subsections.
-A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.
+**ggalluvial** recognizes two formats of "alluvial data", treated in detail in the following subsections, but which basically correspond to the "wide" and "long" formats of categorical repeated measures data.
+A third, tabular (or array), form is popular for storing data with multiple categorical dimensions, such as the `Titanic` and `UCBAdmissions` datasets.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input; `base::data.frame()` converts such an array to an acceptable data frame.
 
@@
 
-### Alluvia format
+### Alluvia (wide) format
 
-The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g.
+The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g.
 the number of observational units in the cohort.
@@
 
-This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**.
-It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:
+This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:
 
@@
   geom_label(stat = "stratum", label.strata = TRUE) +
-  scale_x_continuous(breaks = 1:2, labels = c("Gender", "Dept")) +
+  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
+  scale_fill_brewer(type = "qual", palette = "Set1") +
   ggtitle("UC Berkeley admissions and rejections, by sex and department")
@@
 An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations.
-The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).[^ggforce]
+The diagrams produced by **ggalluvial** conform (somewhat; keep reading) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).[^ggforce]
 **ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]
@@
 
-This format and functionality are useful and will be retained in future versions. They also involve some conspicuous deviations from **ggplot2** norms:
+This format and functionality are useful for many applications and will be retained in future versions. They also involve some conspicuous deviations from **ggplot2** norms:
 
-- The `axis[0-9]*` position aesthetics are non-standard.
+- The `axis[0-9]*` position aesthetics are non-standard: they are not an explicit set of parameters but a family based on a regular expression pattern; and at least one, but no specific one, is required.
 - `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
 - The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
-- The horizontal axis must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.
+- The horizontal axis must be manually corrected (using `scale_x_discrete()` or `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.
 
@@
 
-### Lodes format
+### Lodes (long) format
 
-The second format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the `key` and the stratum as the `value`.
-This format requires an additional indexing `id` column that links the rows corresponding to a common cohort, i.e.
+The long format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) or "pivoting" (in the Microsoft Excel sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the key and the stratum as the value.
+This format requires an additional indexing column that links the rows corresponding to a common cohort, i.e.
 the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:
@@
 
-The same stat and geom can receive this data format using a different set of positional aesthetics, also specific to **ggalluvial**:
+The functions that convert data between wide (alluvia) and long (lodes) format include several parameters that help preserve ancillary information. See `help("alluvial-data")` for examples.
+
+The same stat and geom can receive data in this format using a different set of positional aesthetics, also specific to **ggalluvial**:
 
@@
 
-Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted.
-For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:
+Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and often not plotted.
+For convenience, both `stat_alluvium()` and `stat_flow()` will accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:
 
-[^geom-area]: If bumping is unnecessary, consider using [`geom_area()`](https://chrisalbon.com/r-stats/stacked-area-graph.html) instead.
+[^geom-area]: If bumping is unnecessary, consider using [`geom_area()`](http://www.r-graph-gallery.com/136-stacked-area-chart/) instead.
 [^arguments]: `stat_stratum()` will similarly accept arguments for `x` and `stratum` without `alluvium`.
@@
            fill = curriculum, label = curriculum)) +
+  scale_fill_brewer(type = "qual", palette = "Set2") +
   geom_flow(stat = "alluvium", lode.guidance = "rightleft",

### 2018-05-17 R/devel.r
update version + deprecate is_alluvial() + deprecate logical parameter

@@ -7,6 +7,19 @@
 #' @keywords internal
 NULL
 
+deprecate_parameter <- function(old, new = NA) {
+  .Deprecated(msg = paste0(
+    "The parameter `", old, "` is deprecated.",
+    if (is.null(new)) {
+      "\nPass unparameterized arguments instead."
+    } else if (!is.na(new)) {
+      paste0("\nPass arguments to `", new, "` instead.")
+    } else {
+      ""
+    }
+  ))
+}
+
 release_questions <- function() {
   c(
     "Have previous CRAN NOTEs been addressed?"

### 2018-05-17 R/geom-alluvium.r

@@ -68,13 +68,13 @@ GeomAlluvium <- ggproto(
     }
     
     if (!is.null(params$axis_width)) {
-      warning("Parameter `axis_width` is deprecated; use `width` instead.")
+      deprecate_parameter("axis_width", "width")
       params$width <- params$axis_width
       params$axis_width <- NULL
     }
     
     if (!is.null(params$ribbon_bend)) {
-      warning("Parameter `ribbon_bend` is deprecated; use `knot.pos` instead.")
+      deprecate_parameter("ribbon_bend", "knot.pos")
       params$knot.pos <- params$ribbon_bend
       params$ribbon_bend <- NULL
     }

### 2018-05-17 R/geom-flow.r

@@ -66,13 +66,13 @@ GeomFlow <- ggproto(
   setup_params = function(data, params) {
     
     if (!is.null(params$axis_width)) {
-      warning("Parameter `axis_width` is deprecated; use `width` instead.")
+      deprecate_parameter("axis_width", "width")
       params$width <- params$axis_width
       params$axis_width <- NULL
     }
     
     if (!is.null(params$ribbon_bend)) {
-      warning("Parameter `ribbon_bend` is deprecated; use `knot.pos` instead.")
+      deprecate_parameter("ribbon_bend", "knot.pos")
       params$knot.pos <- params$ribbon_bend
       params$ribbon_bend <- NULL
     }

### 2018-05-17 R/geom-stratum.r

@@ -53,7 +53,7 @@ GeomStratum <- ggproto(
   setup_params = function(data, params) {
     
     if (!is.null(params$axis_width)) {
-      warning("Parameter `axis_width` is deprecated; use `width` instead.")
+      deprecate_parameter("axis_width", "width")
       params$width <- params$axis_width
       params$axis_width <- NULL
     }

### 2018-05-17 inst/examples/ex-alluvial-data.r

@@ -1,42 +1,39 @@
 # Titanic data in alluvia format
 titanic_alluvia <- as.data.frame(Titanic)
 head(titanic_alluvia)
-is_alluvial(titanic_alluvia,
-            weight = "Freq")
+is_alluvia_form(titanic_alluvia,
+                weight = "Freq")
 # Titanic data in lodes format
 titanic_lodes <- to_lodes_form(titanic_alluvia,
                                key = "x", value = "stratum", id = "alluvium",
                                axes = 1:4)
 head(titanic_lodes)
-is_alluvial(titanic_lodes,
-            key = "x", value = "stratum", id = "alluvium",
-            weight = "Freq")
+is_lodes_form(titanic_lodes,
+              key = "x", value = "stratum", id = "alluvium",
+              weight = "Freq")
 # again in lodes format, this time diffusing the `Class` variable
 titanic_lodes2 <- to_lodes_form(titanic_alluvia,
                                 key = variable, value = value,
                                 id = passenger,
                                 1:3, diffuse = Class)
 head(titanic_lodes2)
-is_alluvial(titanic_lodes2,
-            key = variable, value = value, id = passenger,
-            weight = Freq)
+is_lodes_form(titanic_lodes2,
+              key = variable, value = value, id = passenger,
+              weight = Freq)
 
 # curriculum data in lodes format
 data(majors)
 head(majors)
-is_alluvial(majors,
-            key = "semester", value = "curriculum", id = "student",
-            logical = FALSE)
+is_lodes_form(majors,
+              key = "semester", value = "curriculum", id = "student")
 # curriculum data in alluvia format
 majors_alluvia <- to_alluvia_form(majors,
                                   key = "semester", value = "curriculum",
                                   id = "student")
 head(majors_alluvia)
-is_alluvial(majors_alluvia,
-            2:9,
-            logical = FALSE)
+is_alluvia_form(majors_alluvia, axes = 2:9)
 
-# distill variables that vary within 'id' values
+# distill variables that vary within `id` values
 set.seed(1)
 majors$hypo_grade <- LETTERS[sample(5, size = nrow(majors), replace = TRUE)]
 majors_alluvia2 <- to_alluvia_form(majors,

### 2018-05-17 inst/examples/ex-stat-stratum.r

@@ -1,4 +1,4 @@
-# only 'stratum' assignment is necessary to generate strata
+# only `stratum` assignment is necessary to generate strata
 data(vaccinations)
 ggplot(vaccinations,
        aes(weight = freq,

### 2018-05-17 tests/testthat/test-is-alluvial.r

@@ -1,16 +1,13 @@
 library(ggalluvial)
-context("is_alluvial")
+context("alluvial-data")
 
 titanic_alluvia <- as.data.frame(Titanic)
 
-test_that("is_alluvial recognizes alluvia-format Titanic data", {
-  expect_message(is_alluvial(titanic_alluvia),
-                 "[Mm]issing")
-  expect_true(is_alluvial(titanic_alluvia, axes = 1:4))
-  expect_true(is_alluvial(titanic_alluvia, axes = c("Class", "Sex")))
-  expect_warning(is_alluvial(titanic_alluvia,
-                             key = "Class", value = "Freq", id = "Age"),
-                 "[Dd]uplicate")
+# updated `is_alluvia_form()` tests
+test_that("`is_alluvia_form()` recognizes alluvia-format Titanic data", {
+  expect_message(is_alluvia_form(titanic_alluvia), "[Mm]issing")
+  expect_true(is_alluvia_form(titanic_alluvia, axes = 1:4))
+  expect_true(is_alluvia_form(titanic_alluvia, axes = c("Class", "Sex")))
 })
 
 titanic_lodes <- suppressWarnings(tidyr::gather(
@@ -21,14 +18,12 @@ titanic_lodes$Value <- factor(titanic_lodes$Value,
                               levels = do.call(c, lapply(titanic_alluvia[, 1:4],
                                                          levels)))
 
-test_that("is_alluvial recognizes lodes-format Titanic data", {
-  expect_message(is_alluvial(titanic_lodes),
-                 "[Mm]issing")
-  expect_true(is_alluvial(titanic_lodes,
-                          key = "Variable", value = "Value", id = "Index"))
-  expect_true(is_alluvial(titanic_lodes,
-                          key = "Variable", value = "Value", id = "Index",
-                          weight = "Freq"))
-  expect_message(is_alluvial(titanic_lodes, axes = 3:4),
-                 "[Mm]issing")
+# updated `is_alluvia_form()` tests
+test_that("`is_lodes_form()` recognizes lodes-format Titanic data", {
+  expect_error(is_lodes_form(titanic_lodes))
+  expect_true(is_lodes_form(titanic_lodes,
+                            key = Variable, value = Value, id = Index))
+  expect_true(is_lodes_form(titanic_lodes,
+                            key = "Variable", value = "Value", id = "Index",
+                            weight = "Freq"))
 })

### 2018-05-17 inst/doc/ggalluvial.rmd
rebuild site

@@
 ```{r example alluvial diagram using Titanic dataset, echo=FALSE}
-ggplot(data = to_lodes(as.data.frame(Titanic),
+ggplot(data = to_lodes_form(as.data.frame(Titanic),
                             key = "Demographic",
@@
            weight = Freq, label = stratum)) +
+  scale_x_discrete(expand = c(.05, .05)) +
   geom_alluvium(aes(fill = Survived)) +
@@
 head(as.data.frame(UCBAdmissions), n = 12)
-is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE, silent = TRUE)
+is_alluvia_form(as.data.frame(UCBAdmissions), axes = 1:3, silent = TRUE)
 ```
@@
 This format requires an additional indexing column that links the rows corresponding to a common cohort, i.e.
-the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:
+the lodes of a single alluvium:
 
 ```{r lodes format of Berkeley admissions dataset}
-UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
+UCB_lodes <- to_lodes_form(as.data.frame(UCBAdmissions),
+                           axes = 1:3,
+                           id = "Cohort")
 head(UCB_lodes, n = 12)
-is_alluvial(UCB_lodes, logical = FALSE, silent = TRUE)
+is_lodes_form(UCB_lodes, key = x, value = stratum, id = Cohort, silent = TRUE)
 ```
@@
   scale_x_continuous(breaks = seq(2003, 2013, 2)) +
+  theme_bw() +
   theme(axis.text.x = element_text(angle = -30, hjust = 0)) +
@@
            fill = response, label = response)) +
+  scale_x_discrete(expand = c(.05, .05)) +
   geom_flow() +

### 2018-05-17 R/alluvial-data.r

@@ -82,9 +82,9 @@ is_lodes_form <- function(data,
                           logical = TRUE, silent = FALSE) {
   if (!isTRUE(logical)) deprecate_parameter("logical")
   
-  key_var <- tidyselect::vars_pull(names(data), !!rlang::enquo(key))
-  value_var <- tidyselect::vars_pull(names(data), !!rlang::enquo(value))
-  id_var <- tidyselect::vars_pull(names(data), !!rlang::enquo(id))
+  key_var <- vars_pull(names(data), !!rlang::enquo(key))
+  value_var <- vars_pull(names(data), !!rlang::enquo(value))
+  id_var <- vars_pull(names(data), !!rlang::enquo(id))
   
   if (any(duplicated(cbind(data[c(key_var, id_var)])))) {
     if (!silent) warning("Duplicated id-axis pairings.")
@@ -98,7 +98,7 @@ is_lodes_form <- function(data,
   
   # if `weight` is not `NULL`, use NSE to identify `weight_var`
   if (!is.null(rlang::enexpr(weight))) {
-    weight_var <- tidyselect::vars_select(names(data), !!rlang::enquo(weight))
+    weight_var <- vars_select(names(data), !!rlang::enquo(weight))
     if (!is.numeric(data[[weight_var]])) {
       if (!silent) message("Lode weights are non-numeric.")
       return(return(if (logical) FALSE else "none"))
@@ -122,7 +122,7 @@ is_alluvia_form <- function(data,
   if (is.null(rlang::enexpr(weight))) {
     weight_var <- NULL
   } else {
-    weight_var <- tidyselect::vars_select(names(data), !!rlang::enquo(weight))
+    weight_var <- vars_select(names(data), !!rlang::enquo(weight))
     if (!is.numeric(data[[weight_var]])) {
       if (!silent) message("Alluvium weights are non-numeric.")
       return(return(if (logical) FALSE else "none"))
@@ -133,15 +133,13 @@ is_alluvia_form <- function(data,
   }
   
   if (!is.null(rlang::enexpr(axes))) {
-    #deprecate_parameter("axes", NULL)
-    #axes <- ensure_vars(axes, data)
     axes <- data_at_vars(data, axes)
   } else {
     quos <- rlang::quos(...)
     if (rlang::is_empty(quos)) {
       axes <- setdiff(names(data), c(weight_var))
     } else {
-      axes <- unname(tidyselect::vars_select(names(data), !!!quos))
[…]

### 2018-05-23 inst/examples/ex-geom-alluvium.r
switch weight to y in examples + vignette

@@ -1,13 +1,13 @@
 # basic
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age,
            fill = Survived)) +
   geom_alluvium() +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 gg <- ggplot(alluvial::Refugees,
-             aes(weight = refugees, x = year, alluvium = country))
+             aes(y = refugees, x = year, alluvium = country))
 # time series bump chart
 gg + geom_alluvium(aes(fill = country, colour = country),
                    width = 1/4, alpha = 2/3, decreasing = FALSE)

### 2018-05-23 inst/examples/ex-stat-alluvium.r

@@ -1,6 +1,6 @@
 # illustrate positioning
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age,
            color = Survived)) +
   stat_stratum(geom = "errorbar") +
@@ -10,7 +10,7 @@ ggplot(as.data.frame(Titanic),
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 gg <- ggplot(as.data.frame(Titanic),
-             aes(weight = Freq,
+             aes(y = Freq,
                  axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
@@ -48,7 +48,7 @@ gg + geom_flow(stat = "alluvium", lode.guidance = "rightleft",
                       prop >= .01 & sex == "F" &
                         year > 1962 & year < 1968)
   ggplot(data = bn,
-         aes(x = year, alluvium = name, weight = prop)) +
+         aes(x = year, alluvium = name, y = prop)) +
     geom_alluvium(aes(fill = name, color = name == "Tammy"),
                   decreasing = TRUE, show.legend = FALSE) +
     scale_color_manual(values = c("#00000000", "#000000"))
@@ -58,7 +58,7 @@ gg + geom_flow(stat = "alluvium", lode.guidance = "rightleft",
                all = TRUE)
   bn2$prop[is.na(bn2$prop)] <- 0
   ggplot(data = bn2,
-         aes(x = year, alluvium = name, weight = prop)) +
+         aes(x = year, alluvium = name, y = prop)) +
     geom_alluvium(aes(fill = name, color = name == "Tammy"),
                   decreasing = TRUE, show.legend = FALSE) +
     scale_color_manual(values = c("#00000000", "#000000"))

### 2018-05-23 inst/examples/ex-stat-flow.r

@@ -1,6 +1,6 @@
 # illustrate positioning
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age,
            color = Survived)) +
   stat_stratum(geom = "errorbar") +
@@ -11,7 +11,7 @@ ggplot(as.data.frame(Titanic),
 
 # use of lode controls
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_flow(aes(fill = Survived), aes.bind = TRUE, reverse = FALSE) +
   geom_stratum(reverse = FALSE) +
@@ -21,7 +21,7 @@ ggplot(as.data.frame(Titanic),
 data(vaccinations)
 gg <- ggplot(vaccinations,
              aes(x = survey, stratum = response, alluvium = subject,
-                 weight = freq, fill = response)) +
+                 y = freq, fill = response)) +
   geom_stratum(alpha = .5) +
   geom_text(aes(label = response), stat = "stratum")
 # rightward alluvial aesthetics for vaccine survey data
@@ -36,7 +36,7 @@ vaccinations$subgroup <- LETTERS[1:2][rbinom(
 ) + 1][vaccinations$subject]
 ggplot(vaccinations,
        aes(x = survey, stratum = response, alluvium = subject,
-           weight = freq, fill = response, label = response)) +
+           y = freq, fill = response, label = response)) +
   geom_flow(aes(alpha = subgroup)) +
   scale_alpha_discrete(range = c(1/3, 2/3)) +
   geom_stratum(alpha = .5) +
@@ -44,7 +44,7 @@ ggplot(vaccinations,
 # can even set aesthetics that vary both ways
 ggplot(vaccinations,
        aes(x = survey, stratum = response, alluvium = subject,
-           weight = freq, label = response)) +
+           y = freq, label = response)) +
   geom_flow(aes(fill = interaction(response, subgroup)), aes.bind = TRUE) +
   scale_alpha_discrete(range = c(1/3, 2/3)) +
   geom_stratum(alpha = .5) +

### 2018-05-23 R/stat-alluvium.r
tweak vignette + replace aggregate.wts parameter with aggregate.y

@@ -15,10 +15,11 @@
 #'   \code{\link{geom_lode}}, and
 #'   \code{\link{geom_flow}} for the corresponding geoms.
 #' @inheritParams stat_flow
-#' @param aggregate.wts Whether to aggregate weights across otherwise equivalent
+#' @param aggregate.y Whether to aggregate weights across otherwise equivalent
 #'   rows before computing lode and flow positions. Set to \code{TRUE} to group
 #'   observations into cohorts.
 #'   \strong{Warning}: This is currently an expensive operation.
+#' @param aggregate.wts Deprecated alias for \code{aggregate.y}.
 #' @param lode.guidance The function to prioritize the axis variables for 
 #'   ordering the lodes within each stratum. Options are "zigzag", "rightleft",
 #'   "leftright", "rightward", and "leftward" (see 
@@ -38,7 +39,7 @@ stat_alluvium <- function(mapping = NULL,
                           decreasing = NA,
                           reverse = TRUE,
                           discern = FALSE,
-                          aggregate.wts = FALSE,
+                          aggregate.y = FALSE, aggregate.wts = NULL,
                           lode.guidance = "zigzag",
                           lode.ordering = NULL,
                           na.rm = FALSE,
@@ -57,7 +58,7 @@ stat_alluvium <- function(mapping = NULL,
       decreasing = decreasing,
       reverse = reverse,
       discern = discern,
-      aggregate.wts = FALSE,
+      aggregate.y = FALSE, aggregate.wts = NULL,
       lode.guidance = lode.guidance,
       lode.ordering = lode.ordering,
       na.rm = na.rm,
@@ -144,13 +145,17 @@ StatAlluvium <- ggproto(
   compute_panel = function(data, scales,
                            decreasing = NA, reverse = TRUE,
                            discern = FALSE,
-                           aggregate.wts = FALSE,
+                           aggregate.y = FALSE, aggregate.wts = NULL,
                            lode.guidance = "zigzag",
                            aes.bind = FALSE,
                            lode.ordering = NULL) {
     
     # aggregate weights over otherwise equivalent alluvia
-    if (aggregate.wts) data <- aggregate_along(data, "x", "alluvium", "y")
+    if (!is.null(aggregate.wts)) {
+      deprecate_parameter("aggregate.wts", "aggregate.y")
+      aggregate.y <- aggregate.wts
+    }
[…]

### 2018-05-23 inst/doc/ggalluvial.R

@@ -7,7 +7,7 @@ ggplot(data = to_lodes_form(as.data.frame(Titanic),
                             key = "Demographic",
                             axes = 1:3),
        aes(x = Demographic, stratum = stratum, alluvium = alluvium,
-           weight = Freq, label = stratum)) +
+           y = Freq, label = stratum)) +
   scale_x_discrete(expand = c(.05, .05)) +
   geom_alluvium(aes(fill = Survived)) +
   geom_stratum() + geom_text(stat = "stratum") +
@@ -20,7 +20,7 @@ is_alluvia_form(as.data.frame(UCBAdmissions), axes = 1:3, silent = TRUE)
 
 ## ----alluvial diagram of UC Berkeley admissions dataset------------------
 ggplot(as.data.frame(UCBAdmissions),
-       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
+       aes(y = Freq, axis1 = Gender, axis2 = Dept)) +
   geom_alluvium(aes(fill = Admit), width = 1/12) +
   geom_stratum(width = 1/12, fill = "black", color = "grey") +
   geom_label(stat = "stratum", label.strata = TRUE) +
@@ -30,7 +30,7 @@ ggplot(as.data.frame(UCBAdmissions),
 
 ## ----parallel sets plot of Titanic dataset-------------------------------
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Survived, axis2 = Sex, axis3 = Class)) +
   geom_alluvium(aes(fill = Class),
                 width = 0, knot.pos = 0, reverse = FALSE) +
@@ -64,7 +64,7 @@ country_regions <- c(
 )
 Refugees$region <- country_regions[Refugees$country]
 ggplot(data = Refugees,
-       aes(x = year, weight = refugees, alluvium = country)) +
+       aes(x = year, y = refugees, alluvium = country)) +
   geom_alluvium(aes(fill = country, colour = country),
                 alpha = .75, decreasing = FALSE) +
   scale_x_continuous(breaks = seq(2003, 2013, 2)) +
@@ -93,7 +93,7 @@ data(vaccinations)
 levels(vaccinations$response) <- rev(levels(vaccinations$response))
 ggplot(vaccinations,
        aes(x = survey, stratum = response, alluvium = subject,
-           weight = freq,
+           y = freq,
            fill = response, label = response)) +
   scale_x_discrete(expand = c(.05, .05)) +
   geom_flow() +

### 2018-05-23 inst/examples/ex-stat-alluvium.r

@@ -25,10 +25,17 @@ gg + geom_flow(aes(fill = Survived, alpha = Sex), stat = "alluvium",
 data(majors)
 # omit missing lodes and incident flows
 ggplot(majors,
-       aes(x = semester, stratum = curriculum, alluvium = student)) +
+             aes(x = semester, stratum = curriculum, alluvium = student)) +
   geom_alluvium(fill = "darkgrey", na.rm = TRUE) +
   geom_stratum(aes(fill = curriculum), color = NA, na.rm = TRUE) +
   theme_bw()
+# reverse the vertical axis (requires an explicit `y` aesthetic)
+ggplot(majors,
+       aes(x = semester, stratum = curriculum, alluvium = student, y = 1)) +
+  geom_alluvium(fill = "darkgrey", na.rm = TRUE) +
+  geom_stratum(aes(fill = curriculum), color = NA, na.rm = TRUE) +
+  theme_bw() +
+  scale_y_reverse()
 
 gg <- ggplot(majors,
              aes(x = semester, stratum = curriculum, alluvium = student,
@@ -39,7 +46,7 @@ gg + geom_flow(stat = "alluvium", lode.guidance = "rightleft",
                color = "black")
 # same diagram with students are aggregated into cohorts
 gg + geom_flow(stat = "alluvium", lode.guidance = "rightleft",
-               color = "black", aggregate.wts = TRUE)
+               color = "black", aggregate.y = TRUE)
 
 \dontrun{
   data(babynames, package = "babynames")

### 2018-06-02 README.rmd

--- 2018-02-05 README.rmd
@@
        aes(axis1 = Class, axis2 = Sex, axis3 = Age,
-           weight = Freq)) +
-  scale_x_discrete(limits = c("Class", "Sex", "Age")) +
+           y = Freq)) +
+  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
   geom_alluvium(aes(fill = Survived)) +
@@
        aes(x = Demographic, stratum = stratum, alluvium = alluvium,
-           weight = Freq, label = stratum)) +
+           y = Freq, label = stratum)) +
   geom_alluvium(aes(fill = Survived)) +

### 2018-06-02 inst/doc/ggalluvial.rmd

@@
        aes(x = Demographic, stratum = stratum, alluvium = alluvium,
-           weight = Freq, label = stratum)) +
+           y = Freq, label = stratum)) +
   scale_x_discrete(expand = c(.05, .05)) +
@@
 
-The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g.
-the number of observational units in the cohort.
+The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column.
+An additional column contains the weight of each row, e.g.
+the number of observational units in the cohort, which may be used to control the heights of the strata.[^weight-y] Basically, the wide format consists of _one row per alluvium_.
 This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:
@@
 ggplot(as.data.frame(UCBAdmissions),
-       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
+       aes(y = Freq, axis1 = Gender, axis2 = Dept)) +
   geom_alluvium(aes(fill = Admit), width = 1/12) +
@@
 ggplot(as.data.frame(Titanic),
-       aes(weight = Freq,
+       aes(y = Freq,
            axis1 = Survived, axis2 = Sex, axis3 = Class)) +
@@
 
-Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and often not plotted.
+Heights can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata contain no more information than the alluvia and often not plotted.
 For convenience, both `stat_alluvium()` and `stat_flow()` will accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:
@@
 ggplot(data = Refugees,
-       aes(x = year, weight = refugees, alluvium = country)) +
+       aes(x = year, y = refugees, alluvium = country)) +
   geom_alluvium(aes(fill = country, colour = country),
@@
 
-No `weight` is specified, so each row is assigned unit weight.
+The stratum heights `y` are unspecified, so each row is given unit height.
 This example demonstrates one way **ggalluvial** handles missing data.
@@
        aes(x = survey, stratum = response, alluvium = subject,
-           weight = freq,
+           y = freq,
            fill = response, label = response)) +
-  scale_x_discrete(expand = c(.05, .05)) +
+  scale_x_discrete(expand = c(.1, .1)) +
   geom_flow() +
@@
 ```
+
+[^weight-y]: Previously, weights were passed to the `weight` aesthetic rather than to `y`.
+This prevented `scale_y_continuous()` from correctly transforming scales, and anyway it was inconsistent with the behavior of `geom_bar()`.

### 2018-07-04 inst/doc/labels.rmd
re-build vignettes

---
title: "Labeling small strata"
author: "Jason Cory Brunson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{labeling small strata}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Setup

This brief vignette uses the `vaccinations` dataset included in **ggalluvial**. As in [the technical introduction](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html), the order of the levels is reversed to be more intuitive. Objects from other **ggplot2** extensions are accessed via `::` and `:::`.

```{r setup}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.align = "center")
library(ggalluvial)
data(vaccinations)
levels(vaccinations$response) <- rev(levels(vaccinations$response))
```

## Problem

The issue on the table: Strata are most helpful when they're overlaid with text labels. Yet the strata often vary in height, and the labels in length, to such a degree that fitting the text inside the strata at a uniform size renders them illegible. In principle, the user could treat `size` as a variable aesthetic and manually fit text to strata, but this is cumbersome, and doesn't help anyway in cases where large text is needed. Fortunately, two wonderful packages provide more elegant solutions.

To illustrate the problem, check out the diagram below. It's by no means an egregious case, but it'll do. (For a more practical example, see [this question on StackOverflow](https://stackoverflow.com/questions/50720718/labelling-and-theme-of-ggalluvial-plot-in-r), which prompted this vignette.)

```{r raw}
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject, y = freq,
           fill = response, label = response)) +
  scale_x_discrete(expand = c(.1, 0)) +
  geom_flow(width = 1/4) +
  geom_stratum(alpha = .5, width = 1/4) +
  geom_text(stat = "stratum", size = 4) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses", "labeled using `geom_text()`")
```

## Solutions

Two **ggplot2** extensions are well-suited to this problem: [**ggrepel**](https://github.com/slowkow/ggrepel) and [**ggfittext**](https://github.com/wilkox/ggfittext). They provide new geom layers that use the output of existing stat layers to situate text:
`ggrepel::geom_text_repel()` takes the same aesthetics as `ggplot2::geom_text()`, namely `x`, `y`, and `label`. In contrast, `ggfittext::geom_fit_text()` only specifically requires `label` but also needs enough information to determine the rectangle that will contain the text. This can be encoded as `xmin` and `xmax` or as `x` and `width` for the horizontal direction, and as `ymin` and `ymax` or as `y` and `height` for the vertical direction. Conveniently, `ggalluvial::stat_stratum()` produces more than enough information for both geoms, including `x`, `xmin`, `xmax`, and their vertical counterparts. All this can be gleaned from the `ggproto` objects that construct the layers:

```{r aesthetics}
print(ggrepel::GeomTextRepel$required_aes)
print(ggfittext:::GeomFitText$required_aes)
print(ggfittext:::GeomFitText$setup_data)
print(StatStratum$compute_panel)
```

I reached the specific solutions through trial and error. They may not be the best tricks for most cases, but they demonstrate what these packages can do. For many more examples, see the respective package vignettes: [for **ggrepel**](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html), and [for **ggfittext**](https://cran.r-project.org/web/packages/ggfittext/vignettes/introduction-to-ggfittext.html).

### Solution 1: **ggrepel**

**ggrepel** is most often (in my experience) used to repel text away from symbols in a scatterplot, in whatever directions effectively prevent them from overlapping the symbols and each other. In this case, however, it makes much more sense to align them vertically a fixed horizontal distance (`nudge_x`) away from the strata, repelling them vertically from each other (`direction = "y"`) just enough to print them without overlap. It takes an extra bit of effort to render text _only_ for the strata at the first (or at the last) axis, but the result is worth it.

```{r ggrepel}
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject, y = freq,
           fill = response)) +
  scale_x_discrete(expand = c(.4, 0)) +
  geom_flow(width = 1/4) +
  geom_stratum(alpha = .5, width = 1/4) +
  scale_linetype_manual(values = c("blank", "solid")) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(as.numeric(survey) == 1, as.character(response), NA)),
    stat = "stratum", size = 4, direction = "y", nudge_x = -.5
  ) +
  ggrepel::geom_text_repel(
    aes(label = ifelse(as.numeric(survey) == 3, as.character(response), NA)),
    stat = "stratum", size = 4, direction = "y", nudge_x = .5
  ) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses", "labeled using `geom_text_repel()`")
```

### Solution 2: **ggfittext**

**ggfittext** is simplicity itself: The stratum are just rectangles, so no more parameter specifications are necessary to fit the text into them. One key parameter is `min.size`, which defaults to `4` and controls how small the text is allowed to get without being omitted.

```{r ggfittext}
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject, y = freq,
           fill = response, label = response)) +
  scale_x_discrete(expand = c(.1, 0)) +
  geom_flow(width = 1/4) +
  geom_stratum(alpha = .5, width = 1/4) +
  ggfittext::geom_fit_text(stat = "stratum", width = 1/4, min.size = 3) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses", "labeled using `geom_fit_text()`")
```

Note that **ggfittext** in particular is under development, so the functionality may change in the future.

## Appendix

```{r session info}
devtools::session_info()
```

### 2018-07-06 README.rmd
issues/pulls links + pkre-build site

@@
 The
-[optimization branch](https://github.com/corybrunson/ggalluvial/tree/optimization) contains a development version with experimental functions to reduce the number or area of alluvial overlaps (see issue [#6](/../../issues/6)).
+[optimization branch](https://github.com/corybrunson/ggalluvial/tree/optimization) contains a development version with experimental functions to reduce the number or area of alluvial overlaps (see issue [#6](https://github.com/corybrunson/ggalluvial/issues/6)).
 Install it as follows:
@@
 
-Issues and pull requests are more than welcome!
+[Issues](https://github.com/corybrunson/ggalluvial/issues) and [pull requests](https://github.com/corybrunson/ggalluvial/pulls) are more than welcome!
 Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.

### 2018-10-11 README.rmd
add x-label to README

@@
   scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
+  xlab("Demographic") +
   geom_alluvium(aes(fill = Survived)) +

### 2018-10-24 README.rmd

@@
 ```{r eval=TRUE}
-titanic_long <- to_lodes(data.frame(Titanic),
+titanic_long <- to_lodes_form(data.frame(Titanic),
                               key = "Demographic",

### 2019-01-18 README.rmd
remotes pkg to install dev version - resolves #26

@@
 ```{r eval=FALSE}
-devtools::install_github("corybrunson/ggalluvial", build_vignettes = TRUE)
+remotes::install_github("corybrunson/ggalluvial", build_vignettes = TRUE)
 ```
@@
 ```{r eval=FALSE}
-devtools::install_github("corybrunson/ggalluvial", ref = "optimization")
+remotes::install_github("corybrunson/ggalluvial", ref = "optimization")
 ```

### 2019-01-19 inst/doc/labels.rmd
demo height bounds in labels vignette + update NEWS + v0.9.2

@@
 
+One option is to simply omit those labels that don't fit within their strata.
+In response to [an issue](https://github.com/corybrunson/ggalluvial/issues/27), `v0.9.2` includes parameters in `stat_stratum()` to exclude strata outside a specified height range; while few would use this to omit the rectangles themselves, it can be used in tandem with `geom_text()` to shirk this problem, at least when the labels are concise:
+
+```{r omit}
+ggplot(vaccinations,
+       aes(x = survey, stratum = response, alluvium = subject, y = freq,
+           fill = response, label = response)) +
+  scale_x_discrete(expand = c(.1, 0)) +
+  geom_flow(width = 1/4) +
+  geom_stratum(alpha = .5, width = 1/4) +
+  geom_text(stat = "stratum", size = 4, min.height = 100) +
+  theme(legend.position = "none") +
+  ggtitle(
+    "vaccination survey responses",
+    "labeled using `geom_text()` with `min.height = 100`"
+  )
+```
+
 ## Solutions
@@
 Conveniently, `ggalluvial::stat_stratum()` produces more than enough information for both geoms, including `x`, `xmin`, `xmax`, and their vertical counterparts.
+
 All this can be gleaned from the `ggproto` objects that construct the layers:
@@
 They may not be the best tricks for most cases, but they demonstrate what these packages can do.
-For many more examples, see the respective package vignettes: [for **ggrepel**](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html), and [for **ggfittext**](https://cran.r-project.org/web/packages/ggfittext/vignettes/introduction-to-ggfittext.html).
+For many more examples, see the respective package vignettes: [for **ggrepel**](https://CRAN.R-project.org/package=ggrepel/vignettes/ggrepel.html), and [for **ggfittext**](https://CRAN.R-project.org/package=ggfittext/vignettes/introduction-to-ggfittext.html).
 
@@
 
-**ggrepel** is most often (in my experience) used to repel text away from symbols in a scatterplot, in whatever directions effectively prevent them from overlapping the symbols and each other.
-In this case, however, it makes much more sense to align them vertically a fixed horizontal distance (`nudge_x`) away from the strata, repelling them vertically from each other (`direction = "y"`) just enough to print them without overlap.
+**ggrepel** is most often (in my experience) used to repel text away from symbols in a scatterplot, in whatever directions prevent them from overlapping the symbols and each other.
+In this case, however, it makes much more sense to align them vertically a fixed horizontal distance (`nudge_x`) away from the strata and repel them vertically from each other (`direction = "y"`) just enough to print them without overlap.
 It takes an extra bit of effort to render text _only_ for the strata at the first (or at the last) axis, but the result is worth it.
@@
 
-**ggfittext** is simplicity itself: The stratum are just rectangles, so no more parameter specifications are necessary to fit the text into them.
+**ggfittext** is simplicity itself: The strata are just rectangles, so no more parameter specifications are necessary to fit the text into them.
 One key parameter is `min.size`, which defaults to `4` and controls how small the text is allowed to get without being omitted.
@@
 
-Note that **ggfittext** in particular is under development, so the functionality may change in the future.
+Note that this solution requires **ggfittext** v0.6.0.
 
@@
 ```{r session info}
-devtools::session_info()
+sessioninfo::session_info()
 ```

### 2019-01-19 inst/doc/labels.R

@@ -15,6 +15,20 @@ ggplot(vaccinations,
   theme(legend.position = "none") +
   ggtitle("vaccination survey responses", "labeled using `geom_text()`")
 
+## ----omit----------------------------------------------------------------
+ggplot(vaccinations,
+       aes(x = survey, stratum = response, alluvium = subject, y = freq,
+           fill = response, label = response)) +
+  scale_x_discrete(expand = c(.1, 0)) +
+  geom_flow(width = 1/4) +
+  geom_stratum(alpha = .5, width = 1/4) +
+  geom_text(stat = "stratum", size = 4, min.height = 100) +
+  theme(legend.position = "none") +
+  ggtitle(
+    "vaccination survey responses",
+    "labeled using `geom_text()` with `min.height = 100`"
+  )
+
 ## ----aesthetics----------------------------------------------------------
 print(ggrepel::GeomTextRepel$required_aes)
 print(ggfittext:::GeomFitText$required_aes)

### 2019-07-02 README.rmd

@@
 
-<!--
+[![Travis build status](https://travis-ci.org/corybrunson/ggalluvial.svg?branch=master)](https://travis-ci.org/corybrunson/ggalluvial)
 [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
--->
 

### 2019-09-02 inst/examples/ex-stat-flow.r
rename lode-guidance functions

@@ -25,7 +25,7 @@ gg <- ggplot(vaccinations,
   geom_stratum(alpha = .5) +
   geom_text(aes(label = response), stat = "stratum")
 # rightward alluvial aesthetics for vaccine survey data
-gg + geom_flow(stat = "alluvium", lode.guidance = "rightward")
+gg + geom_flow(stat = "alluvium", lode.guidance = "forward")
 # memoryless flows for vaccine survey data
 gg + geom_flow()
 

### 2019-09-02 R/stat-alluvium.r
drop position variable from stat-alluvium output

@@ -235,16 +235,16 @@ StatAlluvium <- ggproto(
     # gather lode positions into alluvium-axis-order table
     alluv_pos <- tidyr::gather(
       alluv,
-      key = "x", value = "position",
+      key = "x", value = "pos",
       alluv_ind
     )
     rm(alluv) # avoid confusion
     alluv_pos$x <- as.integer(alluv_pos$x)
-    # join position variable into `data`
+    # join 'pos' variable into `data`
     data <- dplyr::left_join(data, alluv_pos, by = c("x", "alluvium"))
     
     # calculate lode floors and ceilings from positions by axis
-    data <- data[order(data$position), , drop = FALSE]
+    data <- data[order(data$pos), , drop = FALSE]
     data <- dplyr::ungroup(dplyr::mutate(dplyr::group_by(data, x),
                                          ymax = cumsum(y),
                                          ymin = dplyr::lag(ymax, default = 0)))
@@ -264,6 +264,7 @@ StatAlluvium <- ggproto(
     # add `group` to group contiguous alluvial subsets
     data <- transform(data, group = as.numeric(interaction(alluvium, flow)))
     # remove unused fields
+    data$pos <- NULL
     data$cont <- NULL
     data$axis <- NULL
     data$flow <- NULL

### 2019-09-05 README.rmd
acknowledge UConn Health + re-build home

@@
 
-```{r eval=TRUE, echo=FALSE, message=FALSE}
+```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
 devtools::load_all()
@@
 
-### Resources
+### Documentation
 
-For detailed discussion of the data formats recognized by **ggalluvial** and several examples that illustrate its flexibility and limitations, read the vignette:
+For detailed discussion of the data formats recognized by **ggalluvial** and several examples that illustrate its flexibility and limitations, read the technical vignette:
 
@@
 
-The documentation contains several examples; use `help()` to call forth examples of any layer (`stat_*` or `geom_*`).
+The documentation contains several examples; use `help()` to call forth examples of any layer (`stat_*` or `geom_*`) or of the conversion functions (`to_*_form`).
 
-## Feedback
+## Acknowledgment
 
-### Cite
+### Resources
 
-If you use **ggalluvial**-generated figures in publication, i'd be grateful to hear about it! You can also cite the package according to `citation("ggalluvial")`.
+Development of this package benefitted from the use of equipment and the support of colleagues at [UConn Health](https://health.uconn.edu/).
 
@@
 Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.
+
+### Cite
+
+If you use **ggalluvial**-generated figures in publication, i'd be grateful to hear about it! You can also cite the package according to `citation("ggalluvial")`.

### 2019-10-01 inst/examples/ex-stat-flow.r
deprecate min-height to min-y + extend min-y to stat-flow & stat-alluvium - resolves #39

@@ -29,6 +29,23 @@ gg + geom_flow(stat = "alluvium", lode.guidance = "forward")
 # memoryless flows for vaccine survey data
 gg + geom_flow()
 
+# omit small flows
+ggplot(vaccinations,
+       aes(y = freq,
+           x = survey, stratum = response, alluvium = subject,
+           fill = response, label = response)) +
+  stat_stratum(alpha = .5) +
+  geom_flow(min.y = 50) +
+  geom_text(stat = "stratum")
+# omit large flows
+ggplot(vaccinations,
+       aes(y = freq,
+           x = survey, stratum = response, alluvium = subject,
+           fill = response, label = response)) +
+  stat_stratum(alpha = .5) +
+  geom_flow(max.y = 100) +
+  geom_text(stat = "stratum")
+
 # aesthetics that vary betwween and within strata
 data(vaccinations)
 vaccinations$subgroup <- LETTERS[1:2][rbinom(

### 2019-10-01 inst/examples/ex-stat-stratum.r

@@ -43,4 +43,4 @@ ggplot(vaccinations,
            x = survey, stratum = response,
            fill = response, label = response)) +
   stat_stratum(width = .5) +
-  geom_text(stat = "stratum", min.height = 100)
+  geom_text(stat = "stratum", min.y = 100)

### 2019-11-22 README.rmd
change 'diagram' to 'plot' + debug infer-label param

@@
 
-This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial diagrams.
+This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial plots.
 
@@
 The alluvial plots implemented here can be used to visualize frequency distributions over time or frequency tables involving several categorical variables.
-The design is derived mostly from the [**alluvial**](https://github.com/mbojan/alluvial) package, but the **ggplot2** framework induced several conspicuous differences:
+The design is inspired by the [**alluvial**](https://github.com/mbojan/alluvial) package, but the **ggplot2** framework induced several conspicuous differences:
 
-- **alluvial** understands a variety of inputs (vectors, lists, data frames), while **ggalluvial** requires a single data frame;
+- **alluvial** understands a variety of inputs (vectors, lists, data frames), whereas **ggalluvial** requires a single data frame;
 - **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](http://tidyr.tidyverse.org/);
 - **alluvial** produces both the *alluvia*, which link cohorts across multiple dimensions, and (what are here called) the *strata*, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called *lodes* and *flows*.
+
+Additionally, **ggalluvial** arranges these layers vertically without gaps, so that the secondary plotting axis indicates the cumulative values of the strata at each dimension.
 
@@
 ```
+Note, however, that this branch has not kept pace with the `master` branch or with recent upgrades on CRAN.
 
@@
 
-Here is how to generate an alluvial diagram representation of the multi-dimensional categorical dataset of passengers on the Titanic:
+Here is how to generate an alluvial plot representation of the multi-dimensional categorical dataset of passengers on the Titanic:
 

### 2019-11-22 R/alluvial-data.r

@@ -1,6 +1,6 @@
 #' Check for alluvial structure and convert between alluvial formats
 #'
-#' Alluvial diagrams consist of multiple horizontally-distributed columns (axes)
+#' Alluvial plots consist of multiple horizontally-distributed columns (axes)
 #' representing factor variables, vertical divisions (strata) of these axes
 #' representing these variables' values; and splines (alluvial flows) connecting
 #' vertical subdivisions (lodes) within strata of adjacent axes representing
@@ -19,14 +19,14 @@
 #'   for this structure.
 #'
 #' `to_lodes_form` takes a data frame with several designated variables to
-#' be used as axes in an alluvial diagram, and reshapes the data frame so that
+#' be used as axes in an alluvial plot, and reshapes the data frame so that
 #' the axis variable names constitute a new factor variable and their values
 #' comprise another. Other variables' values will be repeated, and a
 #' row-grouping variable can be introduced. This function invokes
 #' [tidyr::gather()].
 #'
 #' `to_alluvia_form` takes a data frame with axis and axis value variables
-#' to be used in an alluvial diagram, and reshape the data frame so that the
+#' to be used in an alluvial plot, and reshape the data frame so that the
 #' axes constitute separate variables whose values are given by the value
 #' variable. This function invokes [tidyr::spread()].
 #'

### 2019-11-22 R/geom-lode.r

@@ -1,7 +1,7 @@
 #' Lodes at intersections of alluvia and strata
 #'
 #' `geom_alluvium` receives a dataset of the horizontal (`x`) and vertical (`y`,
-#' `ymin`, `ymax`) positions of the **lodes** of an alluvial diagram, the
+#' `ymin`, `ymax`) positions of the **lodes** of an alluvial plot, the
 #' intersections of the alluvia with the strata. It plots rectangles for these
 #' lodes of a provided `width`.
 #' @template geom-aesthetics

### 2019-11-22 R/geom-stratum.r

@@ -1,7 +1,7 @@
 #' Strata at axes
 #'
 #' `geom_stratum` receives a dataset of the horizontal (`x`) and vertical (`y`,
-#' `ymin`, `ymax`) positions of the strata of an alluvial diagram. It plots
+#' `ymin`, `ymax`) positions of the strata of an alluvial plot. It plots
 #' rectangles for these strata of a provided `width`.
 #' @template geom-aesthetics
 #' @template defunct-geom-params

### 2019-11-22 R/stat-alluvium.r

@@ -43,7 +43,7 @@ stat_alluvium <- function(mapping = NULL,
                           lode.guidance = "zigzag",
                           lode.ordering = NULL,
                           aes.bind = "none",
-                          overlay.label = FALSE,
+                          infer.label = FALSE,
                           min.y = NULL, max.y = NULL,
                           na.rm = FALSE,
                           show.legend = NA,
@@ -65,7 +65,7 @@ stat_alluvium <- function(mapping = NULL,
       lode.guidance = lode.guidance,
       lode.ordering = lode.ordering,
       aes.bind = aes.bind,
-      overlay.label = overlay.label,
+      infer.label = infer.label,
       min.y = min.y, max.y = max.y,
       na.rm = na.rm,
       ...
@@ -164,17 +164,17 @@ StatAlluvium <- ggproto(
                            aggregate.y = NULL, cement.alluvia = FALSE,
                            lode.guidance = "zigzag",
                            aes.bind = "none",
-                           overlay.label = FALSE,
+                           infer.label = FALSE,
                            min.y = NULL, max.y = NULL,
                            lode.ordering = NULL) {
     
     # introduce label
-    if (overlay.label) {
+    if (infer.label) {
       if (is.null(data$label)) {
         data$label <- data$alluvium
       } else {
         warning("Aesthetic `label` is specified, ",
-                "so parameter `overlay.label` will be ignored.")
+                "so parameter `infer.label` will be ignored.")
       }
     }
     

### 2019-11-22 inst/examples/ex-alluvial-data.r

@@ -48,11 +48,11 @@ gg <- ggplot(majors_alluvia,
 gg +
   geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = TRUE) +
   geom_stratum(width = 2/5, discern = TRUE) +
-  geom_text(stat = "stratum", discern = TRUE, overlay.label = TRUE)
+  geom_text(stat = "stratum", discern = TRUE, infer.label = TRUE)
 gg +
   geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = FALSE) +
   geom_stratum(width = 2/5, discern = FALSE) +
-  geom_text(stat = "stratum", discern = FALSE, overlay.label = TRUE)
+  geom_text(stat = "stratum", discern = FALSE, infer.label = TRUE)
 # warning when inappropriate
 ggplot(majors[majors$semester %in% paste0("CURR", c(1, 7, 13)), ],
        aes(x = semester, stratum = curriculum, alluvium = student,

### 2019-11-22 inst/examples/ex-geom-flow.r

@@ -4,8 +4,8 @@ ggplot(as.data.frame(Titanic),
            axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_flow() +
   scale_x_discrete(limits = c("Class", "Sex", "Age")) +
-  geom_stratum() + geom_text(stat = "stratum", overlay.label = TRUE) +
-  ggtitle("Alluvial diagram of Titanic passenger demographic data")
+  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
+  ggtitle("Alluvial plot of Titanic passenger demographic data")
 
 # use of facets
 ggplot(as.data.frame(Titanic),
@@ -13,7 +13,7 @@ ggplot(as.data.frame(Titanic),
            axis1 = Class, axis2 = Sex)) +
   geom_flow(aes(fill = Age), width = .4) +
   geom_stratum(width = .4) +
-  geom_text(stat = "stratum", overlay.label = TRUE, size = 3) +
+  geom_text(stat = "stratum", infer.label = TRUE, size = 3) +
   scale_x_discrete(limits = c("Class", "Sex")) +
   facet_wrap(~ Survived, scales = "fixed")
 

### 2019-11-22 inst/examples/ex-geom-stratum.r

@@ -2,7 +2,7 @@
 ggplot(as.data.frame(Titanic),
        aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age, axis4 = Survived)) +
-  geom_stratum(width = 1) + geom_text(stat = "stratum", overlay.label = TRUE) +
+  geom_stratum(width = 1) + geom_text(stat = "stratum", infer.label = TRUE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age", "Survived"))
   
 # use of facets
@@ -10,6 +10,6 @@ ggplot(as.data.frame(Titanic),
        aes(y = Freq,
            axis1 = Class, axis2 = Sex)) +
   geom_flow(aes(fill = Survived)) +
-  geom_stratum() + geom_text(stat = "stratum", overlay.label = TRUE) +
+  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
   scale_x_discrete(limits = c("Class", "Sex")) +
   facet_wrap(~ Age, scales = "free_y")

### 2019-11-22 inst/examples/ex-stat-flow.r

@@ -6,7 +6,7 @@ ggplot(as.data.frame(Titanic),
   stat_stratum(geom = "errorbar") +
   geom_line(stat = "flow") +
   stat_flow(geom = "pointrange") +
-  geom_text(stat = "stratum", overlay.label = TRUE) +
+  geom_text(stat = "stratum", infer.label = TRUE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 # use of lode controls
@@ -15,7 +15,7 @@ ggplot(as.data.frame(Titanic),
            axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_flow(aes(fill = Survived), aes.bind = TRUE, reverse = FALSE) +
   geom_stratum(reverse = FALSE) +
-  geom_text(stat = "stratum", overlay.label = TRUE, reverse = FALSE) +
+  geom_text(stat = "stratum", infer.label = TRUE, reverse = FALSE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 data(vaccinations)

### 2019-11-22 inst/examples/ex-stat-stratum.r

@@ -33,7 +33,7 @@ ggplot(as.data.frame(Titanic),
 ggplot(as.data.frame(Titanic),
        aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age, axis4 = Survived)) +
-  geom_text(stat = "stratum", overlay.label = TRUE) +
+  geom_text(stat = "stratum", infer.label = TRUE) +
   stat_stratum(geom = "errorbar") +
   scale_x_discrete(limits = c("Class", "Sex", "Age", "Survived"))
 

### 2019-11-24 R/stat-alluvium.r
debug aes-bind + increment sub-version + stat tests

@@ -193,12 +193,12 @@ StatAlluvium <- ggproto(
     # match arguments for `aes.bind`
     if (! is.null(aes.bind)) {
       if (is.logical(aes.bind)) {
-        aes.bind.rep <- if (aes.bind) "index" else "linked"
+        aes.bind.rep <- if (aes.bind) "flow" else "none"
         warning("Logical values of `aes.bind` are deprecated; ",
                 "replacing ", aes.bind, " with '", aes.bind.rep, "'.")
         aes.bind <- aes.bind.rep
       }
-      aes.bind <- match.arg(aes.bind, c("none", "linked", "index"))
+      aes.bind <- match.arg(aes.bind, c("none", "flows", "alluvia"))
     }
     
     # sign variable (sorts positives before negatives)
@@ -298,9 +298,9 @@ StatAlluvium <- ggproto(
     sort_fields <- c(
       "x",
       "deposit",
-      if (aes.bind == "index") "fissure",
+      if (aes.bind == "alluvia") "fissure",
       "rem_deposit",
-      if (aes.bind == "linked") "fissure",
+      if (aes.bind == "flows") "fissure",
       "fan"
     )
     data <- data[do.call(order, data[, sort_fields]), , drop = FALSE]

### 2019-11-24 R/stat-flow.r

@@ -15,16 +15,17 @@
 #' @inheritParams stat_stratum
 #' @param aes.bind At what grouping level, if any, to prioritize differentiation
 #'   aesthetics when ordering the lodes within each stratum. Defaults to
-#'   `"none"` (no aesthetic binding) with intermediate option `"linked"` to bind
+#'   `"none"` (no aesthetic binding) with intermediate option `"flows"` to bind
 #'   aesthetics after stratifying by axes linked to the index axis (the one
 #'   adjacent axis in `stat_flow()`; all remaining axes in `stat_alluvium()`)
-#'   and strongest option `"index"` to bind aesthetics after stratifying by the
-#'   index axis but before stratifying by linked axes. Stratification by any
-#'   axis is done with respect to the strata at that axis, after separating
-#'   positive and negative strata, consistent with the values of `decreasing`,
-#'   `reverse`, and `absolute`. Thus, if `"none"`, then lode orderings will not
-#'   depend on aesthetic variables. All aesthetic variables are used, in the
-#'   order in which they are specified in `aes()`.
+#'   and strongest option `"alluvia"` to bind aesthetics after stratifying by
+#'   the index axis but before stratifying by linked axes (only available for
+#'   `stat_alluvium()`). Stratification by any axis is done with respect to the
+#'   strata at that axis, after separating positive and negative strata,
+#'   consistent with the values of `decreasing`, `reverse`, and `absolute`.
+#'   Thus, if `"none"`, then lode orderings will not depend on aesthetic
+#'   variables. All aesthetic variables are used, in the order in which they are
+#'   specified in `aes()`.
 #' @example inst/examples/ex-stat-flow.r
 #' @export
 stat_flow <- function(mapping = NULL,
@@ -158,12 +159,16 @@ StatFlow <- ggproto(
     # match arguments for `aes.bind`
     if (! is.null(aes.bind)) {
       if (is.logical(aes.bind)) {
-        aes.bind.rep <- if (aes.bind) "index" else "linked"
+        aes.bind.rep <- if (aes.bind) "flow" else "none"
         warning("Logical values of `aes.bind` are deprecated; ",
                 "replacing ", aes.bind, " with '", aes.bind.rep, "'.")
         aes.bind <- aes.bind.rep
       }
-      aes.bind <- match.arg(aes.bind, c("none", "linked", "index"))
[…]

### 2019-11-24 inst/examples/ex-stat-flow.r

@@ -13,7 +13,7 @@ ggplot(as.data.frame(Titanic),
 ggplot(as.data.frame(Titanic),
        aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age)) +
-  geom_flow(aes(fill = Survived), aes.bind = TRUE, reverse = FALSE) +
+  geom_flow(aes(fill = Survived), aes.bind = "alluvia", reverse = FALSE) +
   geom_stratum(reverse = FALSE) +
   geom_text(stat = "stratum", infer.label = TRUE, reverse = FALSE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
@@ -62,7 +62,7 @@ ggplot(vaccinations,
 ggplot(vaccinations,
        aes(x = survey, stratum = response, alluvium = subject,
            y = freq, label = response)) +
-  geom_flow(aes(fill = interaction(response, subgroup)), aes.bind = TRUE) +
+  geom_flow(aes(fill = interaction(response, subgroup)), aes.bind = "alluvia") +
   scale_alpha_discrete(range = c(1/3, 2/3)) +
   geom_stratum(alpha = .5) +
   geom_text(stat = "stratum")

### 2019-11-24 tests/testthat/test-stat-alluvium.r

@@ -29,27 +29,27 @@ test_that("`stat_alluvium` orders alluvia correctly according to `aes.bind`", {
   data$stratum <- LETTERS[1:2][c(1, 1, 2, 2, 2, 2, 2, 1)]
   data$y <- 1
   data$fill <- c("red", "blue", "blue", "blue")
-  # order by index strata, linked strata, and alluvium
+  # order by index strata, linked strata (flows), and alluvia
   #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
   #  geom_alluvium(aes(fill = fill)) +
   #  geom_text(stat = "alluvium", aes(fill = fill, label = alluvium))
   comp <- StatAlluvium$compute_panel(data)
   expect_identical(comp[with(comp, order(x, alluvium)), ]$y,
                    c(3.5, 2.5, 0.5, 1.5, 2.5, 1.5, 0.5, 3.5))
-  # order by index strata, linked strata, aesthetics, and alluvium
+  # order by index strata, linked strata (flows), aesthetics, and alluvia
   #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
-  #  geom_alluvium(aes(fill = fill), aes.bind = "linked") +
+  #  geom_alluvium(aes(fill = fill), aes.bind = "flows") +
   #  geom_text(stat = "alluvium", aes(fill = fill, label = alluvium),
-  #            aes.bind = "linked")
-  comp <- StatAlluvium$compute_panel(data, aes.bind = "linked")
+  #            aes.bind = "flows")
+  comp <- StatAlluvium$compute_panel(data, aes.bind = "flows")
   expect_identical(comp[with(comp, order(x, alluvium)), ]$y,
                    c(2.5, 3.5, 0.5, 1.5, 1.5, 2.5, 0.5, 3.5))
-  # order by index strata, aesthetics, linked strata, and alluvium
+  # order by index strata, aesthetics, linked strata (flows), and alluvia
   #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
-  #  geom_alluvium(aes(fill = fill), aes.bind = "index") +
+  #  geom_alluvium(aes(fill = fill), aes.bind = "alluvia") +
   #  geom_text(stat = "alluvium", aes(fill = fill, label = alluvium),
-  #            aes.bind = "index")
[…]

### 2019-11-24 tests/testthat/test-stat-flow.r

@@ -1,0 +1,51 @@
+context("stat-flow")
+
+# reverse and absolute, negative values
+
+test_that("`stat_flow` orders flows correctly with negative values", {
+  data <- expand.grid(alluvium = letters[1:3], x = 1:2)
+  data$stratum <- LETTERS[c(1, 1, 2)]
+  data$y <- c(1, 1, 1, -1, 1, -1)
+  # order by stratum, `reverse = TRUE`
+  #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
+  #  geom_flow()
+  comp <- StatFlow$compute_panel(data)
+  expect_identical(comp[with(comp, order(x, alluvium)), ]$y,
+                   c(1.5, 0.5, 2.5, -1.5, -0.5, 0.5))
+  # order by stratum, `reverse = FALSE`
+  #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
+  #  geom_flow(reverse = FALSE)
+  comp <- StatFlow$compute_panel(data, reverse = FALSE)
+  expect_identical(comp[with(comp, order(x, alluvium)), ]$y,
+                   c(2.5, 0.5, 1.5, -1.5, -0.5, 0.5))
+  # order by stratum, `reverse = FALSE, absolute = FALSE`
+  #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
+  #  geom_flow(reverse = FALSE, absolute = FALSE)
+  comp <- StatFlow$compute_panel(data, reverse = FALSE, absolute = FALSE)
+  expect_identical(comp[with(comp, order(x, alluvium)), ]$y,
+                   c(0.5, 2.5, 1.5, -1.5, -0.5, 0.5))
+})
+
+# aesthetic binding
+
+test_that("`stat_flow` orders alluvia correctly according to `aes.bind`", {
+  data <- expand.grid(alluvium = letters[1:4], x = 1:2)
+  data$stratum <- LETTERS[c(1, 1, 1, 2, 2, 1, 3, 1)]
+  data$y <- 1
+  data$fill <- c("red", "blue", "blue", "blue",
+                 "red", "blue", "blue", "blue")
+  # order by index strata and linked strata (flows)
+  #ggplot(data, aes(x = x, stratum = stratum, alluvium = alluvium, y = y)) +
+  #  geom_flow(aes(fill = fill))
[…]

### 2019-11-24 tests/testthat/test-stat-stratum.r

@@ -1,0 +1,37 @@
+context("stat-stratum")
+
+# reverse and absolute parameters, negative values
+
+test_that("`stat_stratum` orders strata correctly with negative values", {
+  data <- expand.grid(stratum = LETTERS[1:2], x = 1:2)
+  data$y <- c(1, 1, -1, -1)
+  # order by stratum, `reverse = TRUE`
+  #ggplot(data, aes(x = x, stratum = stratum, y = y)) +
+  #  geom_stratum() +
+  #  geom_text(stat = "stratum", aes(label = stratum))
+  comp <- StatStratum$compute_panel(data)
+  expect_identical(comp[with(comp, order(x, stratum)), ]$y,
+                   c(1.5, 0.5, -1.5, -0.5))
+  # order by stratum, `reverse = FALSE`
+  #ggplot(data, aes(x = x, stratum = stratum, y = y)) +
+  #  geom_stratum(reverse = FALSE) +
+  #  geom_text(stat = "stratum", aes(label = stratum), reverse = FALSE)
+  comp <- StatStratum$compute_panel(data, reverse = FALSE)
+  expect_identical(comp[with(comp, order(x, stratum)), ]$y,
+                   c(0.5, 1.5, -0.5, -1.5))
+  # order by stratum, `absolute = FALSE`
+  #ggplot(data, aes(x = x, stratum = stratum, y = y)) +
+  #  geom_stratum(absolute = FALSE) +
+  #  geom_text(stat = "stratum", aes(label = stratum), absolute = FALSE)
+  comp <- StatStratum$compute_panel(data, absolute = FALSE)
+  expect_identical(comp[with(comp, order(x, stratum)), ]$y,
+                   c(1.5, 0.5, -0.5, -1.5))
+  # order by stratum, `reverse = FALSE, absolute = FALSE`
+  #ggplot(data, aes(x = x, stratum = stratum, y = y)) +
+  #  geom_stratum(reverse = FALSE, absolute = FALSE) +
+  #  geom_text(stat = "stratum", aes(label = stratum),
+  #            reverse = FALSE, absolute = FALSE)
+  comp <- StatStratum$compute_panel(data, reverse = FALSE, absolute = FALSE)
+  expect_identical(comp[with(comp, order(x, stratum)), ]$y,
+                   c(0.5, 1.5, -1.5, -0.5))
+})

### 2019-12-03 README.rmd

@@
 
-Development versions can be installed from GitHub:
+The development version can be installed from GitHub:
 ```{r eval=FALSE}
@@
 ```
+Note that, in order to build the vignettes, the imported packages **alluvial**, **ggfittext**, and **ggrepel** must be installed.
+To skip this step, leave `build_vignettes` unspecified or set it to `FALSE`.
+
 The
@@
 
-## Acknowledgment
+## Acknowledgments
 

### 2019-12-05 paper/paper.rmd
refine joss paper

---
title: "ggalluvial: A Layered Grammar for Alluvial Plots"
tags:
  - R
  - ggplot2
  - alluvial plots
  - statistical graphics
  - data visualization
  - repeated measures data
  - categorical data
authors:
  - name: Jason Cory Brunson
    orcid: 0000-0003-3126-9494
    affiliation: "1"
affiliations:
 - name: Center for Quantitative Medicine, UConn Health
   index: 1
date: 2019-12-05
bibliography: paper.bib
---

<!--
To generate `paper.md`, execute this from the console:
```r
knitr::knit("paper/paper.rmd", "paper/paper.md")
```
-->

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "paper/figures/"
)
```

# Summary

Alluvial plots use variable-width ribbons and stacked bar plots to represent multi-dimensional or repeated-measures data with categorical or ordinal variables [@Rosvall2010;@Bojanowski2016]. The **ggalluvial** package extends the layered grammar of graphics [@Wickham2016] to generate alluvial plots from tidy data [@Wickham2014].
The package makes two key contributions to the R ecosystem, which are summarized here and discussed in more detail below.

First, **ggalluvial** anchors the somewhat amorphous notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2005], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g. with the text and point--line geoms of **ggplot2**. Moreover, the package gives users a high degree of control over the underlying statistical transformation, whereas other extensions are not so flexible.[^easy] As an extention of **ggplot2**, the package also introduces alluvial plots to a **tidyverse** workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow **tidyr** and **dplyr** conventions.

Second, **ggalluvial** adopts a set of conventions meant to cement some of the concepts and terms that are inconsistently applied by developers and users of Sankey diagrams, parallel sets plots, and alluvial diagrams or plots.[^startup]
Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types. Instead, **ggalluvial** anchors its graphical elements to a distinctive, geological nomenclature. While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it. I hope that both the conventions and the terms can serve as useful points of reference as these visualization tools converge toward common standards.

[^easy]: Indeed, the dependency package **easyalluvial** [@Koneswarakantha2019] was built on top of **ggalluvial** to exchange much of this flexibility for expedited data exploration using alluvial plots.
[^startup]: Indeed, the diagrams that came to be known as parallel sets plots [@Kosara2006], consisting of parallel sequences of rectangles connected by parallelogramic ribbons, were anticipated by the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005].

## Functionality

[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of **ggalluvial**, and the reader is encouraged to browse [the documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations. In brief, the package contains stat and geom functions to add the following layers to a **ggplot2** object:

* _strata_, or stacked bar plots, located at each of multiple horizontal positions (_axes_ or _dimensions_)
* _alluvia_, ribbons through strata at different axes that encode individual cases or cohorts measured along each axis
* _lodes_, subdivisions of strata by their intersections with alluvia
* _flows_, segments of alluvia between adjacent axes (excluding lodes)

The layers of a **ggplot2** graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation. Default pairings, other within-package pairings, and pairings of alluvial stats with other geoms are differentiated by fill color:

```{r documented pairings with alluvial stats and geoms, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
library(ggalluvial)
combos <- data.frame(
  stat = c(
    rep("stratum", 2),
    rep("alluvium", 4),
    "flow",
    # text geom
    c("stratum", "alluvium", "flow"),
    # errorbar and pointrange geoms
    "stratum", "alluvium", "flow"
  ),
  geom = c(
    c("stratum", "lode"),
    c("stratum", "alluvium", "lode", "flow"),
    "flow",
    rep("text", 3),
    "errorbar", "pointrange", "pointrange"
  ),
  pairing = c(
    c("default", "alluvial"),
    c("alluvial", "default", "default", "alluvial"),
    "default",
    rep("external", 6)
  )
)
levs <- c("stratum", "alluvium", "lode", "flow",
          "text", "errorbar", "pointrange")
combos <- transform(
  combos,
  stat = factor(stat, levs),
  geom = factor(geom, levs),
  pairing = factor(pairing, c("default", "alluvial", "external"))
)
ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = pairing)) +
  scale_fill_brewer(type = "qual") +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  theme_minimal()
```

Note that the alluvial geoms are only paired with alluvial stats. This is because these geoms require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format. Because these aesthetics are not recognized by **ggplot2**, they produce warnings under some conditions. The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of **tidyr**.

Most of the stat parameters control how the strata at each axis, and the lodes within each stratum, are ordered vertically. By default, these orderings are independent of differentiation aesthetics, so that layers are consistent within and across plots unless otherwise specified. [An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of each of these parameters. For convenience, they can also be set as global options.

## Concepts

Statistical graphics of multivariate categorical---or otherwise discrete---data tend to take one of two forms: ordinations of continuous-valued transformations (e.g. residual plots from multiple regression; ternary plots from tripartition data; biplots from correspondence analysis and factor analysis) and arrangements of length- or area-encoded shapes (e.g. mosaic plots [@Friendly2002]; product plots [@Wickham2011]; stacked area charts).
In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to categorical data, represent cohorts of equivalent cases as ribbons connecting sets represented as boxes [@Kosara2006].

Meanwhile, visualizations of flow processes have long encoded magnitudes, such as volumes and counts, as ribbon widths, and this broad genre have come to be called Sankey diagrams [@Schmidt2008]; see <http://www.sankey-diagrams.com/> for many examples.
A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by curved ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].

Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cases (or cohorts) encoded as ribbons intersecting stacked bar plots at each time point. These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions (with bar plots at each categorical variable axis, arranged as in parallel sets plots).

Several R packages, including **ggplot2** extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:

- **alluvial** for alluvial diagrams in base R [@Bojanowski2016]
- **ggforce** for parallel sets plots in **ggplot2** [@Pedersen2019]
- **ggparallel** and **ggpcp** for parallel sets, hammock, and common angle plots in **ggplot2** [@Hofmann2013;@Ge2019]
- **ggalluvial** for alluvial diagrams in **ggplot2** [@Brunson2019]

While Sankey diagrams, parallel sets plots, and alluvial diagrams have accrued their own characteristic use cases, the terms are often used interchangeably, and there is currently no consensus on what features, if any, are distinctive to each type. Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, others are wholly determined by the data; most give the user limited control through parameter settings.

The diagrams produced by **ggalluvial** are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2005]. The package adopts for them the less common term _alluvial plot_.[^plots]
These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most alluvial diagrams and of Sankey diagrams of a more limited type), features shared by **ggparallel**. The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the `+` syntax of **ggplot2**, a feature shared by **ggpcp**.

[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.

## Applications

**ggalluvial** has seen widespread use in research and scholarship.
While many of these uses would be served equally well by other parallel sets plots or Sankey diagrams, i showcase here three settings to which alluvial plots seem exceptionally well-suited: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.
To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.

**Repeated ordinal measures data:**
Parallel sets plots and Sankey diagrams are commonly used to represent repeated categorical measures data. Most implementations stack each bar plot in order of name or of size, and some allow arbitrary orderings guided by the user, often with vertical space between category for easy visual discrimination. In contrast, ordinal variables are more appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from time to time, vertical separations can obscure whether the cases and cohorts changed in number as well. A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in order across axes (in this case preference ranks rather than time points). In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up. In the next setting, the changing heights of the bar plots are an important part of the data encoded in the plot.
<!--
A related example is provided by @Kissinger2019, who encode the carbon footprints of 64 intra-city commutes as ribbon widths in an alluvial plot on two axes: origin and destination. The strata at each axis are the eight quarters of Tel Aviv, which are follow a municipal numbering scheme and are ordered accordingly at both axes to improve readability.
-->

**Incomplete longitudinal data:**
Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type. While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
<!--
Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point. (See the **babynames** example in the `stat_alluvium()` examples.)
-->

**Signed categorical data:**
@Edwards2019 recently produced a novel alluvial plot to represent changes in ownership category (owner-operated, corporate, etc.) of owners in a halibut fishery. Not only the categorical distribution but the total number of owners changed from year to year as new entrants were not exactly matched by exeters. In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, below the horizontal axis. Such a feature has no analogue in Sankey diagrams or parallel sets plots (which do not commonly employ a cumulative axis) but potentially wide-ranging applications: Bar plots may include both "positive" and "negative" bars to represent definitionally negative groups, such as revenues versus deficits, or to contrast the bars divided along a binary variable such as gender across age groups in a population (so-called "pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such bar plots, even when cases may change sign.

# Acknowledgments

I am grateful to many users for their feedback on every version of this package. Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant from the National Institute of Dental and Craniofacial Research (5T90DE021989-07).

# References

### 2019-12-05 paper/paper.rmd
un-boldface package names in joss paper

@@
 Alluvial plots use variable-width ribbons and stacked bar plots to represent multi-dimensional or repeated-measures data with categorical or ordinal variables [@Rosvall2010;@Bojanowski2016].
-The **ggalluvial** package extends the layered grammar of graphics [@Wickham2016] to generate alluvial plots from tidy data [@Wickham2014].
+The ggalluvial package extends the layered grammar of graphics [@Wickham2016] to generate alluvial plots from tidy data [@Wickham2014].
 The package makes two key contributions to the R ecosystem, which are summarized here and discussed in more detail below.
 
-First, **ggalluvial** anchors the somewhat amorphous notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2005], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
-with the text and point--line geoms of **ggplot2**.
-Moreover, the package gives users a high degree of control over the underlying statistical transformation, whereas other extensions are not so flexible.[^easy] As an extention of **ggplot2**, the package also introduces alluvial plots to a **tidyverse** workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow **tidyr** and **dplyr** conventions.
+First, ggalluvial anchors the somewhat amorphous notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2005], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
+with the text and point--line geoms of ggplot2.
+Moreover, the package gives users a high degree of control over the underlying statistical transformation, whereas other extensions are not so flexible.[^easy] As an extention of ggplot2, the package also introduces alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.
 
-Second, **ggalluvial** adopts a set of conventions meant to cement some of the concepts and terms that are inconsistently applied by developers and users of Sankey diagrams, parallel sets plots, and alluvial diagrams or plots.[^startup]
+Second, ggalluvial adopts a set of conventions meant to cement some of the concepts and terms that are inconsistently applied by developers and users of Sankey diagrams, parallel sets plots, and alluvial diagrams or plots.[^startup]
 Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types.
-Instead, **ggalluvial** anchors its graphical elements to a distinctive, geological nomenclature.
+Instead, ggalluvial anchors its graphical elements to a distinctive, geological nomenclature.
 While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it.
@@
 
-[^easy]: Indeed, the dependency package **easyalluvial** [@Koneswarakantha2019] was built on top of **ggalluvial** to exchange much of this flexibility for expedited data exploration using alluvial plots.
+[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for expedited data exploration using alluvial plots.
 [^startup]: Indeed, the diagrams that came to be known as parallel sets plots [@Kosara2006], consisting of parallel sequences of rectangles connected by parallelogramic ribbons, were anticipated by the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005].
@@
 
-[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of **ggalluvial**, and the reader is encouraged to browse [the documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
-In brief, the package contains stat and geom functions to add the following layers to a **ggplot2** object:
+[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
+In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:
 
@@
 
-The layers of a **ggplot2** graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
+The layers of a ggplot2 graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
 The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation.
@@
 This is because these geoms require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.
-Because these aesthetics are not recognized by **ggplot2**, they produce warnings under some conditions.
-The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of **tidyr**.
+Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
+The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of tidyr.
 
@@
 
-Several R packages, including **ggplot2** extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:
+Several R packages, including ggplot2 extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:
 
-- **alluvial** for alluvial diagrams in base R [@Bojanowski2016]
-- **ggforce** for parallel sets plots in **ggplot2** [@Pedersen2019]
-- **ggparallel** and **ggpcp** for parallel sets, hammock, and common angle plots in **ggplot2** [@Hofmann2013;@Ge2019]
-- **ggalluvial** for alluvial diagrams in **ggplot2** [@Brunson2019]
+- alluvial for alluvial diagrams in base R [@Bojanowski2016]
+- ggforce for parallel sets plots in ggplot2 [@Pedersen2019]
+- ggparallel and ggpcp for parallel sets, hammock, and common angle plots in ggplot2 [@Hofmann2013;@Ge2019]
+- ggalluvial for alluvial diagrams in ggplot2 [@Brunson2019]
 
@@
 
-The diagrams produced by **ggalluvial** are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2005].
+The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2005].
 The package adopts for them the less common term _alluvial plot_.[^plots]
-These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most alluvial diagrams and of Sankey diagrams of a more limited type), features shared by **ggparallel**.
-The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the `+` syntax of **ggplot2**, a feature shared by **ggpcp**.
+These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most alluvial diagrams and of Sankey diagrams of a more limited type), features shared by ggparallel.
+The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the `+` syntax of ggplot2, a feature shared by ggpcp.
 
@@
 
-**ggalluvial** has seen widespread use in research and scholarship.
+ggalluvial has seen widespread use in research and scholarship.
 While many of these uses would be served equally well by other parallel sets plots or Sankey diagrams, i showcase here three settings to which alluvial plots seem exceptionally well-suited: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.
@@
 Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point.
-(See the **babynames** example in the `stat_alluvium()` examples.)
+(See the example in the `stat_alluvium()` examples using the babynames package.)
 -->

### 2019-12-05 paper/paper.rmd
proofread joss paper

@@
 ---
-title: "ggalluvial: A Layered Grammar for Alluvial Plots"
+title: "ggalluvial: Layered Grammar for Alluvial Plots"
 tags:
@@
 bibliography: paper.bib
+output: md_document
 ---
@@
 
-Alluvial plots use variable-width ribbons and stacked bar plots to represent multi-dimensional or repeated-measures data with categorical or ordinal variables [@Rosvall2010;@Bojanowski2016].
-The ggalluvial package extends the layered grammar of graphics [@Wickham2016] to generate alluvial plots from tidy data [@Wickham2014].
-The package makes two key contributions to the R ecosystem, which are summarized here and discussed in more detail below.
+Alluvial diagrams use stacked bar plots and variable-width ribbons to represent multi-dimensional or repeated-measures data comprising categorical or ordinal variables [@Rosvall2010;@Bojanowski2016].
+The ggalluvial package extends the layered grammar of graphics of ggplot2 [@Wickham2016] to generate alluvial diagrams from tidy data [@Wickham2014].
+The package makes two key contributions to the R ecosystem, which i summarize here and discuss in more detail below.
 
-First, ggalluvial anchors the somewhat amorphous notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2005], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
+First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
 with the text and point--line geoms of ggplot2.
-Moreover, the package gives users a high degree of control over the underlying statistical transformation, whereas other extensions are not so flexible.[^easy] As an extention of ggplot2, the package also introduces alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.
+Meanwhile, the package gives users a high degree of control over underlying data transformations, and thereby over the messages conveyed in the plot, whereas other extensions are not so flexible.[^easy] The package also integrates alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.
 
-Second, ggalluvial adopts a set of conventions meant to cement some of the concepts and terms that are inconsistently applied by developers and users of Sankey diagrams, parallel sets plots, and alluvial diagrams or plots.[^startup]
+Second, ggalluvial cements some concepts and terms that are inconsistently applied by developers and users of Sankey, parallel sets, and alluvial diagrams.
+For example, the diagrams that came to be known as parallel sets plots [@Kosara2006] were anticipated in the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005], using wholly different terms for their graphical elements.
 Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types.
-Instead, ggalluvial anchors its graphical elements to a distinctive, geological nomenclature.
+Instead, ggalluvial adopts a distinctive, geological nomenclature for its graphical elements.
 While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it.
-I hope that both the conventions and the terms can serve as useful points of reference as these visualization tools converge toward common standards.
+I hope that these conventions can serve as useful points of reference as these visualization tools converge toward common standards.
 
 [^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for expedited data exploration using alluvial plots.
-[^startup]: Indeed, the diagrams that came to be known as parallel sets plots [@Kosara2006], consisting of parallel sequences of rectangles connected by parallelogramic ribbons, were anticipated by the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005].
 
@@
 
-[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
+[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
 In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:
@@
 The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation.
-Default pairings, other within-package pairings, and pairings of alluvial stats with other geoms are differentiated by fill color:
+Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by `fill` color:
 
-```{r documented pairings with alluvial stats and geoms, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
 library(ggalluvial)
@@
 
-Note that the alluvial geoms are only paired with alluvial stats.
-This is because these geoms require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.
+The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.
 Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
 The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of tidyr.
+Because the alluvial geoms are highly specialized to these stats, no pairings with outside stats are currently supported.
 
@@
 
+While questions and requests by users have prompted improvements throughout ggalluvial's development, the current version is expected to be feature complete, with one exception: alternative curves will be implemented for the `alluvium` and `flow` geoms, controlled by a new parameter.
+The current splines will remain the default, so that the output of existing code will not change.
+This is expected to be the last change before version `1.0`.
+
 ## Concepts
 
-Statistical graphics of multivariate categorical---or otherwise discrete---data tend to take one of two forms: ordinations of continuous-valued transformations (e.g.
+Statistical graphics of multivariate categorical data tend to take one of two forms: ordinations of continuous-valued transformations (e.g.
 residual plots from multiple regression; ternary plots from tripartition data; biplots from correspondence analysis and factor analysis) and arrangements of length- or area-encoded shapes (e.g.
 mosaic plots [@Friendly2002]; product plots [@Wickham2011]; stacked area charts).
-In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to categorical data, represent cohorts of equivalent cases as ribbons connecting sets represented as boxes [@Kosara2006].
+In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990], represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
 
 Meanwhile, visualizations of flow processes have long encoded magnitudes, such as volumes and counts, as ribbon widths, and this broad genre have come to be called Sankey diagrams [@Schmidt2008]; see <http://www.sankey-diagrams.com/> for many examples.
-A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by curved ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].
+A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].
 
-Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cases (or cohorts) encoded as ribbons intersecting stacked bar plots at each time point.
-These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions (with bar plots at each categorical variable axis, arranged as in parallel sets plots).
+Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cohorts of equivalent cases encoded as ribbons intersecting stacked bar plots at each time point.
+These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions.
 
@@
 While Sankey diagrams, parallel sets plots, and alluvial diagrams have accrued their own characteristic use cases, the terms are often used interchangeably, and there is currently no consensus on what features, if any, are distinctive to each type.
-Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, others are wholly determined by the data; most give the user limited control through parameter settings.
+Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, while others are wholly determined by the data; most give the user limited control through parameter settings.
 
-The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2005].
+The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
 The package adopts for them the less common term _alluvial plot_.[^plots]
-These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most alluvial diagrams and of Sankey diagrams of a more limited type), features shared by ggparallel.
-The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the `+` syntax of ggplot2, a feature shared by ggpcp.
+These plots are distinct from Sankey diagrams (they exhibit a much narrower range of behavior, e.g.
+no cycles, and do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most diagrams of each type), features shared only by ggparallel.
+The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.
 
@@
 Parallel sets plots and Sankey diagrams are commonly used to represent repeated categorical measures data.
-Most implementations stack each bar plot in order of name or of size, and some allow arbitrary orderings guided by the user, often with vertical space between category for easy visual discrimination.
+Most implementations stack each bar plot in order of name or of size, though some follow hierarchies provided by the user, often with vertical space between category for easy visual discrimination.
 In contrast, ordinal variables are more appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from time to time, vertical separations can obscure whether the cases and cohorts changed in number as well.
@@
 Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in order across axes (in this case preference ranks rather than time points).
-In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.
-In the next setting, the changing heights of the bar plots are an important part of the data encoded in the plot.
-<!--
-A related example is provided by @Kissinger2019, who encode the carbon footprints of 64 intra-city commutes as ribbon widths in an alluvial plot on two axes: origin and destination.
-The strata at each axis are the eight quarters of Tel Aviv, which are follow a municipal numbering scheme and are ordered accordingly at both axes to improve readability.
--->
 
@@
 @Edwards2019 recently produced a novel alluvial plot to represent changes in ownership category (owner-operated, corporate, etc.) of owners in a halibut fishery.
-Not only the categorical distribution but the total number of owners changed from year to year as new entrants were not exactly matched by exeters.
+Not only the categorical distribution but the total number of owners changed from year to year as exeters were not exactly matched by new entrants.
 In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, below the horizontal axis.
@@
 Alluvial plots provide a way to track cases and cohorts across such bar plots, even when cases may change sign.
+Future use cases may demonstrate the practical utility of this functionality.
 

### 2019-12-05 paper.rmd
move paper to root

--- 2019-12-05 paper/paper.rmd
@@
 
-<!--
-To generate `paper.md`, execute this from the console:
-```r
-knitr::knit("paper/paper.rmd", "paper/paper.md")
-```
--->
-
 ```{r, echo=FALSE}
@@
   comment = "#>",
-  fig.path = "paper/figures/"
+  fig.path = "paper/"
 )

### 2019-12-05 paper.rmd
restore code to manually knit rmd to md
try to rm figure caption

@@
 bibliography: paper.bib
-output: md_document
 ---
+
+<!--
+To generate `paper.md`, execute this from the console:
+```r
+knitr::knit("paper.rmd", "paper.md")
+```
+-->
 
@@
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap=NA}
 library(ggalluvial)
@@
 @Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
-<!--
-Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point.
-(See the example in the `stat_alluvium()` examples using the babynames package.)
--->
 

### 2019-12-05 paper.rmd
try to rm figure caption 2

@@
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap=NA}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap=""}
 library(ggalluvial)

### 2019-12-05 paper.rmd
shorten joss paper

@@
 The ggalluvial package extends the layered grammar of graphics of ggplot2 [@Wickham2016] to generate alluvial diagrams from tidy data [@Wickham2014].
-The package makes two key contributions to the R ecosystem, which i summarize here and discuss in more detail below.
 
-First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
-with the text and point--line geoms of ggplot2.
-Meanwhile, the package gives users a high degree of control over underlying data transformations, and thereby over the messages conveyed in the plot, whereas other extensions are not so flexible.[^easy] The package also integrates alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.
-
-Second, ggalluvial cements some concepts and terms that are inconsistently applied by developers and users of Sankey, parallel sets, and alluvial diagrams.
-For example, the diagrams that came to be known as parallel sets plots [@Kosara2006] were anticipated in the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005], using wholly different terms for their graphical elements.
-Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types.
-Instead, ggalluvial adopts a distinctive, geological nomenclature for its graphical elements.
-While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it.
-I hope that these conventions can serve as useful points of reference as these visualization tools converge toward common standards.
-
-[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for expedited data exploration using alluvial plots.
+The package makes two key contributions to the R ecosystem.
+First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which lends the plots more precise meaning and opens up many combinatorial possibilities.
+Second, ggalluvial adopts a distinctive geological nomenclature to distinguish "alluvial plots" and their graphical elements from Sankey diagrams and parallel sets plots, which i hope prove useful as these visualization tools converge toward common standards.
 
@@
 
-[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
+[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples.
 In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:
 
-* _strata_, or stacked bar plots, located at each of multiple horizontal positions (_axes_ or _dimensions_)
-* _alluvia_, ribbons through strata at different axes that encode individual cases or cohorts measured along each axis
+* _strata_, or stacked bar plots, located in parallel along a (plotting) axis of (variable) _axes_ or _dimensions_
+* _alluvia_, ribbons through strata that connect the categories of individual cases or cohorts at different axes
 * _lodes_, subdivisions of strata by their intersections with alluvia
-* _flows_, segments of alluvia between adjacent axes (excluding lodes)
+* _flows_, segments of alluvia between strata
 
-The layers of a ggplot2 graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
-The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation.
-Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by `fill` color:
+Plot layers are formed by pairing stats (statistical transformations) with geoms (mappings to graphical elements and properties); while every stat and geom has a conventional default, alternative grammatical pairings provide combinatorial richness to plotting possibilities.
+The following alluvial plot depicts several meaningful stat--geom combinations from the documentation.
+Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by color:
 
@@
 
-The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.
-Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
-The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of tidyr.
-Because the alluvial geoms are highly specialized to these stats, no pairings with outside stats are currently supported.
+Alluvial layers can interpret two forms of tidy data, long (one row per lode) and wide (one row per alluvium), related by the pivot operations of tidyr [@Wickham2019].
+The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.[^warnings] Because the alluvial geoms are specialized to these stats, no pairings with outside stats are currently supported.
+
+[^warnings]: Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
 
@@
 By default, these orderings are independent of differentiation aesthetics, so that layers are consistent within and across plots unless otherwise specified.
-[An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of each of these parameters.
-For convenience, they can also be set as global options.
+[An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of these parameters.
+They can also be set as global options.
 
-While questions and requests by users have prompted improvements throughout ggalluvial's development, the current version is expected to be feature complete, with one exception: alternative curves will be implemented for the `alluvium` and `flow` geoms, controlled by a new parameter.
-The current splines will remain the default, so that the output of existing code will not change.
-This is expected to be the last change before version `1.0`.
+While questions and requests by users have prompted many improvements, the current version is feature complete with one exception: Alternative flow elements will be implemented and controlled by a new parameter.
+The current splines will remain the default.
 
@@
 
-Statistical graphics of multivariate categorical data tend to take one of two forms: ordinations of continuous-valued transformations (e.g.
-residual plots from multiple regression; ternary plots from tripartition data; biplots from correspondence analysis and factor analysis) and arrangements of length- or area-encoded shapes (e.g.
-mosaic plots [@Friendly2002]; product plots [@Wickham2011]; stacked area charts).
-In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990], represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
-
-Meanwhile, visualizations of flow processes have long encoded magnitudes, such as volumes and counts, as ribbon widths, and this broad genre have come to be called Sankey diagrams [@Schmidt2008]; see <http://www.sankey-diagrams.com/> for many examples.
-A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].
-
-Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cohorts of equivalent cases encoded as ribbons intersecting stacked bar plots at each time point.
-These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions.
-
-Several R packages, including ggplot2 extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:
-
-- alluvial for alluvial diagrams in base R [@Bojanowski2016]
-- ggforce for parallel sets plots in ggplot2 [@Pedersen2019]
-- ggparallel and ggpcp for parallel sets, hammock, and common angle plots in ggplot2 [@Hofmann2013;@Ge2019]
-- ggalluvial for alluvial diagrams in ggplot2 [@Brunson2019]
-
-While Sankey diagrams, parallel sets plots, and alluvial diagrams have accrued their own characteristic use cases, the terms are often used interchangeably, and there is currently no consensus on what features, if any, are distinctive to each type.
-Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, while others are wholly determined by the data; most give the user limited control through parameter settings.
+Visualizations of flow processes have long encoded magnitudes as ribbon widths, constituting a type called Sankey diagrams [@Schmidt2008].
+A widely-used subtype for longitudinal categorical data represent categories as nodes threaded by edges that represent the trajectories and magnitudes of cases [@Riehmann2005].
+Their design anticipated parallel sets plots, which were adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to visualize multivariate categorical data, and which represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
+These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
+Several R packages have been developed to generate these diagrams, in particular alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].
 
 The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
-The package adopts for them the less common term _alluvial plot_.[^plots]
-These plots are distinct from Sankey diagrams (they exhibit a much narrower range of behavior, e.g.
-no cycles, and do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most diagrams of each type), features shared only by ggparallel.
-The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.
+The package adopts for them the term _alluvial plot_.[^plots]
+These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them, features shared only by ggparallel.
+The package includes several levers of control over the statistical transformations, and thereby over the messages conveyed by the plot.[^easy]
+The plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.
 
 [^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
+[^easy]: Other ggplot2 extensions are not so flexible; in particular, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
+
+Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
+Moreover, their graphical elements go by a variety of names, often interchangeably.
+While adopting new terms risks adding to this confusion, these alluvial plots are, i think, distinctive enough to warrant it.
 
@@
 
-ggalluvial has seen widespread use in research and scholarship.
-While many of these uses would be served equally well by other parallel sets plots or Sankey diagrams, i showcase here three settings to which alluvial plots seem exceptionally well-suited: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.
-To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.
+While most uses might be served equally well by other parallel sets plots or Sankey diagrams, ggalluvial seems exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]
+
+[^opinion]: To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.
 
 **Repeated ordinal measures data:**
-Parallel sets plots and Sankey diagrams are commonly used to represent repeated categorical measures data.
-Most implementations stack each bar plot in order of name or of size, though some follow hierarchies provided by the user, often with vertical space between category for easy visual discrimination.
-In contrast, ordinal variables are more appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from time to time, vertical separations can obscure whether the cases and cohorts changed in number as well.
+Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination.
+Ordinal variables are most appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
 A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
-Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in order across axes (in this case preference ranks rather than time points).
+Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order.
+In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.
 
 **Incomplete longitudinal data:**
-Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
-@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times.
-Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type.
-While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
-@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
+Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
+@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected patients who had their infections ribotyped at multiple times.
+Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification.
+While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the dimensional axis.
+@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.
 
 **Signed categorical data:**
-@Edwards2019 recently produced a novel alluvial plot to represent changes in ownership category (owner-operated, corporate, etc.) of owners in a halibut fishery.
-Not only the categorical distribution but the total number of owners changed from year to year as exeters were not exactly matched by new entrants.
-In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, below the horizontal axis.
-Such a feature has no analogue in Sankey diagrams or parallel sets plots (which do not commonly employ a cumulative axis) but potentially wide-ranging applications: Bar plots may include both "positive" and "negative" bars to represent definitionally negative groups, such as revenues versus deficits, or to contrast the bars divided along a binary variable such as gender across age groups in a population (so-called "pyramid plots").
-Alluvial plots provide a way to track cases and cohorts across such bar plots, even when cases may change sign.
+@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery.
+The total number of owners changed from year to year as exeters were not exactly matched by new entrants.
+In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, reflected across the dimensional axis.
+Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
+Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign.
 Future use cases may demonstrate the practical utility of this functionality.

### 2019-12-19 paper.rmd
resize image for joss paper pdf

@@
 ---
-title: "ggalluvial: Layered Grammar for Alluvial Plots"
+title: "Layered Grammar for Alluvial Plots"
 tags:
@@
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap=""}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap=""}
 library(ggalluvial)

### 2020-01-01 paper.rmd
typos in paper

@@
 Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination.
-Ordinal variables are most appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
+Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
 A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
@@
 @Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery.
-The total number of owners changed from year to year as exeters were not exactly matched by new entrants.
-In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, reflected across the dimensional axis.
+The total number of owners changed from year to year as exiters were not exactly matched by new entrants.
+In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.
 Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
@@
 I am grateful to many users for their feedback on every version of this package.
-Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant from the National Institute of Dental and Craniofacial Research (5T90DE021989-07).
+Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
 

### 2020-01-06 paper.rmd
improvements + alternatives to splines

@@
 
-While questions and requests by users have prompted many improvements, the current version is feature complete with one exception: Alternative flow elements will be implemented and controlled by a new parameter.
+While questions and requests by users have prompted many improvements, the current version is feature complete with one exception: improvements and alternatives to the flow elements connecting lodes.
 The current splines will remain the default.

### 2020-01-17 README.rmd
replace deprecated param from readme

@@
   geom_alluvium(aes(fill = Survived)) +
-  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
+  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
   theme_minimal() +

### 2020-01-17 paper.rmd
revise & extend comparisons with other packages

@@
 
-[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples.
+[The primary vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples.
 In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:
@@
 
-While questions and requests by users have prompted many improvements, the current version is feature complete with one exception: improvements and alternatives to the flow elements connecting lodes.
-The current splines will remain the default.
-
 ## Concepts
@@
 
-The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
-The package adopts for them the term _alluvial plot_.[^plots]
-These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them, features shared only by ggparallel.
-The package includes several levers of control over the statistical transformations, and thereby over the messages conveyed by the plot.[^easy]
-The plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.
+Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
+Moreover, their graphical elements go by a variety of names, often interchangeably.
+In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a formal demarcation among these diagram types.
 
-[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
+_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
+By design, graphics produced by ggplot2 extensions (and by the alluvial package) are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
+Sankey diagrams are much more flexible: The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008], and software implementations may be interactive and enable manual adjustments of graphical elements [@Riehmann2005].
+Moreover, the plots generated by these extensions are suitable for a wide variety of data types, whereas Sankey diagrams are still overwhelmingly used to represent flow.
+Arguably, the extensions are better understood as belonging to a different type of diagram.
+
+[^exceptions]: There are exceptions.
+For example, the jitter geom in ggplot2 introduces randomness to symbol positions, and the repel geoms of ggrepel [@Slowikowski2019] use heuristic algorithms to position text.
+
+Parallel sets plots might be viewed as a special subtype of Sankey diagram.
+In turn, the plots produced by the aforecited packages are all parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
+They vary in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
+The geoms of ggalluvial offer only a one-dimensional continuum between straight lines and x-splines for the ribbon shapes, though alternatives are planned.
+The stats vertically arrange the elements (boxes and ribbons) at each axis.
+ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
+Notably, in ggalluvial the plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared by ggforce and ggpcp but not by ggparallel.
+
 [^easy]: Other ggplot2 extensions are not so flexible; in particular, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
 
-Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
-Moreover, their graphical elements go by a variety of names, often interchangeably.
-While adopting new terms risks adding to this confusion, these alluvial plots are, i think, distinctive enough to warrant it.
+The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] which are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them.
+The first feature is shared by the other packages but are not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes.
+While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots.
+(ggparallel and ggalluvial alone _only_ produce alluvial plots.)
+This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
+It is for this reason that i believe the new geological terminology is warranted.
+
+[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than i use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
+[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
 

### 2020-01-17 paper.rmd
edit 'easy' footnote following edits to surrounding text

@@
 
-[^easy]: Other ggplot2 extensions are not so flexible; in particular, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
+[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
 

### 2020-01-18 paper.rmd
revise concepts section of joss paper

@@
 These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
-Several R packages have been developed to generate these diagrams, in particular alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].
+Several R packages have been developed to generate diagrams of these types, including riverplot [@Weiner2017], networkD3 [@Allaire2017], sankey [@Csardi2017], alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].
 
@@
 Moreover, their graphical elements go by a variety of names, often interchangeably.
-In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a formal demarcation among these diagram types.
+In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a careful demarcation among these diagram types.
 
 _Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
-By design, graphics produced by ggplot2 extensions (and by the alluvial package) are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
-Sankey diagrams are much more flexible: The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008], and software implementations may be interactive and enable manual adjustments of graphical elements [@Riehmann2005].
-Moreover, the plots generated by these extensions are suitable for a wide variety of data types, whereas Sankey diagrams are still overwhelmingly used to represent flow.
-Arguably, the extensions are better understood as belonging to a different type of diagram.
+By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
+Sankey diagrams are much more flexible.
+The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
+Software implementations may use heuristic algorithms to position graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
+Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
+Arguably, the extensions are better understood as producing a different type of diagram.
 
@@
 
-Parallel sets plots might be viewed as a special subtype of Sankey diagram.
-In turn, the plots produced by the aforecited packages are all parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
-They vary in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
-The geoms of ggalluvial offer only a one-dimensional continuum between straight lines and x-splines for the ribbon shapes, though alternatives are planned.
+Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and e very ribbon encounters a box at every axis.
+These graphical constraints correspond to combinatorial constraints on the data, which amount to a id--key--value structure in which every id--key pair takes exactly one value (possibly 0 or `NA`).
+In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
+
+The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
+The exceptional geoms of ggparallel offer common-angle as well as linear ribbons.
+Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
 The stats vertically arrange the elements (boxes and ribbons) at each axis.
+These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
 ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
-Notably, in ggalluvial the plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared by ggforce and ggpcp but not by ggparallel.
 
@@
 
-The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] which are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them.
-The first feature is shared by the other packages but are not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes.
+The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology itemized above.
+These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them.
+The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes.
 While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots.
@@
 This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
-It is for this reason that i believe the new geological terminology is warranted.
+It is for this reason that i believe the typology and terminology are warranted.
 
@@
 
-While most uses might be served equally well by other parallel sets plots or Sankey diagrams, ggalluvial seems exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]
+While most uses might be served equally well by other parallel sets plots or Sankey diagrams, alluvial plots seem exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]
 

### 2020-01-20 paper.rmd
typo

@@
 
-Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and e very ribbon encounters a box at every axis.
+Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis.
 These graphical constraints correspond to combinatorial constraints on the data, which amount to a id--key--value structure in which every id--key pair takes exactly one value (possibly 0 or `NA`).

### 2020-01-20 paper.rmd
small paper revisions

@@
 
-Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis.
-These graphical constraints correspond to combinatorial constraints on the data, which amount to a id--key--value structure in which every id--key pair takes exactly one value (possibly 0 or `NA`).
+Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis.
+These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing).
 In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
@@
 The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
-The exceptional geoms of ggparallel offer common-angle as well as linear ribbons.
+The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons.
 Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
@@
 
-The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology itemized above.
-These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them.
+The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology introduced above.
+These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
 The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes.
@@
 The total number of owners changed from year to year as exiters were not exactly matched by new entrants.
-In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.
-Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
+In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
 Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign.
-Future use cases may demonstrate the practical utility of this functionality.
+Future use cases may demonstrate additional practical uses for this functionality.
+
+[^negative]: The authors should be credited with this innovation, which i only implemented in ggalluvial after learning about their workaround to create it using a previous version.
 

### 2020-02-05 paper.rmd
paper - declutter refs + add curves footnote

@@
 The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons.
-Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
+Those of alluvial, ggforce, ggalluvial, and ggpcp offer one-parameter families that interpolate between straight and x-spline ribbons.[^curves]
 The stats vertically arrange the elements (boxes and ribbons) at each axis.
@@
 
+[^curves]: Several alternative curves, based on @Shaffer2019, are in development.
 [^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
@@
 This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
-It is for this reason that i believe the typology and terminology are warranted.
+It is for this reason that i believe the new typology and terminology are warranted.
 

### 2020-02-20 paper.rmd

@@
 ---
-title: "Layered Grammar for Alluvial Plots"
+title: "ggalluvial: Layered Grammar for Alluvial Plots"
 tags:
@@
 First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which lends the plots more precise meaning and opens up many combinatorial possibilities.
-Second, ggalluvial adopts a distinctive geological nomenclature to distinguish "alluvial plots" and their graphical elements from Sankey diagrams and parallel sets plots, which i hope prove useful as these visualization tools converge toward common standards.
+Second, ggalluvial adopts a distinctive geological nomenclature to distinguish "alluvial plots" and their graphical elements from Sankey diagrams and parallel sets plots, which I hope prove useful as these visualization tools converge toward common standards.
 
@@
 This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
-It is for this reason that i believe the new typology and terminology are warranted.
+It is for this reason that I believe the new typology and terminology are warranted.
 
-[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than i use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
+[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than I use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
 [^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
@@
 
-[^negative]: The authors should be credited with this innovation, which i only implemented in ggalluvial after learning about their workaround to create it using a previous version.
+[^negative]: The authors should be credited with this innovation, which I only implemented in ggalluvial after learning about their workaround to create it using a previous version.
 
@@
 I am grateful to many users for their feedback on every version of this package.
-Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
+Development benefitted from the use of resources and support of colleagues at UConn Health, and I have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
 

### 2020-03-04 README.rmd
link to contributing guidelines & code of conduct from readme

@@
 
-[Issues](https://github.com/corybrunson/ggalluvial/issues) and [pull requests](https://github.com/corybrunson/ggalluvial/pulls) are more than welcome!
+Contributions in any form are more than welcome!
 Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.
+See the [CONTRIBUTING](CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](CODE_OF_CONDUCT.md).
 

### 2020-03-16 paper.rmd
more illustrative example in paper

@@
 
-Plot layers are formed by pairing stats (statistical transformations) with geoms (mappings to graphical elements and properties); while every stat and geom has a conventional default, alternative grammatical pairings provide combinatorial richness to plotting possibilities.
-The following alluvial plot depicts several meaningful stat--geom combinations from the documentation.
-Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by color:
+These layers, and other plot elements, are illustrated in the alluvial plot below, which visualizes changes in several students' curricula (based on their declared majors) across several academic terms:
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap=""}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=8, fig.cap=""}
 library(ggalluvial)
-combos <- data.frame(
-  stat = c(
-    rep("stratum", 2),
-    rep("alluvium", 4),
-    "flow",
-    # text geom
-    c("stratum", "alluvium", "flow"),
-    # errorbar and pointrange geoms
-    "stratum", "alluvium", "flow"
-  ),
-  geom = c(
-    c("stratum", "lode"),
-    c("stratum", "alluvium", "lode", "flow"),
-    "flow",
-    rep("text", 3),
-    "errorbar", "pointrange", "pointrange"
-  ),
-  pairing = c(
-    c("default", "alluvial"),
-    c("alluvial", "default", "default", "alluvial"),
-    "default",
-    rep("external", 6)
-  )
-)
-levs <- c("stratum", "alluvium", "lode", "flow",
-          "text", "errorbar", "pointrange")
-combos <- transform(
-  combos,
-  stat = factor(stat, levs),
-  geom = factor(geom, levs),
-  pairing = factor(pairing, c("default", "alluvial", "external"))
-)
-ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
-  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
-  geom_alluvium(aes(fill = pairing)) +
-  scale_fill_brewer(type = "qual") +
-  geom_stratum() +
-  geom_text(stat = "stratum", infer.label = TRUE) +
-  theme_minimal()
+data(majors)
+majors$curriculum <- as.factor(majors$curriculum)
+ggplot(majors,
+       aes(x = semester, stratum = curriculum, alluvium = student,
+           fill = curriculum)) +
+  scale_fill_brewer(type = "qual", palette = "Set2") +
+  scale_y_continuous(breaks = seq(-2, 10, 2)) +
+  geom_flow(stat = "alluvium", lode.guidance = "frontback",
+            color = "darkgray", negate.strata = NA) +
+  geom_stratum(negate.strata = NA) +
+  theme(legend.position = "right") +
+  labs(x = NULL) +
+  ggtitle("Changes in curricula in a student cohort")
 ```
 
-Alluvial layers can interpret two forms of tidy data, long (one row per lode) and wide (one row per alluvium), related by the pivot operations of tidyr [@Wickham2019].
+Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc.
+The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but switched to Painting in the 11th term, where the ribbon turns pink.
+The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves.
+When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.
+
+Plot layers are formed by pairing stats (statistical transformations) with geoms (mappings to graphical elements and properties); while every stat and geom has a conventional default, alternative grammatical pairings provide combinatorial richness to plotting possibilities.
+In the above example, the alluvium geom was paired with the flow stat, so that the flows of each alluvium could change color across the axes.
+Other meaningful stat--geom combinations can be found in the documentation, including pairings of the three alluvial stats (stratum, alluvium, and flow) with the text, errorbar, and pointrange geoms.
+
+Alluvial layers can interpret tidy data in either of two formats: long (one row per lode) and wide (one row per alluvium).
+These are related by the pivot operations of tidyr [@Wickham2019] and can be toggled between using the custom functions `to_lodes_form()` and `to_alluvia_form()`.
 The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.[^warnings] Because the alluvial geoms are specialized to these stats, no pairings with outside stats are currently supported.

### 2020-03-16 paper.rmd
add figure caption w/ cross-ref

@@
 
-These layers, and other plot elements, are illustrated in the alluvial plot below, which visualizes changes in several students' curricula (based on their declared majors) across several academic terms:
+Figure \ref{fig:example} illustrates these and other plot elements by visualizing changes in several students' curricula (based on their declared majors) across several academic terms.
+Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc.
+The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but switched to Painting in the 11th term, where the ribbon turns pink.
+The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves.
+When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=8, fig.cap=""}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap="\\label{fig:example}Alluvial plot of changes in curricula in a student cohort"}
 library(ggalluvial)
@@
   theme(legend.position = "right") +
-  labs(x = NULL) +
-  ggtitle("Changes in curricula in a student cohort")
+  labs(x = NULL)
 ```
-
-Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc.
-The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but switched to Painting in the 11th term, where the ribbon turns pink.
-The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves.
-When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.
 

### 2020-03-16 paper.rmd
proofread joss paper

@@
 Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc.
-The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but switched to Painting in the 11th term, where the ribbon turns pink.
+The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but had switched to Painting by the 11th term, where the ribbon turns pink.
 The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves.
@@
 Moreover, their graphical elements go by a variety of names, often interchangeably.
-In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a careful demarcation among these diagram types.
+In order to more clearly describe the features of ggalluvial in relation to similar packages, I have found it useful to adopt a careful demarcation among these diagram types.
 
@@
 The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
-Software implementations may use heuristic algorithms to position graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
+Software implementations may use heuristic algorithms to position their graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
 Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
-Arguably, the extensions are better understood as producing a different type of diagram.
+Arguably, these extensions are better understood as producing a different type of diagram.
 
@@
 Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis.
-These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing).
+These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing, and optionally weighted).
 In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
@@
 These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
-ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
+ggalluvial provides more levers of control over the statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
 
@@
 
-The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology introduced above.
-These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
-The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes.
+The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces, with the geological terminology introduced above.[^plots]
+These alluvial plots are distinguished by two features: a prescribed order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
+In combination, these features confer greater meaning on the second plotting dimension.
+
+The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could, for example, arrange boxes corresponding to repeated categorical decompositions differently at different axes.
 While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots.
 (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
-This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
+These features are particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
 It is for this reason that I believe the new typology and terminology are warranted.
@@
 
-**Repeated ordinal measures data:**
-Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination.
-Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
-A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
+**Repeated ordinal measures data.**
+Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size (though some follow user-provided hierarchies), and most insert gaps between categories for easy visual discrimination.
+Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories (hence of gaps) changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
+A use case by @Schlotter2019, to represent patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
 Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order.
@@
 
-**Incomplete longitudinal data:**
+**Incomplete longitudinal data.**
 Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@@
 Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification.
-While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the dimensional axis.
+While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the main dimension.
 @Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.
 
-**Signed categorical data:**
+**Signed categorical data.**
 @Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery.
 The total number of owners changed from year to year as exiters were not exactly matched by new entrants.
-In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
-Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign.
-Future use cases may demonstrate additional practical uses for this functionality.
+In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
+Alluvial plots provide a way to track cases and cohorts across such graphics, even when cases change sign.
+Future applications may demonstrate additional uses for this functionality.
 

### 2020-03-20 R/alluvial-data.r
deprecate infer-label param + defunct label-strata param

@@ -271,13 +271,6 @@ to_alluvia_form <- function(data,
   res
 }
 
-# distilling functions
-first <- dplyr::first
-last <- dplyr::last
-most <- function(x) {
-  x[which(factor(x) == names(which.max(table(factor(x)))))[1]]
-}
-
 # require different character strings to represent strata at different axes
 discern_data <- function(data, axes, sep = ".") {
   # strata at each axis in order

### 2020-03-20 R/stat-flow.r

@@ -144,6 +144,8 @@ StatFlow <- ggproto(
     
     # introduce label
     if (infer.label) {
+      deprecate_parameter("infer.label",
+                          msg = "Use `aes(label = after_stat(alluvium))`.")
       if (is.null(data$label)) {
         data$label <- data$alluvium
       } else {

### 2020-03-20 R/utils.r

@@ -10,3 +10,10 @@
   "vjust", "hjust", "angle",
   "family", "fontface", "lineheight"
 )
+
+# distilling functions
+first <- dplyr::first
+last <- dplyr::last
+most <- function(x) {
+  x[which(factor(x) == names(which.max(table(factor(x)))))[1]]
+}

### 2020-03-20 inst/examples/ex-alluvial-data.r

@@ -48,11 +48,11 @@ gg <- ggplot(majors_alluvia,
 gg +
   geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = TRUE) +
   geom_stratum(width = 2/5, discern = TRUE) +
-  geom_text(stat = "stratum", discern = TRUE, infer.label = TRUE)
+  geom_text(stat = "stratum", discern = TRUE, aes(label = after_stat(stratum)))
 gg +
   geom_alluvium(aes(fill = as.factor(student)), width = 2/5, discern = FALSE) +
   geom_stratum(width = 2/5, discern = FALSE) +
-  geom_text(stat = "stratum", discern = FALSE, infer.label = TRUE)
+  geom_text(stat = "stratum", discern = FALSE, aes(label = after_stat(stratum)))
 # warning when inappropriate
 ggplot(majors[majors$semester %in% paste0("CURR", c(1, 7, 13)), ],
        aes(x = semester, stratum = curriculum, alluvium = student,

### 2020-03-20 inst/examples/ex-geom-flow.r

@@ -4,7 +4,8 @@ ggplot(as.data.frame(Titanic),
            axis1 = Class, axis2 = Sex, axis3 = Age)) +
   geom_flow() +
   scale_x_discrete(limits = c("Class", "Sex", "Age")) +
-  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
+  geom_stratum() +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
   ggtitle("Alluvial plot of Titanic passenger demographic data")
 
 # use of facets
@@ -13,7 +14,7 @@ ggplot(as.data.frame(Titanic),
            axis1 = Class, axis2 = Sex)) +
   geom_flow(aes(fill = Age), width = .4) +
   geom_stratum(width = .4) +
-  geom_text(stat = "stratum", infer.label = TRUE, size = 3) +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
   scale_x_discrete(limits = c("Class", "Sex")) +
   facet_wrap(~ Survived, scales = "fixed")
 

### 2020-03-20 inst/examples/ex-stat-alluvium.r

@@ -6,14 +6,15 @@ ggplot(as.data.frame(Titanic),
   stat_stratum(geom = "errorbar") +
   geom_line(stat = "alluvium") +
   stat_alluvium(geom = "pointrange") +
-  geom_text(stat = "stratum", infer.label = TRUE) +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 
 # lode ordering examples
 gg <- ggplot(as.data.frame(Titanic),
              aes(y = Freq,
                  axis1 = Class, axis2 = Sex, axis3 = Age)) +
-  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
+  geom_stratum() +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"))
 # use of lode controls
 gg + geom_flow(aes(fill = Survived, alpha = Sex), stat = "alluvium",
@@ -110,6 +111,7 @@ ggplot(subset(titanic, Class != "Crew"),
        aes(axis1 = Class, axis2 = Sex, axis3 = Age, y = Lives)) +
   geom_alluvium(aes(alpha = Survived, fill = Class), absolute = FALSE) +
   geom_stratum(absolute = FALSE) +
-  geom_text(stat = "stratum", infer.label = TRUE, absolute = FALSE) +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
+            absolute = FALSE) +
   scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
   scale_alpha_discrete(range = c(.25, .75), guide = FALSE)

### 2020-03-20 inst/examples/ex-stat-stratum.r

@@ -17,7 +17,7 @@ ggplot(vaccinations,
 ggplot(as.data.frame(Titanic),
        aes(y = Freq,
            axis1 = Class, axis2 = Sex, axis3 = Age, axis4 = Survived)) +
-  geom_text(stat = "stratum", infer.label = TRUE) +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
   stat_stratum(geom = "errorbar") +
   scale_x_discrete(limits = c("Class", "Sex", "Age", "Survived"))
 
@@ -50,5 +50,5 @@ ggplot(admissions,
        aes(y = Count, axis1 = Dept, axis2 = Gender)) +
   geom_alluvium(aes(fill = Dept), width = 1/12) +
   geom_stratum(width = 1/12, fill = "black", color = "grey") +
-  geom_label(stat = "stratum", infer.label = TRUE, min.y = 200) +
+  geom_label(stat = "stratum", aes(label = after_stat(stratum)), min.y = 200) +
   scale_x_discrete(limits = c("Department", "Gender"), expand = c(.05, .05))

### 2020-03-22 paper.rmd
annotate figure in joss paper

@@
 majors$curriculum <- as.factor(majors$curriculum)
+arr <- arrow(angle = 15, type = "closed", length = unit(.03, "npc"))
 ggplot(majors,
@@
   theme(legend.position = "right") +
-  labs(x = NULL)
+  labs(x = NULL) +
+  # annotate axes
+  annotate("segment", x = 2.33, y = -1.33, xend = 2, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3, y = -1.33, xend = 3, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3.67, y = -1.33, xend = 4, yend = -.33, arrow = arr) +
+  annotate("text", x = 3, y = -1.67, label = "axes / dimensions", size = 4) +
+  # annotate strata
+  annotate("segment", x = 1.5, xend = 1, y = 3.5, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4.5, yend = 6, arrow = arr) +
+  annotate("text", x = 1.67, y = 4, label = "strata", angle = 90, size = 4) +
+  # annotate alluvia
+  annotate("segment", x = 4.33, y = 6.5, xend = 1, yend = 8.5, arrow = arr) +
+  annotate("segment", x = 5, y = 6.5, xend = 4.5, yend = 9, arrow = arr) +
+  annotate("segment", x = 5.67, y = 6.5, xend = 7.5, yend = 7, arrow = arr) +
+  annotate("text", x = 5, y = 6.17, label = "one alluvium", size = 4) +
+  # annotate flows
+  annotate("segment", x = 5.33, y = 2.5, xend = 4.25, yend = 0.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3, xend = 4.25, yend = 2.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3.5, xend = 3.75, yend = 4.5, arrow = arr) +
+  annotate("text", x = 5.5, y = 3, label = "flows", angle = 90, size = 4) +
+  # annotate lodes
+  annotate("segment", x = 7.5, y = 3, xend = 8, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 3.5, xend = 8, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 4, xend = 8, yend = 5.5, arrow = arr) +
+  annotate("text", x = 7.33, y = 3.5, label = "lodes", angle = -90, size = 4)
 ```

### 2020-03-22 paper.rmd
clarify labels in joss paper figure

--- 2020-03-16 paper.rmd
@@
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap="\\label{fig:example}Alluvial plot of changes in curricula in a student cohort"}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap="\\label{fig:example}Alluvial plot of changes in curricula by a cohort of art students"}
 library(ggalluvial)
@@
 majors$curriculum <- as.factor(majors$curriculum)
+levels(majors$semester) <- gsub("CURR", "Term ", levels(majors$semester))
+arr <- arrow(angle = 15, type = "closed", length = unit(.03, "npc"))
 ggplot(majors,
@@
            fill = curriculum)) +
-  scale_fill_brewer(type = "qual", palette = "Set2") +
+  scale_fill_brewer(type = "qual", palette = "Set2", name = "Curriculum") +
   scale_y_continuous(breaks = seq(-2, 10, 2)) +
@@
   theme(legend.position = "right") +
-  labs(x = NULL)
+  labs(x = NULL, y = "Number of students") +
+  # annotate axes
+  annotate("segment", x = 2.33, y = -1.33, xend = 2, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3, y = -1.33, xend = 3, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3.67, y = -1.33, xend = 4, yend = -.33, arrow = arr) +
+  annotate("text", x = 3, y = -1.67, label = "axes / dimensions", size = 4) +
+  # annotate strata
+  annotate("segment", x = 1.5, xend = 1, y = 3.5, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4.5, yend = 6, arrow = arr) +
+  annotate("text", x = 1.67, y = 4, label = "strata", angle = 90, size = 4) +
+  # annotate alluvia
+  annotate("segment", x = 4.33, y = 6.5, xend = 1, yend = 8.5, arrow = arr) +
+  annotate("segment", x = 5, y = 6.5, xend = 4.5, yend = 9, arrow = arr) +
+  annotate("segment", x = 5.67, y = 6.5, xend = 7.5, yend = 7, arrow = arr) +
+  annotate("text", x = 5, y = 6.17, label = "one alluvium", size = 4) +
+  # annotate flows
+  annotate("segment", x = 5.33, y = 2.5, xend = 4.25, yend = 0.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3, xend = 4.25, yend = 2.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3.5, xend = 3.75, yend = 4.5, arrow = arr) +
+  annotate("text", x = 5.5, y = 3, label = "flows", angle = 90, size = 4) +
+  # annotate lodes
+  annotate("segment", x = 7.5, y = 3, xend = 8, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 3.5, xend = 8, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 4, xend = 8, yend = 5.5, arrow = arr) +
+  annotate("text", x = 7.33, y = 3.5, label = "lodes", angle = -90, size = 4)
 ```

### 2020-03-24 paper/paper.rmd
simplify joss paper knit procedure
mv paper files to paper subdirectory

@@
 bibliography: paper.bib
-output: md_document
 ---
@@
 The ggalluvial package extends the layered grammar of graphics of ggplot2 [@Wickham2016] to generate alluvial diagrams from tidy data [@Wickham2014].
-The package makes two key contributions to the R ecosystem, which i summarize here and discuss in more detail below.
 
-First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g.
-with the text and point--line geoms of ggplot2.
-Meanwhile, the package gives users a high degree of control over underlying data transformations, and thereby over the messages conveyed in the plot, whereas other extensions are not so flexible.[^easy] The package also integrates alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.
-
-Second, ggalluvial cements some concepts and terms that are inconsistently applied by developers and users of Sankey, parallel sets, and alluvial diagrams.
-For example, the diagrams that came to be known as parallel sets plots [@Kosara2006] were anticipated in the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005], using wholly different terms for their graphical elements.
-Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types.
-Instead, ggalluvial adopts a distinctive, geological nomenclature for its graphical elements.
-While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it.
-I hope that these conventions can serve as useful points of reference as these visualization tools converge toward common standards.
-
-[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for expedited data exploration using alluvial plots.
+The package makes two key contributions to the R ecosystem.
+First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which lends the plots more precise meaning and opens up many combinatorial possibilities.
+Second, ggalluvial adopts a distinctive geological nomenclature to distinguish "alluvial plots" and their graphical elements from Sankey diagrams and parallel sets plots, which I hope prove useful as these visualization tools converge toward common standards.
 
@@
 
-[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations.
+[The primary vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples.
 In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:
 
-* _strata_, or stacked bar plots, located at each of multiple horizontal positions (_axes_ or _dimensions_)
-* _alluvia_, ribbons through strata at different axes that encode individual cases or cohorts measured along each axis
+* _strata_, or stacked bar plots, located in parallel along a (plotting) axis of (variable) _axes_ or _dimensions_
+* _alluvia_, ribbons through strata that connect the categories of individual cases or cohorts at different axes
 * _lodes_, subdivisions of strata by their intersections with alluvia
-* _flows_, segments of alluvia between adjacent axes (excluding lodes)
+* _flows_, segments of alluvia between strata
 
-The layers of a ggplot2 graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
-The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation.
-Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by `fill` color:
+Figure \ref{fig:example} illustrates these and other plot elements by visualizing changes in several students' curricula (based on their declared majors) across several academic terms.
+Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc.
+The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but had switched to Painting by the 11th term, where the ribbon turns pink.
+The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves.
+When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.
 
-```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
+```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap="\\label{fig:example}Alluvial plot of changes in curricula by a cohort of art students"}
 library(ggalluvial)
-combos <- data.frame(
-  stat = c(
-    rep("stratum", 2),
-    rep("alluvium", 4),
-    "flow",
-    # text geom
-    c("stratum", "alluvium", "flow"),
-    # errorbar and pointrange geoms
-    "stratum", "alluvium", "flow"
-  ),
-  geom = c(
-    c("stratum", "lode"),
-    c("stratum", "alluvium", "lode", "flow"),
-    "flow",
-    rep("text", 3),
-    "errorbar", "pointrange", "pointrange"
-  ),
-  pairing = c(
-    c("default", "alluvial"),
-    c("alluvial", "default", "default", "alluvial"),
-    "default",
-    rep("external", 6)
-  )
-)
-levs <- c("stratum", "alluvium", "lode", "flow",
-          "text", "errorbar", "pointrange")
-combos <- transform(
-  combos,
-  stat = factor(stat, levs),
-  geom = factor(geom, levs),
-  pairing = factor(pairing, c("default", "alluvial", "external"))
-)
-ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
-  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
-  geom_alluvium(aes(fill = pairing)) +
-  scale_fill_brewer(type = "qual") +
-  geom_stratum() +
-  geom_text(stat = "stratum", infer.label = TRUE) +
-  theme_minimal()
+data(majors)
+majors$curriculum <- as.factor(majors$curriculum)
+levels(majors$semester) <- gsub("CURR", "Term ", levels(majors$semester))
+arr <- arrow(angle = 15, type = "closed", length = unit(.03, "npc"))
+ggplot(majors,
+       aes(x = semester, stratum = curriculum, alluvium = student,
+           fill = curriculum)) +
+  scale_fill_brewer(type = "qual", palette = "Set2", name = "Curriculum") +
+  scale_y_continuous(breaks = seq(-2, 10, 2)) +
+  geom_flow(stat = "alluvium", lode.guidance = "frontback",
+            color = "darkgray", negate.strata = NA) +
+  geom_stratum(negate.strata = NA) +
+  theme(legend.position = "right") +
+  labs(x = NULL, y = "Number of students") +
+  # annotate axes
+  annotate("segment", x = 2.33, y = -1.33, xend = 2, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3, y = -1.33, xend = 3, yend = -.33, arrow = arr) +
+  annotate("segment", x = 3.67, y = -1.33, xend = 4, yend = -.33, arrow = arr) +
+  annotate("text", x = 3, y = -1.67, label = "axes / dimensions", size = 4) +
+  # annotate strata
+  annotate("segment", x = 1.5, xend = 1, y = 3.5, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 1.5, xend = 1, y = 4.5, yend = 6, arrow = arr) +
+  annotate("text", x = 1.67, y = 4, label = "strata", angle = 90, size = 4) +
+  # annotate alluvia
+  annotate("segment", x = 4.33, y = 6.5, xend = 1, yend = 8.5, arrow = arr) +
+  annotate("segment", x = 5, y = 6.5, xend = 4.5, yend = 9, arrow = arr) +
+  annotate("segment", x = 5.67, y = 6.5, xend = 7.5, yend = 7, arrow = arr) +
+  annotate("text", x = 5, y = 6.17, label = "one alluvium", size = 4) +
+  # annotate flows
+  annotate("segment", x = 5.33, y = 2.5, xend = 4.25, yend = 0.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3, xend = 4.25, yend = 2.5, arrow = arr) +
+  annotate("segment", x = 5.33, y = 3.5, xend = 3.75, yend = 4.5, arrow = arr) +
+  annotate("text", x = 5.5, y = 3, label = "flows", angle = 90, size = 4) +
+  # annotate lodes
+  annotate("segment", x = 7.5, y = 3, xend = 8, yend = 1.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 3.5, xend = 8, yend = 3.5, arrow = arr) +
+  annotate("segment", x = 7.5, y = 4, xend = 8, yend = 5.5, arrow = arr) +
+  annotate("text", x = 7.33, y = 3.5, label = "lodes", angle = -90, size = 4)
 ```
 
-The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.
-Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
-The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of tidyr.
-Because the alluvial geoms are highly specialized to these stats, no pairings with outside stats are currently supported.
+Plot layers are formed by pairing stats (statistical transformations) with geoms (mappings to graphical elements and properties); while every stat and geom has a conventional default, alternative grammatical pairings provide combinatorial richness to plotting possibilities.
+In the above example, the alluvium geom was paired with the flow stat, so that the flows of each alluvium could change color across the axes.
+Other meaningful stat--geom combinations can be found in the documentation, including pairings of the three alluvial stats (stratum, alluvium, and flow) with the text, errorbar, and pointrange geoms.
+
+Alluvial layers can interpret tidy data in either of two formats: long (one row per lode) and wide (one row per alluvium).
+These are related by the pivot operations of tidyr [@Wickham2019] and can be toggled between using the custom functions `to_lodes_form()` and `to_alluvia_form()`.
+The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.[^warnings] Because the alluvial geoms are specialized to these stats, no pairings with outside stats are currently supported.
+
+[^warnings]: Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.
 
@@
 By default, these orderings are independent of differentiation aesthetics, so that layers are consistent within and across plots unless otherwise specified.
-[An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of each of these parameters.
-For convenience, they can also be set as global options.
-
-While questions and requests by users have prompted improvements throughout ggalluvial's development, the current version is expected to be feature complete, with one exception: alternative curves will be implemented for the `alluvium` and `flow` geoms, controlled by a new parameter.
-The current splines will remain the default, so that the output of existing code will not change.
-This is expected to be the last change before version `1.0`.
+[An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of these parameters.
+They can also be set as global options.
 
@@
 
-Statistical graphics of multivariate categorical data tend to take one of two forms: ordinations of continuous-valued transformations (e.g.
-residual plots from multiple regression; ternary plots from tripartition data; biplots from correspondence analysis and factor analysis) and arrangements of length- or area-encoded shapes (e.g.
-mosaic plots [@Friendly2002]; product plots [@Wickham2011]; stacked area charts).
-In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990], represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
+Visualizations of flow processes have long encoded magnitudes as ribbon widths, constituting a type called Sankey diagrams [@Schmidt2008].
+A widely-used subtype for longitudinal categorical data represent categories as nodes threaded by edges that represent the trajectories and magnitudes of cases [@Riehmann2005].
+Their design anticipated parallel sets plots, which were adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to visualize multivariate categorical data, and which represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
+These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
+Several R packages have been developed to generate diagrams of these types, including riverplot [@Weiner2017], networkD3 [@Allaire2017], sankey [@Csardi2017], alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].
 
-Meanwhile, visualizations of flow processes have long encoded magnitudes, such as volumes and counts, as ribbon widths, and this broad genre have come to be called Sankey diagrams [@Schmidt2008]; see <http://www.sankey-diagrams.com/> for many examples.
-A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].
+Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
+Moreover, their graphical elements go by a variety of names, often interchangeably.
+In order to more clearly describe the features of ggalluvial in relation to similar packages, I have found it useful to adopt a careful demarcation among these diagram types.
 
-Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cohorts of equivalent cases encoded as ribbons intersecting stacked bar plots at each time point.
-These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions.
+_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
+By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
+Sankey diagrams are much more flexible.
+The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
+Software implementations may use heuristic algorithms to position their graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
+Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
+Arguably, these extensions are better understood as producing a different type of diagram.
 
-Several R packages, including ggplot2 extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:
+[^exceptions]: There are exceptions.
+For example, the jitter geom in ggplot2 introduces randomness to symbol positions, and the repel geoms of ggrepel [@Slowikowski2019] use heuristic algorithms to position text.
 
-- alluvial for alluvial diagrams in base R [@Bojanowski2016]
-- ggforce for parallel sets plots in ggplot2 [@Pedersen2019]
-- ggparallel and ggpcp for parallel sets, hammock, and common angle plots in ggplot2 [@Hofmann2013;@Ge2019]
-- ggalluvial for alluvial diagrams in ggplot2 [@Brunson2019]
+Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis.
+These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing, and optionally weighted).
+In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
 
-While Sankey diagrams, parallel sets plots, and alluvial diagrams have accrued their own characteristic use cases, the terms are often used interchangeably, and there is currently no consensus on what features, if any, are distinctive to each type.
-Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, while others are wholly determined by the data; most give the user limited control through parameter settings.
+The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
+The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons.
+Those of alluvial, ggforce, ggalluvial, and ggpcp offer one-parameter families that interpolate between straight and x-spline ribbons.[^curves]
+The stats vertically arrange the elements (boxes and ribbons) at each axis.
+These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
+ggalluvial provides more levers of control over the statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
 
-The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
-The package adopts for them the less common term _alluvial plot_.[^plots]
-These plots are distinct from Sankey diagrams (they exhibit a much narrower range of behavior, e.g.
-no cycles, and do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most diagrams of each type), features shared only by ggparallel.
-The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.
+[^curves]: Several alternative curves, based on @Shaffer2019, are in development.
+[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.
 
+The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces, with the geological terminology introduced above.[^plots]
+These alluvial plots are distinguished by two features: a prescribed order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
+In combination, these features confer greater meaning on the second plotting dimension.
+
+The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could, for example, arrange boxes corresponding to repeated categorical decompositions differently at different axes.
+While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots.
+(ggparallel and ggalluvial alone _only_ produce alluvial plots.)
+These features are particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
+It is for this reason that I believe the new typology and terminology are warranted.
+
+[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than I use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
 [^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
@@
 
-ggalluvial has seen widespread use in research and scholarship.
-While many of these uses would be served equally well by other parallel sets plots or Sankey diagrams, i showcase here three settings to which alluvial plots seem exceptionally well-suited: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.
-To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.
+While most uses might be served equally well by other parallel sets plots or Sankey diagrams, alluvial plots seem exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]
 
-**Repeated ordinal measures data:**
-Parallel sets plots and Sankey diagrams are commonly used to represent repeated categorical measures data.
-Most implementations stack each bar plot in order of name or of size, though some follow hierarchies provided by the user, often with vertical space between category for easy visual discrimination.
-In contrast, ordinal variables are more appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from time to time, vertical separations can obscure whether the cases and cohorts changed in number as well.
-A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
-Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in order across axes (in this case preference ranks rather than time points).
+[^opinion]: To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.
 
-**Incomplete longitudinal data:**
-Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
-@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times.
-Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type.
-While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
-@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
-<!--
-Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point.
-(See the example in the `stat_alluvium()` examples using the babynames package.)
--->
+**Repeated ordinal measures data.**
+Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size (though some follow user-provided hierarchies), and most insert gaps between categories for easy visual discrimination.
+Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories (hence of gaps) changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well.
+A use case by @Schlotter2019, to represent patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification).
+Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order.
+In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.
 
-**Signed categorical data:**
-@Edwards2019 recently produced a novel alluvial plot to represent changes in ownership category (owner-operated, corporate, etc.) of owners in a halibut fishery.
-Not only the categorical distribution but the total number of owners changed from year to year as exeters were not exactly matched by new entrants.
-In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, below the horizontal axis.
-Such a feature has no analogue in Sankey diagrams or parallel sets plots (which do not commonly employ a cumulative axis) but potentially wide-ranging applications: Bar plots may include both "positive" and "negative" bars to represent definitionally negative groups, such as revenues versus deficits, or to contrast the bars divided along a binary variable such as gender across age groups in a population (so-called "pyramid plots").
-Alluvial plots provide a way to track cases and cohorts across such bar plots, even when cases may change sign.
-Future use cases may demonstrate the practical utility of this functionality.
+**Incomplete longitudinal data.**
+Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
+@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected patients who had their infections ribotyped at multiple times.
+Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification.
+While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the main dimension.
+@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.
+
+**Signed categorical data.**
+@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery.
+The total number of owners changed from year to year as exiters were not exactly matched by new entrants.
+In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots").
+Alluvial plots provide a way to track cases and cohorts across such graphics, even when cases change sign.
+Future applications may demonstrate additional uses for this functionality.
+
+[^negative]: The authors should be credited with this innovation, which I only implemented in ggalluvial after learning about their workaround to create it using a previous version.
 
@@
 I am grateful to many users for their feedback on every version of this package.
-Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant from the National Institute of Dental and Craniofacial Research (5T90DE021989-07).
+Development benefitted from the use of resources and support of colleagues at UConn Health, and I have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
 

### 2020-03-25 paper/paper.rmd

@@
 ```r
-knitr::knit("paper/paper.rmd", "paper/paper.md")
+knitr::knit("paper.rmd", "paper.md")
 ```
@@
   comment = "#>",
-  fig.path = "paper/"
+  fig.path = ""
 )

### 2020-04-10 README.rmd

@@
   geom_alluvium(aes(fill = Survived)) +
-  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
+  geom_stratum() +
+  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
   theme_minimal() +

### 2020-04-13 README.rmd
change format function params to mimic pivot functions

--- 2020-03-04 README.rmd
@@
 titanic_long <- to_lodes_form(data.frame(Titanic),
-                              key = "Demographic",
-                              axes = 1:3)
+                              axes = 1:3,
+                              axes_to = "Demographic")
 head(titanic_long)

### 2020-04-13 R/alluvial-data.r

@@ -1,55 +1,70 @@
-#' Check for alluvial structure and convert between alluvial formats
+#' @title Test for alluvial structure and convert between alluvial formats
+#'
+#' @description
+#'
+#' The functions `is_*_form()` determine whether a data frame is alluvial---that
+#' it has either the id--key--value structure of lodes format or the tidy
+#' properties of alluvia format. The functions `to_*_form()` convert data from
+#' one format to the other.
+#' 
+
+#' @details
 #'
 #' Alluvial plots consist of multiple horizontally-distributed columns (axes)
-#' representing factor variables, vertical divisions (strata) of these axes
-#' representing these variables' values; and splines (alluvial flows) connecting
-#' vertical subdivisions (lodes) within strata of adjacent axes representing
+#' representing categorical variables, vertical divisions (strata) of these axes
+#' representing these variables' values; and ribbons (alluvial flows) connecting
+#' vertical subdivisions (lodes) within strata of adjacent axes, representing
 #' subsets or amounts of observations that take the corresponding values of the
-#' corresponding variables. This function checks a data frame for either of two
-#' types of alluvial structure:
+#' corresponding variables. **ggalluvial** can generate an alluvial plot from
+#' data in either of two recognized formats:
+#' 
+
+#' - **Lodes format** encodes _one row per measurement_ or _one row per lode_,
+#' using an `id`--`key`--`value` structure: Each row contains the recorded
+#' `value` of one measurement (the `key`) for one subject or cohort (the `id`).
+#' Visually, the lode is the unique intersection of one alluvium (the `id`) and
+#' one axis (the `key`) and belongs to a unique stratum (the `value`).
+#' Additional columns may contain lode-level variables, including magnitudes and
+#' weights.
+
+#' - **Alluvia format** encodes _one row per subject_ or _one row per alluvium_,
+#' using a [tidy](https://tidyr.tidyverse.org/) structure: Each subject (or
+#' cohort) has a row, each measurement has a column, and each measured value is
+#' an entry. Additional columns may contain subject-level variables, including
+#' magnitudes and weights; but lode-level magnitudes and weights may also be
+#' encoded as column ranges having the same cardinality as those of the
+#' measurements.
+
 #'
[…]

### 2020-04-13 R/stat-alluvium.r

@@ -135,7 +135,8 @@ StatAlluvium <- ggproto(
     # ensure that data is in lode form
     if (type == "alluvia") {
       axis_ind <- get_axes(names(data))
-      data <- to_lodes_form(data = data, axes = axis_ind,
+      data <- to_lodes_form(data = data,
+                            axes = axis_ind,
                             discern = params$discern)
       # positioning requires numeric `x`
       data <- data[with(data, order(x, stratum, alluvium)), , drop = FALSE]
@@ -227,7 +228,7 @@ StatAlluvium <- ggproto(
       }
       stopifnot(is.function(lode.guidance))
       # summary data of alluvial deposits
-      alluv_dep <- alluviate(data, "x", "deposit", "alluvium")
+      alluv_dep <- alluviate(data, "alluvium", "x", "deposit")
       # axis indices
       alluv_x <- setdiff(names(alluv_dep), "alluvium")
       # calculate `lode.ordering` from `lode.guidance`
@@ -358,7 +359,7 @@ cement_data <- function(data, key, id, fun) {
     addNA, ifany = FALSE
   ), drop = TRUE))
   # convert to alluvia format with 'binding' entries
-  alluv_data <- alluviate(data, key, "binding", id)
+  alluv_data <- alluviate(data, id, key, "binding")
   # sort by all axes (everything except `id`)
   alluv_data <- alluv_data[do.call(
     order,
@@ -405,7 +406,7 @@ cement_data_alt <- function(data, key, id) {
   ), drop = TRUE))
   
   # convert to alluvia format with '.binding' entries
-  alluv_data <- alluviate(data, key, ".binding", id)
+  alluv_data <- alluviate(data, id, key, ".binding")
   
   # redefine 'alluvium' with only as many distinct values as distinct alluvia
   alluv_agg <- stats::aggregate(
@@ -436,9 +437,9 @@ cement_data_alt <- function(data, key, id) {
 }
 
 # build alluvial dataset for reference during lode-ordering
-alluviate <- function(data, key, value, id) {
+alluviate <- function(data, id, key, value) {
   to_alluvia_form(
-    data[, c(key, value, id)],
-    key = key, value = value, id = id
+    data[, c(id, key, value)],
+    alluvia_from = id, axes_from = key, strata_from = value
   )
 }

### 2020-04-13 R/stat-flow.r

@@ -105,7 +105,8 @@ StatFlow <- ggproto(
     # ensure that data is in lode form
     if (type == "alluvia") {
       axis_ind <- get_axes(names(data))
-      data <- to_lodes_form(data = data, axes = axis_ind,
+      data <- to_lodes_form(data = data,
+                            axes = axis_ind,
                             discern = params$discern)
       # positioning requires numeric `x`
       data <- data[with(data, order(x, stratum, alluvium)), , drop = FALSE]

### 2020-04-13 R/stat-stratum.r

@@ -125,7 +125,8 @@ StatStratum <- ggproto(
     # ensure that data is in lode form
     if (type == "alluvia") {
       axis_ind <- get_axes(names(data))
-      data <- to_lodes_form(data = data, axes = axis_ind,
+      data <- to_lodes_form(data = data,
+                            axes = axis_ind,
                             discern = params$discern)
       # positioning requires numeric `x`
       data <- data[with(data, order(x, stratum, alluvium)), , drop = FALSE]

### 2020-04-13 ggplot2-exts/thumbnail.r

@@ -1,10 +1,12 @@
 png(filename = "ggalluvial.png",
     width = 350, height = 300)
-ggplot(data = to_lodes(as.data.frame(Titanic),
-                       key = "Demographic",
-                       axes = 1:3),
-       aes(x = Demographic, stratum = value, alluvium = id,
-           weight = Freq, label = value)) +
+ggplot(data = to_lodes_form(as.data.frame(Titanic),
+                            alluvia_to = "Cohort",
+                            axes_to = "Demographic",
+                            strata_to = "Value",
+                            axes = 1:3),
+       aes(x = Demographic, stratum = Value, alluvium = Cohort,
+           y = Freq, label = Value)) +
   geom_alluvium(aes(fill = Survived)) +
   geom_stratum() + geom_text(stat = "stratum") +
   theme(legend.position = "bottom") +

### 2020-04-13 tests/testthat/test-alluvial-data.r

@@ -38,7 +38,7 @@ test_that("`to_lodes_form` consistently formats Titanic data", {
 # preparation for next tests
 titanic_lodes <- suppressWarnings(to_lodes_form(
   transform(titanic_alluvia, Index = 1:nrow(titanic_alluvia)),
-  key = "Variable", value = "Value", id = "Index", axes = 1:4
+  alluvia_to = "Index", axes_to = "Variable", strata_to = "Value", axes = 1:4
 ))
 titanic_lodes$Value <-
   factor(titanic_lodes$Value,
@@ -49,36 +49,44 @@ titanic_lodes$Value <-
 test_that("`is_lodes_form` recognizes lodes-format Titanic data", {
   expect_error(is_lodes_form(titanic_lodes))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = "Variable", value = "Value", id = "Index"))
+                            alluvia_from = "Index", axes_from = "Variable",
+                            strata_from = "Value"))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = Variable, value = Value, id = Index))
+                            alluvia_from = Index, axes_from = Variable,
+                            strata_from = Value))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = 3, value = 4, id = 2))
+                            alluvia_from = 2, axes_from = 3, strata_from = 4))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = "Variable", value = "Value", id = "Index",
-                            y = "Freq"))
+                            alluvia_from = "Index", axes_from = "Variable",
+                            strata_from = "Value", y = "Freq"))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = 3, value = 4, id = 2,
+                            alluvia_from = 2, axes_from = 3, strata_from = 4,
                             y = 1))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = Variable, value = Value, id = Index,
-                            y = Freq))
+                            alluvia_from = Index, axes_from = Variable,
+                            strata_from = Value, y = Freq))
   expect_true(is_lodes_form(titanic_lodes,
-                            key = Variable, value = Value, id = Index,
-                            y = !! null_wt))
+                            alluvia_from = Index, axes_from = Variable,
+                            strata_from = Value, y = !! null_wt))
 })
 
 # `to_alluvia_form()` tests
 
 test_that("`to_alluvia_form` consistently formats Titanic data", {
   expect_equivalent(to_alluvia_form(titanic_lodes,
[…]

### 2020-04-16 README.rmd
fix links in README per CRAN issues

--- 2020-04-10 README.rmd
@@
 
-This is a [**ggplot2** extension](http://www.ggplot2-exts.org/) for alluvial plots.
+This is a [**ggplot2** extension](https://exts.ggplot2.tidyverse.org/) for alluvial plots.
 
@@
            y = Freq)) +
-  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
+  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.2, .05)) +
   xlab("Demographic") +
   geom_alluvium(aes(fill = Survived)) +
-  geom_stratum() +
-  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
+  geom_stratum() + geom_text(stat = "stratum", infer.label = TRUE) +
   theme_minimal() +
@@
 Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.
-See the [CONTRIBUTING](CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](CODE_OF_CONDUCT.md).
+See the [CONTRIBUTING](https://github.com/corybrunson/ggalluvial/blob/master/CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](https://github.com/corybrunson/ggalluvial/blob/master/CODE_OF_CONDUCT.md).
 

### 2020-05-06 README.rmd
add JOSS status badge

--- 2020-04-16 README.rmd
@@
 [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
+[![JOSS](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723/status.svg)](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723)
 

### 2020-05-21 README.rmd
docs(readme): revert badge names + edit readme.rmd

@@
 [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3836748.svg)](https://doi.org/10.5281/zenodo.3836748)
 [![JOSS](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723/status.svg)](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723)

### 2020-05-21 README.rmd
docs(readme): reorganize badges + add JOSS paper

--- 2020-05-06 README.rmd
@@
 
-[![Travis build status](https://travis-ci.org/corybrunson/ggalluvial.svg?branch=master)](https://travis-ci.org/corybrunson/ggalluvial)
-[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
-[![JOSS](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723/status.svg)](https://joss.theoj.org/papers/2a38fbb08d94994145d01c370f243723)
+[![Travis](https://travis-ci.org/corybrunson/ggalluvial.svg?branch=master)](https://travis-ci.org/corybrunson/ggalluvial)
+[![CRAN](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
+
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3836748.svg)](https://doi.org/10.5281/zenodo.3836748)
+[![JOSS](https://joss.theoj.org/papers/10.21105/joss.02017/status.svg)](https://doi.org/10.21105/joss.02017)
 

### 2020-07-23 README.rmd
docs(readme,home): update branch descriptions from master to main

@@
 
-[![Travis](https://travis-ci.org/corybrunson/ggalluvial.svg?branch=master)](https://travis-ci.org/corybrunson/ggalluvial)
+[![Travis](https://travis-ci.org/corybrunson/ggalluvial.svg?branch=main)](https://travis-ci.org/corybrunson/ggalluvial)
 [![CRAN](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
@@
 
-The [cran branch](https://github.com/corybrunson/ggalluvial/tree/cran) will contain the version most recently submitted to [CRAN](https://cran.r-project.org/package=ggalluvial).
+The [`cran` branch](https://github.com/corybrunson/ggalluvial/tree/cran) will contain the version most recently submitted to [CRAN](https://cran.r-project.org/package=ggalluvial).
+It is duplicated in the `master` branch, from which source the [website](https://corybrunson.github.io/ggalluvial/) is built.
 
-The development version can be installed from GitHub:
+The development version can be installed from the (default) `main` branch on GitHub:
 ```{r eval=FALSE}
-remotes::install_github("corybrunson/ggalluvial", build_vignettes = TRUE)
+remotes::install_github("corybrunson/ggalluvial@main", build_vignettes = TRUE)
 ```
@@
 The
-[optimization branch](https://github.com/corybrunson/ggalluvial/tree/optimization) contains a development version with experimental functions to reduce the number or area of alluvial overlaps (see issue [#6](https://github.com/corybrunson/ggalluvial/issues/6)).
+[`optimization` branch](https://github.com/corybrunson/ggalluvial/tree/optimization) contains a development version with experimental functions to reduce the number or area of alluvial overlaps (see issue [#6](https://github.com/corybrunson/ggalluvial/issues/6)).
 Install it as follows:
@@
 ```
-Note, however, that this branch has not kept pace with the `master` branch or with recent upgrades on CRAN.
+Note, however, that this branch has not kept pace with the `main` branch or with recent upgrades on CRAN.
 
@@
 Pretty much every fix and feature of this package derives from a problem or question posed by someone with datasets or design goals i hadn't anticipated.
-See the [CONTRIBUTING](https://github.com/corybrunson/ggalluvial/blob/master/CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](https://github.com/corybrunson/ggalluvial/blob/master/CODE_OF_CONDUCT.md).
+See the [CONTRIBUTING](https://github.com/corybrunson/ggalluvial/blob/main/CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](https://github.com/corybrunson/ggalluvial/blob/main/CODE_OF_CONDUCT.md).
 

### 2020-08-10 README.rmd
docs(ggalluvial,readme): change http to https and rebuild site

--- 2020-07-23 README.rmd
@@
 - **alluvial** understands a variety of inputs (vectors, lists, data frames), whereas **ggalluvial** requires a single data frame;
-- **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](http://tidyr.tidyverse.org/);
+- **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](https://tidyr.tidyverse.org/);
 - **alluvial** produces both the *alluvia*, which link cohorts across multiple dimensions, and (what are here called) the *strata*, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called *lodes* and *flows*.

### 2020-12-01 README.rmd
pkg: doc formatting + version/date + debug citation + revdep lists

--- 2020-08-10 README.rmd
@@
 - **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](https://tidyr.tidyverse.org/);
-- **alluvial** produces both the *alluvia*, which link cohorts across multiple dimensions, and (what are here called) the *strata*, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called *lodes* and *flows*.
+- **alluvial** produces both the _alluvia_, which link cohorts across multiple dimensions, and (what are here called) the _strata_, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called _lodes_ and _flows_.
 

### 2021-01-26 inst/examples/ex-shiny-wide-data/app.R
edit text of new vig, change expansion on ucba plot

@@ -12,7 +12,7 @@ UCBAdmissions <- UCBAdmissions[order(UCBAdmissions$Admit, UCBAdmissions$Gender, 
 # in both x and y direction.
 offset <- 5
 # Width of node boxes
-node_width <- 1/12
+node_width <- 1/4
 
 # Draw plot.
 p <- ggplot(UCBAdmissions,
@@ -25,7 +25,7 @@ p <- ggplot(UCBAdmissions,
              size = rel(3)) + 
   theme_bw() +
   scale_fill_brewer(type = "qual", palette = "Set1") +
-  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(0, 0)) +
+  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
   scale_y_continuous(expand = c(0, 0)) +
   ggtitle("UC Berkeley admissions and rejections, by sex and department")
 

### 2021-01-26 inst/examples/ex-shiny-long-data/app.R
make both apps server fxn identical

@@ -144,12 +144,10 @@ server <- function(input, output, session) {
         if (any(hover_within_flow)) {
           # Find the alluvium that is plotted on top. (last)
           coord_id <- rev(which(hover_within_flow == 1))[1]
-          # Find the subject and freq corresponding to that alluvium in the data.
-          flow_label <- paste(groups_to_draw[[coord_id]]$label, collapse = ' -> ')
+          # Find the strata labels and n corresponding to that alluvium in the data.
+          flow_label <- paste(groups_to_draw[[coord_id]]$stratum, collapse = ' -> ')
           flow_n <- groups_to_draw[[coord_id]]$count[1]
-          # Get the values for all axes for that row ID.
-          #axis_values <- sapply(UCBAdmissions[coord_id, c('Gender', 'Dept')], as.character)
-          
+
           # Render tooltip
           renderTags(
             tags$div(

### 2021-01-26 inst/examples/ex-shiny-wide-data/app.R

@@ -144,13 +144,15 @@ server <- function(input, output, session) {
         if (any(hover_within_flow)) {
           # Find the alluvium that is plotted on top. (last)
           coord_id <- rev(which(hover_within_flow == 1))[1]
-          # Get the values for all axes for that row ID.
-          axis_values <- sapply(UCBAdmissions[coord_id, c('Gender', 'Dept')], as.character)
+          # Find the strata labels and n corresponding to that alluvium in the data.
+          flow_label <- paste(groups_to_draw[[coord_id]]$stratum, collapse = ' -> ')
+          flow_n <- groups_to_draw[[coord_id]]$count[1]
           
           # Render tooltip
           renderTags(
             tags$div(
-              paste(axis_values, collapse = ' -> '),
+              flow_label, tags$br(),
+              "n =", flow_n,
               style = paste0(
                 "position: absolute; ",
                 "top: ", hover$coords_css$y + offset, "px; ",

### 2021-01-26 inst/examples/ex-shiny-wide-data/app.R
replace placeholders with actual text make a few updates to wide app

@@ -5,8 +5,6 @@ library(sp)
 
 data(UCBAdmissions)
 UCBAdmissions <- as.data.frame(UCBAdmissions)
-# Arrange data so that the row IDs match the order in the data
-UCBAdmissions <- UCBAdmissions[order(UCBAdmissions$Admit, UCBAdmissions$Gender, UCBAdmissions$Dept), ]
 
 # Offset, in pixels, for location of tooltip relative to mouse cursor,
 # in both x and y direction.
@@ -64,7 +62,7 @@ yrange_old <- range(unlist(lapply(
   function(pts) as.numeric(pts$y)
 )))
 xrange_new <- c(1 - 1/6, max(pbuilt$data[[1]]$x) + 1/6) 
-yrange_new <- c(0, sum(UCBAdmissions$Freq)) 
+yrange_new <- c(0, sum(pbuilt$data[[2]]$count[pbuilt$data[[2]]$x == 1])) 
 
 # Define function to convert grid graphics coordinates to data coordinates
 new_range_transform <- function(x_old, range_old, range_new) {

### 2021-01-29 inst/examples/ex-shiny-long-data/app.R
Remove hard-coded 1/3 value for width and change name of ucb dataset

@@ -12,6 +12,8 @@ vaccinations <- transform(vaccinations,
 offset <- 5
 # Width of node boxes
 node_width <- 1/3
+# Width of alluvia
+alluvium_width <- 1/3
 
 # Draw plot and extract coordinates
 p <- ggplot(vaccinations,
@@ -20,7 +22,7 @@ p <- ggplot(vaccinations,
                 fill = response, label = response)) +
   scale_x_discrete(expand = c(.1, .1)) +
   scale_y_continuous(expand = c(0, 0)) +
-  geom_flow(knot.pos = 1/4) +
+  geom_flow(knot.pos = 1/4, width = alluvium_width) +
   geom_stratum(width = node_width) +
   geom_text(stat = "stratum", size = rel(3)) +
   theme_bw() +
@@ -34,7 +36,7 @@ pbuilt <- ggplot_build(p)
 # Use built plot data to recalculate the locations of the flow polygons:
 
 # Add width parameter, and then convert built plot data to xsplines
-data_draw <- transform(pbuilt$data[[1]], width = 1/3)
+data_draw <- transform(pbuilt$data[[1]], width = alluvium_width)
 groups_to_draw <- split(data_draw, data_draw$group) # Global assignment because this is used to extract information for the tooltip later.
 group_xsplines <- lapply(groups_to_draw,
                          ggalluvial:::data_to_xspline,
@@ -62,7 +64,7 @@ yrange_old <- range(unlist(lapply(
   xspline_points,
   function(pts) as.numeric(pts$y)
 )))
-xrange_new <- c(1 - 1/6, max(pbuilt$data[[1]]$x) + 1/6) 
+xrange_new <- c(1 - alluvium_width/2, max(pbuilt$data[[1]]$x) + alluvium_width/2) 
 yrange_new <- c(0, sum(pbuilt$data[[2]]$count[pbuilt$data[[2]]$x == 1])) 
 
 # Define function to convert grid graphics coordinates to data coordinates

### 2021-01-29 inst/examples/ex-shiny-wide-data/app.R

@@ -4,18 +4,20 @@ library(htmltools)
 library(sp)
 
 data(UCBAdmissions)
-UCBAdmissions <- as.data.frame(UCBAdmissions)
+ucb_admissions <- as.data.frame(UCBAdmissions)
 
 # Offset, in pixels, for location of tooltip relative to mouse cursor,
 # in both x and y direction.
 offset <- 5
 # Width of node boxes
 node_width <- 1/4
+# Width of alluvia
+alluvium_width <- 1/3
 
 # Draw plot.
-p <- ggplot(UCBAdmissions,
+p <- ggplot(ucb_admissions,
             aes(y = Freq, axis1 = Gender, axis2 = Dept)) + 
-  geom_alluvium(aes(fill = Admit), knot.pos = 1/4) + 
+  geom_alluvium(aes(fill = Admit), knot.pos = 1/4, width = alluvium_width) + 
   geom_stratum(width = node_width, reverse = TRUE, fill = 'black', color = 'grey') + 
   geom_label(aes(label = after_stat(stratum)), 
              stat = "stratum", 
@@ -33,7 +35,7 @@ pbuilt <- ggplot_build(p)
 # Use built plot data to recalculate the locations of the flow polygons:
 
 # Add width parameter, and then convert built plot data to xsplines
-data_draw <- transform(pbuilt$data[[1]], width = 1/3)
+data_draw <- transform(pbuilt$data[[1]], width = alluvium_width)
 groups_to_draw <- split(data_draw, data_draw$group)
 group_xsplines <- lapply(groups_to_draw,
                          ggalluvial:::data_to_xspline,
@@ -61,7 +63,7 @@ yrange_old <- range(unlist(lapply(
   xspline_points,
   function(pts) as.numeric(pts$y)
 )))
-xrange_new <- c(1 - 1/6, max(pbuilt$data[[1]]$x) + 1/6) 
+xrange_new <- c(1 - alluvium_width/2, max(pbuilt$data[[1]]$x) + alluvium_width/2) 
 yrange_new <- c(0, sum(pbuilt$data[[2]]$count[pbuilt$data[[2]]$x == 1])) 
 
 # Define function to convert grid graphics coordinates to data coordinates

### 2021-02-07 inst/examples/ex-shiny-wide-data/app.R
doc(shiny): adjust sizings of labels and app windows in vignette

@@ -27,8 +27,9 @@ p <- ggplot(ucb_admissions,
   scale_fill_brewer(type = "qual", palette = "Set1") +
   scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
   scale_y_continuous(expand = c(0, 0)) +
-  ggtitle("UC Berkeley admissions and rejections, by sex and department") +
-  theme(plot.title = element_text(size = rel(0.8)))
+  ggtitle("UC Berkeley admissions and rejections", "by sex and department") +
+  theme(plot.title = element_text(size = rel(1)),
+        plot.subtitle = element_text(size = rel(1)))
 
 # Build the plot. 
 pbuilt <- ggplot_build(p)

### 2021-02-23 inst/examples/ex-shiny-long-data/app.R
fix(shiny): shrink labels + attempt link to primary vignette

@@ -24,7 +24,7 @@ p <- ggplot(vaccinations,
   scale_y_continuous(expand = c(0, 0)) +
   geom_flow(knot.pos = 1/4, width = alluvium_width) +
   geom_stratum(width = node_width) +
-  geom_text(stat = "stratum", size = rel(3)) +
+  geom_text(stat = "stratum", size = rel(2)) +
   theme_bw() +
   theme(legend.position = "none") +
   ggtitle("Vaccination responses on three surveys") +

### 2021-04-03 README.rmd

@@
 [![CRAN](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
+[![downloads](https://cranlogs.r-pkg.org/badges/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
 

### 2021-04-06 README.rmd
pkg(site): rebuild site

--- 2020-12-01 README.rmd
@@
 [![CRAN](http://www.r-pkg.org/badges/version/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
-
+[![downloads](https://cranlogs.r-pkg.org/badges/ggalluvial)](https://cran.r-project.org/package=ggalluvial)
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3836748.svg)](https://doi.org/10.5281/zenodo.3836748)

### 2021-04-11 README.rmd
pkg(readme): describe vignettes & credit uf health - fixes #61

@@
 
-The documentation contains several examples; use `help()` to call forth examples of any layer (`stat_*` or `geom_*`) or of the conversion functions (`to_*_form`).
+Several additional vignettes offer detailed solutions to specific needs:
+
+* "Labeling small strata" (`"labels"`) for how to elegantly label strata of a wide range of heights in an alluvial plot;
+* "The Order of the Rectangles" (`"order-rectangles"`) for how to control the positioning of strata and lodes in an alluvial plot; and
+* "Tooltips for ggalluvial plots in Shiny apps" (`"shiny"`) for how to incorporate alluvial plots into interactive apps.
+
+The object documentation includes several more examples. Use `help()` to call forth more detail on
+
+* any layer (`stat_*` or `geom_*`),
+* the conversion functions (`to_*_form`), and
+* the data sets installed with the package (`vaccinations` and `majors`).
 
@@
 
-Development of this package benefitted from the use of equipment and the support of colleagues at [UConn Health](https://health.uconn.edu/).
+Development of this package benefitted from the use of equipment and the support of colleagues at [UConn Health](https://health.uconn.edu/) and at [UF Health](https://ufhealth.org/).
 

### 2021-08-19 R/geom-alluvium.r
Export data_to_alluvium and modify shiny apps accordingly Only placeholder documentation is currently included

@@ -155,6 +155,11 @@ GeomAlluvium <- ggproto(
   draw_key = draw_key_polygon
 )
 
+#' Create alluvium curve points from data
+#' 
+#' Placeholder. Documentation goes here.
+#' 
+#' @export
 data_to_alluvium <- function(data, curve_type = 'spline', knot.prop = TRUE, curve_range = 6, segments = 3) {
   if (nrow(data) == 1) {
     # spline coordinates (one axis)

### 2021-08-19 inst/examples/ex-shiny-long-data/app.R

@@ -31,18 +31,16 @@ p <- ggplot(vaccinations,
   theme(plot.title = element_text(size = rel(1)),
         legend.position = 'bottom')
 
-# Build the plot. Use global assignment so that this object is accessible
-# later.
+# Build the plot.
 pbuilt <- ggplot_build(p)
 
 # Use built plot data to recalculate the locations of the flow polygons:
 
 # Add width parameter, and then convert built plot data to xsplines
 data_draw <- transform(pbuilt$data[[1]], width = alluvium_width)
-groups_to_draw <- split(data_draw, data_draw$group) # Global assignment because this is used to extract information for the tooltip later.
+groups_to_draw <- split(data_draw, data_draw$group) 
 group_xsplines <- lapply(groups_to_draw,
-                         ggalluvial:::data_to_xspline,
-                         knot.prop = TRUE) 
+                         data_to_alluvium) 
 
 # Convert xspline coordinates to grid object.
 xspline_coords <- lapply(

### 2022-08-21 README.rmd
link to edav community contribution project page
credit specific students and links

@@
 
+### Short form
+
+For some more digestible guidance on using **ggalluvial**, check out three cheat sheets and demos by students in Joyce Robbins's [Exploratory Data Analysis and Visualization Community Contribution Project](https://jtr13.github.io/cc21fall2/index.html):
+
+* an [introduction](https://jtr13.github.io/cc21fall2/ggalluvial-cheatsheet.html) by Meggie Wen (Chapter 33)
+* a [cheat sheet](https://cheatography.com/seleven/cheat-sheets/ggalluvial/) by Qingyi Zhang (Chapter 14)
+* a [blog about how this package implements alluvial plots](https://medium.com/@arnavsaxena96/all-about-alluvial-diagrams-21da1505520b) by Arnav Saxena (Chapter 134)
+
 ## Acknowledgments

### 2022-10-20 R/geom-alluvium.r
partially address linewidth change in ggplot2

@@ -83,7 +83,7 @@ GeomAlluvium <- ggproto(
   
   required_aes = c("x", "y", "ymin", "ymax"),
   
-  default_aes = aes(size = .5, linewidth = .5, linetype = 1,
+  default_aes = aes(linewidth = .5, linetype = 1,
                     colour = "transparent", fill = "gray", alpha = .5),
   
   setup_data = function(data, params) {
@@ -157,13 +157,18 @@ GeomAlluvium <- ggproto(
       x = coords$x, y = coords$y, shape = coords$shape,
       open = FALSE,
       gp = grid::gpar(
-        col = coords$colour, fill = coords$fill, alpha = coords$alpha,
-        lty = coords$linetype, lwd = coords$size * .pt
+        col = coords$colour, fill = coords$fill,
+        alpha = coords$alpha,
+        lty = coords$linetype,
+        lwd = (coords$linewidth %||% coords$size) * .pt
       )
     )
   },
   
-  draw_key = draw_key_polygon
+  draw_key = draw_key_polygon,
+  
+  non_missing_aes = "size",
+  rename_size = TRUE
 )
 
 #' @rdname geom_alluvium

### 2022-10-20 R/geom-flow.r

@@ -74,7 +74,7 @@ GeomFlow <- ggproto(
 
   required_aes = c("x", "y", "ymin", "ymax"),
 
-  default_aes = aes(size = .5, linewidth = .5, linetype = 1,
+  default_aes = aes(linewidth = .5, linetype = 1,
                     colour = "transparent", fill = "gray", alpha = .5),
 
   setup_data = function(data, params) {
@@ -112,7 +112,7 @@ GeomFlow <- ggproto(
     flow_pos <- intersect(names(data), c("x", "xmin", "xmax",
                                          "width", "knot.pos",
                                          "y", "ymin", "ymax"))
-    flow_aes <- intersect(names(data), c("size", "linetype",
+    flow_aes <- intersect(names(data), c("linewidth", "size", "linetype",
                                          "colour", "fill", "alpha"))
     flow_fore <- if (aes.flow != "backward") flow_aes else NULL
     flow_back <- if (aes.flow != "forward") flow_aes else NULL
@@ -156,8 +156,10 @@ GeomFlow <- ggproto(
         x = f_coords$x, y = f_coords$y, shape = f_coords$shape,
         open = FALSE,
         gp = grid::gpar(
-          col = f_coords$colour, fill = f_coords$fill, alpha = f_coords$alpha,
-          lty = f_coords$linetype, lwd = f_coords$size * .pt
+          col = f_coords$colour, fill = f_coords$fill,
+          alpha = f_coords$alpha,
+          lty = f_coords$linetype,
+          lwd = (f_coords$linewidth %||% f_coords$size) * .pt
         )
       )
     })
@@ -168,7 +170,10 @@ GeomFlow <- ggproto(
     grob
   },
 
-  draw_key = draw_key_polygon
+  draw_key = draw_key_polygon,
+  
+  non_missing_aes = "size",
+  rename_size = TRUE
 )
 
 #' @rdname geom_flow

### 2022-10-20 R/utils.r

@@ -4,7 +4,7 @@
   "alpha",
   "fg", "col", "colour", "color",
   "lty", "linetype",
-  "cex", "lwd", "size",
+  "cex", "lwd", "linewidth", "size",
   "pch", "shape"
 )
 

### 2023-01-30 README.rmd
tweak titular vignette + reformat package names + suppress warnings in tests

@@
 
-This is a [**ggplot2** extension](https://exts.ggplot2.tidyverse.org/) for alluvial plots.
+This is a [{ggplot2} extension](https://exts.ggplot2.tidyverse.org/) for alluvial plots.
 
@@
 The alluvial plots implemented here can be used to visualize frequency distributions over time or frequency tables involving several categorical variables.
-The design is inspired by the [**alluvial**](https://github.com/mbojan/alluvial) package, but the **ggplot2** framework induced several conspicuous differences:
+The design is inspired by the [{alluvial}](https://github.com/mbojan/alluvial) package, but the {ggplot2} framework induced several conspicuous differences:
 
-- **alluvial** understands a variety of inputs (vectors, lists, data frames), whereas **ggalluvial** requires a single data frame;
-- **alluvial** uses each variable of these inputs as a dimension of the data, whereas **ggalluvial** requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](https://tidyr.tidyverse.org/);
-- **alluvial** produces both the _alluvia_, which link cohorts across multiple dimensions, and (what are here called) the _strata_, which partition the data along each dimension, in a single function; whereas **ggalluvial** relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called _lodes_ and _flows_.
+- {alluvial} understands a variety of inputs (vectors, lists, data frames), whereas {ggalluvial} requires a single data frame;
+- {alluvial} uses each variable of these inputs as a dimension of the data, whereas {ggalluvial} requires the user to specify the dimensions, either as separate aesthetics or as [key-value pairs](https://tidyr.tidyverse.org/);
+- {alluvial} produces both the _alluvia_, which link cohorts across multiple dimensions, and (what are here called) the _strata_, which partition the data along each dimension, in a single function; whereas {ggalluvial} relies on separate layers (stats and geoms) to produce strata, alluvia, and alluvial segments called _lodes_ and _flows_.
 
-Additionally, **ggalluvial** arranges these layers vertically without gaps, so that the secondary plotting axis indicates the cumulative values of the strata at each dimension.
+Additionally, {ggalluvial} arranges these layers vertically without gaps, so that the secondary plotting axis indicates the cumulative values of the strata at each dimension.
 
@@
 ```
-Note that, in order to build the vignettes, the imported packages **alluvial**, **ggfittext**, and **ggrepel** must be installed.
-To skip this step, leave `build_vignettes` unspecified or set it to `FALSE`.
+Note that, in order to build the vignettes, the imported packages {alluvial}, {ggfittext}, and {ggrepel} must be installed. To skip this step, leave `build_vignettes` unspecified or set it to `FALSE`.
 
@@
 
-The data is in "wide" format, but **ggalluvial** also recognizes data in "long" format and can convert between the two:
+The data is in "wide" format, but {ggalluvial} also recognizes data in "long" format and can convert between the two:
 
@@
 
-For detailed discussion of the data formats recognized by **ggalluvial** and several examples that illustrate its flexibility and limitations, read the technical vignette:
+For detailed discussion of the data formats recognized by {ggalluvial} and several examples that illustrate its flexibility and limitations, read the technical vignette:
 
@@
 
-For some more digestible guidance on using **ggalluvial**, check out three cheat sheets and demos by students in Joyce Robbins's [Exploratory Data Analysis and Visualization Community Contribution Project](https://jtr13.github.io/cc21fall2/index.html):
+For some more digestible guidance on using {ggalluvial}, check out three cheat sheets and demos by students in Joyce Robbins's [Exploratory Data Analysis and Visualization Community Contribution Project](https://jtr13.github.io/cc21fall2/index.html):
 
@@
 
-If you use **ggalluvial**-generated figures in publication, i'd be grateful to hear about it! You can also cite the package according to `citation("ggalluvial")`.
+If you use {ggalluvial}-generated figures in publication, i'd be grateful to hear about it! You can also cite the package according to `citation("ggalluvial")`.
